\section{\module{email.Encoders} ---
         Email message payload encoders}

\declaremodule{standard}{email.Encoders}
\modulesynopsis{Encoders for email message payloads.}
\sectionauthor{Barry A. Warsaw}{barry@zope.com}

\versionadded{2.2}

When creating \class{Message} objects from scratch, you often need to
encode the payloads for transport through compliant mail servers.
This is especially true for \code{image/*} and \code{text/*} type
messages containing binary data.

The \module{email} package provides some convenient encodings in its
\module{Encoders} module.  These encoders are actually used by the
\class{MIMEImage} and \class{MIMEText} class constructors to provide default
encodings.  All encoder functions take exactly one argument, the
message object to encode.  They usually extract the payload, encode
it, and reset the payload to this newly encoded value.  They should also
set the \code{Content-Transfer-Encoding:} header as appropriate.

Here are the encoding functions provided:

\begin{funcdesc}{encode_quopri}{msg}
Encodes the payload into \emph{Quoted-Printable} form and sets the
\code{Content-Transfer-Encoding:} header to
\code{quoted-printable}\footnote{Note that encoding with
\method{encode_quopri()} also encodes all tabs and space characters in
the data.}.
This is a good encoding to use when most of your payload is normal
printable data, but contains a few unprintable characters.
\end{funcdesc}

\begin{funcdesc}{encode_base64}{msg}
Encodes the payload into \emph{Base64} form and sets the
\code{Content-Transfer-Encoding:} header to
\code{base64}.  This is a good encoding to use when most of your payload
is unprintable data since it is a more compact form than
Quoted-Printable.  The drawback of Base64 encoding is that it
renders the text non-human readable.
\end{funcdesc}

\begin{funcdesc}{encode_7or8bit}{msg}
This doesn't actually modify the message's payload, but it does set
the \code{Content-Transfer-Encoding:} header to either \code{7bit} or
\code{8bit} as appropriate, based on the payload data.
\end{funcdesc}

\begin{funcdesc}{encode_noop}{msg}
This does nothing; it doesn't even set the
\code{Content-Transfer-Encoding:} header.
\end{funcdesc}
