\section{\module{bz2} ---
         Compression compatible with \program{bzip2}}

\declaremodule{builtin}{bz2}
\modulesynopsis{Interface to compression and decompression
                routines compatible with \program{bzip2}.}
\moduleauthor{Gustavo Niemeyer}{niemeyer@conectiva.com}
\sectionauthor{Gustavo Niemeyer}{niemeyer@conectiva.com}

\versionadded{2.3}

This module provides a comprehensive interface for the bz2 compression library.
It implements a complete file interface, one-shot (de)compression functions,
and types for sequential (de)compression.

Here is a resume of the features offered by the bz2 module:

\begin{itemize}
\item \class{BZ2File} class implements a complete file interface, including
      \method{readline()}, \method{readlines()},
      \method{writelines()}, \method{seek()}, etc;
\item \class{BZ2File} class implements emulated \method{seek()} support;
\item \class{BZ2File} class implements universal newline support;
\item \class{BZ2File} class offers an optimized line iteration using
      the readahead algorithm borrowed from file objects;
\item Sequential (de)compression supported by \class{BZ2Compressor} and
      \class{BZ2Decompressor} classes;
\item One-shot (de)compression supported by \function{compress()} and
      \function{decompress()} functions;
\item Thread safety uses individual locking mechanism;
\item Complete inline documentation;
\end{itemize}


\subsection{(De)compression of files}

Handling of compressed files is offered by the \class{BZ2File} class.

\begin{classdesc}{BZ2File}{filename\optional{, mode\optional{,
                           buffering\optional{, compresslevel}}}}
Open a bz2 file. Mode can be either \code{'r'} or \code{'w'}, for reading 
(default) or writing. When opened for writing, the file will be created if
it doesn't exist, and truncated otherwise. If \var{buffering} is given,
\code{0} means unbuffered, and larger numbers specify the buffer size;
the default is \code{0}. If
\var{compresslevel} is given, it must be a number between \code{1} and
\code{9}; the default is \code{9}.
Add a \character{U} to mode to open the file for input with universal newline
support. Any line ending in the input file will be seen as a
\character{\e n} in Python.  Also, a file so opened gains the
attribute \member{newlines}; the value for this attribute is one of
\code{None} (no newline read yet), \code{'\e r'}, \code{'\e n'},
\code{'\e r\e n'} or a tuple containing all the newline types
seen. Universal newlines are available only when reading.
Instances support iteration in the same way as normal \class{file}
instances.
\end{classdesc}

\begin{methoddesc}[BZ2File]{close}{}
Close the file. Sets data attribute \member{closed} to true. A closed file
cannot be used for further I/O operations. \method{close()} may be called
more than once without error.
\end{methoddesc}

\begin{methoddesc}[BZ2File]{read}{\optional{size}}
Read at most \var{size} uncompressed bytes, returned as a string. If the
\var{size} argument is negative or omitted, read until EOF is reached.
\end{methoddesc}

\begin{methoddesc}[BZ2File]{readline}{\optional{size}}
Return the next line from the file, as a string, retaining newline.
A non-negative \var{size} argument limits the maximum number of bytes to
return (an incomplete line may be returned then). Return an empty
string at EOF.
\end{methoddesc}

\begin{methoddesc}[BZ2File]{readlines}{\optional{size}}
Return a list of lines read. The optional \var{size} argument, if given,
is an approximate bound on the total number of bytes in the lines returned.
\end{methoddesc}

\begin{methoddesc}[BZ2File]{xreadlines}{}
For backward compatibility. \class{BZ2File} objects now include the
performance optimizations previously implemented in the
\module{xreadlines} module.
\deprecated{2.3}{This exists only for compatibility with the method by
                 this name on \class{file} objects, which is
                 deprecated.  Use \code{for line in file} instead.}
\end{methoddesc}

\begin{methoddesc}[BZ2File]{seek}{offset\optional{, whence}}
Move to new file position. Argument \var{offset} is a byte count. Optional
argument \var{whence} defaults to \code{0} (offset from start of file,
offset should be \code{>= 0}); other values are \code{1} (move relative to
current position, positive or negative), and \code{2} (move relative to end
of file, usually negative, although many platforms allow seeking beyond
the end of a file).

Note that seeking of bz2 files is emulated, and depending on the parameters
the operation may be extremely slow.
\end{methoddesc}

\begin{methoddesc}[BZ2File]{tell}{}
Return the current file position, an integer (may be a long integer).
\end{methoddesc}

\begin{methoddesc}[BZ2File]{write}{data}
Write string \var{data} to file. Note that due to buffering, \method{close()}
may be needed before the file on disk reflects the data written.
\end{methoddesc}

\begin{methoddesc}[BZ2File]{writelines}{sequence_of_strings}
Write the sequence of strings to the file. Note that newlines are not added.
The sequence can be any iterable object producing strings. This is equivalent
to calling write() for each string.
\end{methoddesc}


\subsection{Sequential (de)compression}

Sequential compression and decompression is done using the classes
\class{BZ2Compressor} and \class{BZ2Decompressor}.

\begin{classdesc}{BZ2Compressor}{\optional{compresslevel}}
Create a new compressor object. This object may be used to compress
data sequentially. If you want to compress data in one shot, use the
\function{compress()} function instead. The \var{compresslevel} parameter,
if given, must be a number between \code{1} and \code{9}; the default
is \code{9}.
\end{classdesc}

\begin{methoddesc}[BZ2Compressor]{compress}{data}
Provide more data to the compressor object. It will return chunks of compressed
data whenever possible. When you've finished providing data to compress, call
the \method{flush()} method to finish the compression process, and return what
is left in internal buffers.
\end{methoddesc}

\begin{methoddesc}[BZ2Compressor]{flush}{}
Finish the compression process and return what is left in internal buffers. You
must not use the compressor object after calling this method.
\end{methoddesc}

\begin{classdesc}{BZ2Decompressor}{}
Create a new decompressor object. This object may be used to decompress
data sequentially. If you want to decompress data in one shot, use the
\function{decompress()} function instead.
\end{classdesc}

\begin{methoddesc}[BZ2Decompressor]{decompress}{data}
Provide more data to the decompressor object. It will return chunks of
decompressed data whenever possible. If you try to decompress data after the
end of stream is found, \exception{EOFError} will be raised. If any data was
found after the end of stream, it'll be ignored and saved in
\member{unused\_data} attribute.
\end{methoddesc}


\subsection{One-shot (de)compression}

One-shot compression and decompression is provided through the
\function{compress()} and \function{decompress()} functions.

\begin{funcdesc}{compress}{data\optional{, compresslevel}}
Compress \var{data} in one shot. If you want to compress data sequentially,
use an instance of \class{BZ2Compressor} instead. The \var{compresslevel}
parameter, if given, must be a number between \code{1} and \code{9};
the default is \code{9}.
\end{funcdesc}

\begin{funcdesc}{decompress}{data}
Decompress \var{data} in one shot. If you want to decompress data
sequentially, use an instance of \class{BZ2Decompressor} instead.
\end{funcdesc}
