\section{\module{cookielib} ---
         Cookie handling for HTTP clients}

\declaremodule{standard}{cookielib}
\moduleauthor{John J. Lee}{jjl@pobox.com}
\sectionauthor{John J. Lee}{jjl@pobox.com}

\modulesynopsis{Cookie handling for HTTP clients}

The \module{cookielib} module defines classes for automatic handling
of HTTP cookies.  It is useful for accessing web sites that require
small pieces of data -- \dfn{cookies} -- to be set on the client
machine by an HTTP response from a web server, and then returned to
the server in later HTTP requests.

Both the regular Netscape cookie protocol and the protocol defined by
\rfc{2965} are handled.  RFC 2965 handling is switched off by default.
\rfc{2109} cookies are parsed as Netscape cookies and subsequently
treated as RFC 2965 cookies.  Note that the great majority of cookies
on the Internet are Netscape cookies.  \module{cookielib} attempts to
follow the de-facto Netscape cookie protocol (which differs
substantially from that set out in the original Netscape
specification), including taking note of the \code{max-age} and
\code{port} cookie-attributes introduced with RFC 2109.  \note{The
various named parameters found in \mailheader{Set-Cookie} and
\mailheader{Set-Cookie2} headers (eg. \code{domain} and
\code{expires}) are conventionally referred to as \dfn{attributes}.
To distinguish them from Python attributes, the documentation for this
module uses the term \dfn{cookie-attribute} instead}.


The module defines the following exception:

\begin{excdesc}{LoadError}
Instances of \class{FileCookieJar} raise this exception on failure to
load cookies from a file.
\end{excdesc}


The following classes are provided:

\begin{classdesc}{CookieJar}{policy=\constant{None}}
\var{policy} is an object implementing the \class{CookiePolicy}
interface.

The \class{CookieJar} class stores HTTP cookies.  It extracts cookies
from HTTP requests, and returns them in HTTP responses.
\class{CookieJar} instances automatically expire contained cookies
when necessary.  Subclasses are also responsible for storing and
retrieving cookies from a file or database.
\end{classdesc}

\begin{classdesc}{FileCookieJar}{filename, delayload=\constant{None},
 policy=\constant{None}}
\var{policy} is an object implementing the \class{CookiePolicy}
interface.  For the other arguments, see the documentation for the
corresponding attributes.

A \class{CookieJar} which can load cookies from, and perhaps save
cookies to, a file on disk.  Cookies are \strong{NOT} loaded from the
named file until either the \method{load()} or \method{revert()}
method is called.  Subclasses of this class are documented in section
\ref{file-cookie-jar-classes}.
\end{classdesc}

\begin{classdesc}{CookiePolicy}{}
This class is responsible for deciding whether each cookie should be
accepted from / returned to the server.
\end{classdesc}

\begin{classdesc}{DefaultCookiePolicy}{
    blocked_domains=\constant{None},
    allowed_domains=\constant{None},
    netscape=\constant{True}, rfc2965=\constant{False},
    hide_cookie2=\constant{False},
    strict_domain=\constant{False},
    strict_rfc2965_unverifiable=\constant{True},
    strict_ns_unverifiable=\constant{False},
    strict_ns_domain=\constant{DefaultCookiePolicy.DomainLiberal},
    strict_ns_set_initial_dollar=\constant{False},
    strict_ns_set_path=\constant{False}
  }

Constructor arguments should be passed as keyword arguments only.
\var{blocked_domains} is a sequence of domain names that we never
accept cookies from, nor return cookies to. \var{allowed_domains} if
not \constant{None}, this is a sequence of the only domains for which
we accept and return cookies.  For all other arguments, see the
documentation for \class{CookiePolicy} and \class{DefaultCookiePolicy}
objects.

\class{DefaultCookiePolicy} implements the standard accept / reject
rules for Netscape and RFC 2965 cookies.  RFC 2109 cookies
(ie. cookies received in a \mailheader{Set-Cookie} header with a
version cookie-attribute of 1) are treated according to the RFC 2965
rules.  \class{DefaultCookiePolicy} also provides some parameters to
allow some fine-tuning of policy.
\end{classdesc}

\begin{classdesc}{Cookie}{}
This class represents Netscape, RFC 2109 and RFC 2965 cookies.  It is
not expected that users of \module{cookielib} construct their own
\class{Cookie} instances.  Instead, if necessary, call
\method{make_cookies()} on a \class{CookieJar} instance.
\end{classdesc}

\begin{seealso}

\seemodule{urllib2}{URL opening with automatic cookie handling.}

\seemodule{Cookie}{HTTP cookie classes, principally useful for
server-side code.  The \module{cookielib} and \module{Cookie} modules
do not depend on each other.}

\seeurl{http://wwwsearch.sf.net/ClientCookie/}{Extensions to this
module, including a class for reading Microsoft Internet Explorer
cookies on Windows.}

\seeurl{http://www.netscape.com/newsref/std/cookie_spec.html}{The
specification of the original Netscape cookie protocol.  Though this
is still the dominant protocol, the 'Netscape cookie protocol'
implemented by all the major browsers (and \module{cookielib}) only
bears a passing resemblance to the one sketched out in
\code{cookie_spec.html}.}

\seerfc{2109}{HTTP State Management Mechanism}{Obsoleted by RFC 2965.
Uses \mailheader{Set-Cookie} with version=1.}

\seerfc{2965}{HTTP State Management Mechanism}{The Netscape protocol
with the bugs fixed.  Uses \mailheader{Set-Cookie2} in place of
\mailheader{Set-Cookie}.  Not widely used.}

\seeurl{http://kristol.org/cookie/errata.html}{Unfinished errata to
RFC 2965.}

\seerfc{2964}{Use of HTTP State Management}{}

\end{seealso}


\subsection{CookieJar and FileCookieJar Objects \label{cookie-jar-objects}}

\class{CookieJar} objects support the iterator protocol for iterating
over contained \class{Cookie} objects.

\class{CookieJar} has the following methods:

\begin{methoddesc}[CookieJar]{add_cookie_header}{request}
Add correct \mailheader{Cookie} header to \var{request}.

If policy allows (ie. the \member{rfc2965} and \member{hide_cookie2}
attributes of the \class{CookieJar}'s \class{CookiePolicy} instance
are true and false respectively), the \mailheader{Cookie2} header is
also added when appropriate.

The \var{request} object (usually a \class{urllib2.Request} instance)
must support the methods \method{get_full_url()}, \method{get_host()},
\method{get_type()}, \method{unverifiable()},
\method{get_origin_req_host()}, \method{has_header()},
\method{get_header()}, \method{header_items()}, and
\method{add_unredirected_header()},as documented by \module{urllib2}.
\end{methoddesc}

\begin{methoddesc}[CookieJar]{extract_cookies}{response, request}
Extract cookies from HTTP \var{response} and store them in the
\class{CookieJar}, where allowed by policy.

The \class{CookieJar} will look for allowable \mailheader{Set-Cookie}
and \mailheader{Set-Cookie2} headers in the \var{response} argument,
and store cookies as appropriate (subject to the
\method{CookiePolicy.set_ok()} method's approval).

The \var{response} object (usually the result of a call to
\method{urllib2.urlopen()}, or similar) should support an
\method{info()} method, which returns an object with a
\method{getallmatchingheaders()} method (usually a
\class{mimetools.Message} instance).

The \var{request} object (usually a \class{urllib2.Request} instance)
must support the methods \method{get_full_url()}, \method{get_host()},
\method{unverifiable()}, and \method{get_origin_req_host()}, as
documented by \module{urllib2}.  The request is used to set default
values for cookie-attributes as well as for checking that the cookie
is allowed to be set.
\end{methoddesc}

\begin{methoddesc}[CookieJar]{set_policy}{policy}
Set the \class{CookiePolicy} instance to be used.
\end{methoddesc}

\begin{methoddesc}[CookieJar]{make_cookies}{response, request}
Return sequence of \class{Cookie} objects extracted from
\var{response} object.

See the documentation for \method{extract_cookies} for the interfaces
required of the \var{response} and \var{request} arguments.
\end{methoddesc}

\begin{methoddesc}[CookieJar]{set_cookie_if_ok}{cookie, request}
Set a \class{Cookie} if policy says it's OK to do so.
\end{methoddesc}

\begin{methoddesc}[CookieJar]{set_cookie}{cookie}
Set a \class{Cookie}, without checking with policy to see whether or
not it should be set.
\end{methoddesc}

\begin{methoddesc}[CookieJar]{clear}{\optional{domain\optional{,
      path\optional{, name}}}}
Clear some cookies.

If invoked without arguments, clear all cookies.  If given a single
argument, only cookies belonging to that \var{domain} will be removed.
If given two arguments, cookies belonging to the specified
\var{domain} and URL \var{path} are removed.  If given three
arguments, then the cookie with the specified \var{domain}, \var{path}
and \var{name} is removed.

Raises \exception{KeyError} if no matching cookie exists.
\end{methoddesc}

\begin{methoddesc}[CookieJar]{clear_session_cookies}{}
Discard all session cookies.

Discards all contained cookies that have a true \member{discard}
attribute (usually because they had either no \code{max-age} or
\code{expires} cookie-attribute, or an explicit \code{discard}
cookie-attribute).  For interactive browsers, the end of a session
usually corresponds to closing the browser window.

Note that the \method{save()} method won't save session cookies
anyway, unless you ask otherwise by passing a true
\var{ignore_discard} argument.
\end{methoddesc}

\class{FileCookieJar} implements the following additional methods:

\begin{methoddesc}[FileCookieJar]{save}{filename=\constant{None},
    ignore_discard=\constant{False}, ignore_expires=\constant{False}}
Save cookies to a file.

This base class raises \class{NotImplementedError}.  Subclasses may
leave this method unimplemented.

\var{filename} is the name of file in which to save cookies.  If
\var{filename} is not specified, \member{self.filename} is used (whose
default is the value passed to the constructor, if any); if
\member{self.filename} is \constant{None}, \exception{ValueError} is
raised.

\var{ignore_discard}: save even cookies set to be discarded.
\var{ignore_expires}: save even cookies that have expired

The file is overwritten if it already exists, thus wiping all the
cookies it contains.  Saved cookies can be restored later using the
\method{load()} or \method{revert()} methods.
\end{methoddesc}

\begin{methoddesc}[FileCookieJar]{load}{filename=\constant{None},
    ignore_discard=\constant{False}, ignore_expires=\constant{False}}
Load cookies from a file.

Old cookies are kept unless overwritten by newly loaded ones.

Arguments are as for \method{save()}.

The named file must be in the format understood by the class, or
\exception{LoadError} will be raised.
\end{methoddesc}

\begin{methoddesc}[FileCookieJar]{revert}{filename=\constant{None},
    ignore_discard=\constant{False}, ignore_expires=\constant{False}}
Clear all cookies and reload cookies from a saved file.

Raises \exception{cookielib.LoadError} or \exception{IOError} if
reversion is not successful; the object's state will not be altered if
this happens.
\end{methoddesc}

\class{FileCookieJar} instances have the following public attributes:

\begin{memberdesc}{filename}
Filename of default file in which to keep cookies.  This attribute may
be assigned to.
\end{memberdesc}

\begin{memberdesc}{delayload}
If true, load cookies lazily from disk.  This attribute should not be
assigned to.  This is only a hint, since this only affects
performance, not behaviour (unless the cookies on disk are changing).
A \class{CookieJar} object may ignore it.  None of the
\class{FileCookieJar} classes included in the standard library lazily
loads cookies.
\end{memberdesc}


\subsection{FileCookieJar subclasses and co-operation with web browsers
  \label{file-cookie-jar-classes}}

The following \class{CookieJar} subclasses are provided for reading
and writing .  Further \class{CookieJar} subclasses, including one
that reads Microsoft Internet Explorer cookies, are available at
\url{http://wwwsearch.sf.net/ClientCookie/}.

\begin{classdesc}{MozillaCookieJar}{filename, delayload=\constant{None},
 policy=\constant{None}}
A \class{FileCookieJar} that can load from and save cookies to disk in
the Mozilla \code{cookies.txt} file format (which is also used by the
Lynx and Netscape browsers).  \note{This loses information about RFC
2965 cookies, and also about newer or non-standard cookie-attributes
such as \code{port}.}

\warning{Back up your cookies before saving if you have cookies whose
loss / corruption would be inconvenient (there are some subtleties
which may lead to slight changes in the file over a load / save
round-trip).}

Also note that cookies saved while Mozilla is running will get
clobbered by Mozilla.
\end{classdesc}

\begin{classdesc}{LWPCookieJar}{filename, delayload=\constant{None},
 policy=\constant{None}}
A \class{FileCookieJar} that can load from and save cookies to disk in
format compatible with the libwww-perl library's \code{Set-Cookie3}
file format.  This is convenient if you want to store cookies in a
human-readable file.
\end{classdesc}


\subsection{CookiePolicy Objects \label{cookie-policy-objects}}

Objects implementing the \class{CookiePolicy} interface have the
following methods:

\begin{methoddesc}[CookiePolicy]{set_ok}{cookie, request}
Return boolean value indicating whether cookie should be accepted from server.

\var{cookie} is a \class{cookielib.Cookie} instance.  \var{request} is
an object implementing the interface defined by the documentation for
\method{CookieJar.extract_cookies()}.
\end{methoddesc}

\begin{methoddesc}[CookiePolicy]{return_ok}{cookie, request}
Return boolean value indicating whether cookie should be returned to server.

\var{cookie} is a \class{cookielib.Cookie} instance.  \var{request} is
an object implementing the interface defined by the documentation for
\method{CookieJar.add_cookie_header()}.
\end{methoddesc}

\begin{methoddesc}[CookiePolicy]{domain_return_ok}{domain, request}
Return false if cookies should not be returned, given cookie domain.

This method is an optimization.  It removes the need for checking
every cookie with a particular domain (which might involve reading
many files).  Returning true from \method{domain_return_ok()} and
\method{path_return_ok()} leaves all the work to \method{return_ok()}.

If \method{domain_return_ok()} returns true for the cookie domain,
\method{path_return_ok()} is called for the cookie path.  Otherwise,
\method{path_return_ok()} and \method{return_ok()} are never called
for that cookie domain.  If \method{path_return_ok()} returns true,
\method{return_ok()} is called with the \class{Cookie} object itself
for a full check.  Otherwise, \method{return_ok()} is never called for
that cookie path.

Note that \method{domain_return_ok()} is called for every
\emph{cookie} domain, not just for the \emph{request} domain.  For
example, the function might be called with both \code{".example.com"}
and \code{"www.example.com"} if the request domain is
\code{"www.example.com"}.  The same goes for
\method{path_return_ok()}.

The \var{request} argument is as documented for \method{return_ok()}.
\end{methoddesc}

\begin{methoddesc}[CookiePolicy]{path_return_ok}{path, request}
Return false if cookies should not be returned, given cookie path.

See the documentation for \method{domain_return_ok()}.
\end{methoddesc}


In addition to implementing the methods above, implementations of the
\class{CookiePolicy} interface must also supply the following
attributes, indicating which protocols should be used, and how.  All
of these attributes may be assigned to.

\begin{memberdesc}{netscape}
Implement Netscape protocol.
\end{memberdesc}
\begin{memberdesc}{rfc2965}
Implement RFC 2965 protocol.
\end{memberdesc}
\begin{memberdesc}{hide_cookie2}
Don't add \mailheader{Cookie2} header to requests (the presence of
this header indicates to the server that we understand RFC 2965
cookies).
\end{memberdesc}

The most useful way to define a \class{CookiePolicy} class is by
subclassing from \class{DefaultCookiePolicy} and overriding some or
all of the methods above.  \class{CookiePolicy} itself may be used as
a 'null policy' to allow setting and receiving any and all cookies
(this is unlikely to be useful).


\subsection{DefaultCookiePolicy Objects \label{default-cookie-policy-objects}}

Implements the standard rules for accepting and returning cookies.

Both RFC 2965 and Netscape cookies are covered.  RFC 2965 handling is
switched off by default.

The easiest way to provide your own policy is to override this class
and call its methods in your overridden implementations before adding
your own additional checks:

\begin{verbatim}
import cookielib
class MyCookiePolicy(cookielib.DefaultCookiePolicy):
    def set_ok(self, cookie, request):
        if not cookielib.DefaultCookiePolicy.set_ok(self, cookie, request):
            return False
        if i_dont_want_to_store_this_cookie(cookie):
            return False
        return True
\end{verbatim}

In addition to the features required to implement the
\class{CookiePolicy} interface, this class allows you to block and
allow domains from setting and receiving cookies.  There are also some
strictness switches that allow you to tighten up the rather loose
Netscape protocol rules a little bit (at the cost of blocking some
benign cookies).

A domain blacklist and whitelist is provided (both off by default).
Only domains not in the blacklist and present in the whitelist (if the
whitelist is active) participate in cookie setting and returning.  Use
the \var{blocked_domains} constructor argument, and
\method{blocked_domains()} and \method{set_blocked_domains()} methods
(and the corresponding argument and methods for
\var{allowed_domains}).  If you set a whitelist, you can turn it off
again by setting it to \constant{None}.

Domains in block or allow lists that do not start with a dot must
equal the cookie domain to be matched.  For example,
\code{"example.com"} matches a blacklist entry of
\code{"example.com"}, but \code{"www.example.com"} does not.  Domains
that do start with a dot are matched by more specific domains too.
For example, both \code{"www.example.com"} and
\code{"www.coyote.example.com"} match \code{".example.com"} (but
\code{"example.com"} itself does not).  IP addresses are an exception,
and must match exactly.  For example, if blocked_domains contains
\code{"192.168.1.2"} and \code{".168.1.2"}, 192.168.1.2 is blocked,
but 193.168.1.2 is not.

\class{DefaultCookiePolicy} implements the following additional
methods:

\begin{methoddesc}[DefaultCookiePolicy]{blocked_domains}{}
Return the sequence of blocked domains (as a tuple).
\end{methoddesc}

\begin{methoddesc}[DefaultCookiePolicy]{set_blocked_domains}
  {blocked_domains}
Set the sequence of blocked domains.
\end{methoddesc}

\begin{methoddesc}[DefaultCookiePolicy]{is_blocked}{domain}
Return whether \var{domain} is on the blacklist for setting or
receiving cookies.
\end{methoddesc}

\begin{methoddesc}[DefaultCookiePolicy]{allowed_domains}{}
Return \constant{None}, or the sequence of allowed domains (as a tuple).
\end{methoddesc}

\begin{methoddesc}[DefaultCookiePolicy]{set_allowed_domains}
  {allowed_domains}
Set the sequence of allowed domains, or \constant{None}.
\end{methoddesc}

\begin{methoddesc}[DefaultCookiePolicy]{is_not_allowed}{domain}
Return whether \var{domain} is not on the whitelist for setting or
receiving cookies.
\end{methoddesc}

\class{DefaultCookiePolicy} instances have the following attributes,
which are all initialised from the constructor arguments of the same
name, and which may all be assigned to.

General strictness switches:

\begin{memberdesc}{strict_domain}
Don't allow sites to set two-component domains with country-code
top-level domains like \code{.co.uk}, \code{.gov.uk},
\code{.co.nz}.etc.  This is far from perfect and isn't guaranteed to
work!
\end{memberdesc}

RFC 2965 protocol strictness switches:

\begin{memberdesc}{strict_rfc2965_unverifiable}
Follow RFC 2965 rules on unverifiable transactions (usually, an
unverifiable transaction is one resulting from a redirect or a request
for an image hosted on another site).  If this is false, cookies are
\emph{never} blocked on the basis of verifiability
\end{memberdesc}

Netscape protocol strictness switches:

\begin{memberdesc}{strict_ns_unverifiable}
apply RFC 2965 rules on unverifiable transactions even to Netscape
cookies
\end{memberdesc}
\begin{memberdesc}{strict_ns_domain}
Flags indicating how strict to be with domain-matching rules for
Netscape cookies.  See below for acceptable values.
\end{memberdesc}
\begin{memberdesc}{strict_ns_set_initial_dollar}
Ignore cookies in Set-Cookie: headers that have names starting with
\code{'\$'}.
\end{memberdesc}
\begin{memberdesc}{strict_ns_set_path}
Don't allow setting cookies whose path doesn't path-match request URI.
\end{memberdesc}

\member{strict_ns_domain} is a collection of flags.  Its value is
constructed by or-ing together (for example,
\code{DomainStrictNoDots|DomainStrictNonDomain} means both flags are
set).

\begin{memberdesc}{DomainStrictNoDots}
When setting cookies, the 'host prefix' must not contain a dot
(eg. \code{www.foo.bar.com} can't set a cookie for \code{.bar.com},
because \code{www.foo} contains a dot).
\end{memberdesc}
\begin{memberdesc}{DomainStrictNonDomain}
Cookies that did not explicitly specify a \code{domain}
cookie-attribute can only be returned to a domain equal to the domain
that set the cookie (eg. \code{spam.example.com} won't be returned
cookies from \code{example.com} that had no \code{domain}
cookie-attribute).
\end{memberdesc}
\begin{memberdesc}{DomainRFC2965Match}
When setting cookies, require a full RFC 2965 domain-match.
\end{memberdesc}

The following attributes are provided for convenience, and are the
most useful combinations of the above flags:

\begin{memberdesc}{DomainLiberal}
Equivalent to 0 (ie. all of the above Netscape domain strictness flags
switched off).
\end{memberdesc}
\begin{memberdesc}{DomainStrict}
Equivalent to \code{DomainStrictNoDots|DomainStrictNonDomain}.
\end{memberdesc}


\subsection{Cookie Objects \label{cookie-jar-objects}}

\class{Cookie} instances have Python attributes roughly corresponding
to the standard cookie-attributes specified in the various cookie
standards.  The correspondence is not one-to-one, because there are
complicated rules for assigning default values, and because the
\code{max-age} and \code{expires} cookie-attributes contain equivalent
information.

Assignment to these attributes should not be necessary other than in
rare circumstances in a \class{CookiePolicy} method.  The class does
not enforce internal consistency, so you should know what you're
doing if you do that.

\begin{memberdesc}[Cookie]{version}
Integer or \constant{None}.  Netscape cookies have version 0.  RFC
2965 and RFC 2109 cookies have version 1.
\end{memberdesc}
\begin{memberdesc}[Cookie]{name}
Cookie name (a string).
\end{memberdesc}
\begin{memberdesc}[Cookie]{value}
Cookie value (a string), or \constant{None}.
\end{memberdesc}
\begin{memberdesc}[Cookie]{port}
String representing a port or a set of ports (eg. '80', or '80,8080'),
or \constant{None}.
\end{memberdesc}
\begin{memberdesc}[Cookie]{path}
Cookie path (a string, eg. \code{'/acme/rocket_launchers'}).
\end{memberdesc}
\begin{memberdesc}[Cookie]{secure}
True if cookie should only be returned over a secure connection.
\end{memberdesc}
\begin{memberdesc}[Cookie]{expires}
Integer expiry date in seconds since epoch, or \constant{None}.  See
also the \method{is_expired()} method.
\end{memberdesc}
\begin{memberdesc}[Cookie]{discard}
True if this is a session cookie.
\end{memberdesc}
\begin{memberdesc}[Cookie]{comment}
String comment from the server explaining the function of this cookie,
or \constant{None}.
\end{memberdesc}
\begin{memberdesc}[Cookie]{comment_url}
URL linking to a comment from the server explaining the function of
this cookie, or \constant{None}.
\end{memberdesc}

\begin{memberdesc}[Cookie]{port_specified}
True if a port or set of ports was explicitly specified by the server
(in the \mailheader{Set-Cookie} / \mailheader{Set-Cookie2} header).
\end{memberdesc}
\begin{memberdesc}[Cookie]{domain_specified}
True if a domain was explicitly specified by the server.
\end{memberdesc}
\begin{memberdesc}[Cookie]{domain_initial_dot}
True if the domain explicitly specified by the server began with a
dot (\code{'.'}).
\end{memberdesc}

Cookies may have additional non-standard cookie-attributes.  These may
be accessed using the following methods:

\begin{methoddesc}[Cookie]{has_nonstandard_attr}{name}
Return true if cookie has the named cookie-attribute.
\end{methoddesc}
\begin{methoddesc}[Cookie]{get_nonstandard_attr}{name, default=\constant{None}}
If cookie has the named cookie-attribute, return its value.
Otherwise, return \var{default}.
\end{methoddesc}
\begin{methoddesc}[Cookie]{set_nonstandard_attr}{name, value}
Set the value of the named cookie-attribute.
\end{methoddesc}

The \class{Cookie} class also defines the following method:

\begin{methoddesc}[Cookie]{is_expired}{\optional{now=\constant{None}}}
True if cookie has passed the time at which the server requested it
should expire.  If \var{now} is given (in seconds since the epoch),
return whether the cookie has expired at the specified time.
\end{methoddesc}


\subsection{Examples \label{cookielib-examples}}

The first example shows the most common usage of \module{cookielib}:

\begin{verbatim}
import cookielib, urllib2
cj = cookielib.CookieJar()
opener = urllib2.build_opener(urllib2.HTTPCookieProcessor(cj))
r = opener.open("http://example.com/")
\end{verbatim}

This example illustrates how to open a URL using your Netscape,
Mozilla, or Lynx cookies (assumes \UNIX{}/Netscape convention for
location of the cookies file):

\begin{verbatim}
import os, cookielib, urllib2
cj = cookielib.MozillaCookieJar()
cj.load(os.path.join(os.environ["HOME"], ".netscape/cookies.txt"))
opener = urllib2.build_opener(urllib2.HTTPCookieProcessor(cj))
r = opener.open("http://example.com/")
\end{verbatim}

The next example illustrates the use of \class{DefaultCookiePolicy}.
Turn on RFC 2965 cookies, be more strict about domains when setting
and returning Netscape cookies, and block some domains from setting
cookies or having them returned:

\begin{verbatim}
import urllib2
from cookielib import CookieJar, DefaultCookiePolicy
policy = DefaultCookiePolicy(
    rfc2965=True, strict_ns_domain=Policy.DomainStrict,
    blocked_domains=["ads.net", ".ads.net"])
cj = CookieJar(policy)
opener = urllib2.build_opener(urllib2.HTTPCookieProcessor(cj))
r = opener.open("http://example.com/")
\end{verbatim}
