% XXX what order should the types be discussed in?

\section{\module{datetime} ---
         Basic date and time types}

\declaremodule{builtin}{datetime}
\modulesynopsis{Basic date and time types.}
\moduleauthor{Tim Peters}{tim@zope.com}
\sectionauthor{Tim Peters}{tim@zope.com}
\sectionauthor{A.M. Kuchling}{amk@amk.ca}
\sectionauthor{Raymond D. Hettinger}{python@rcn.com}

\versionadded{2.3}


The \module{datetime} module supplies classes for manipulating dates
and times in both simple and complex ways.  While date and time
arithmetic is supported, the focus of the implementation is on
efficient field extraction, for output formatting and manipulation.

There are two kinds of date and time objects: ``naive'' and ``aware''.
This distinction refers to whether the object has any notion of time
zone, daylight savings time, or other kind of algorithmic or political
time adjustment.  Whether a {naive} \class{datetime} object represents
Coordinated Universal Time (UTC), local time, or time in some other
timezone is purely up to the program, just like it's up to the program
whether a particular number represents meters, miles, or mass.  Naive
\class{datetime} objects are easy to understand and to work with, at
the cost of ignoring some aspects of reality.

For applications requiring more, ``aware'' \class{datetime} subclasses add an
optional time zone information object to the basic naive classes.
These \class{tzinfo} objects capture information about the offset from
UTC time, the time zone name, and whether Daylight Savings Time is in
effect.  Note that no concrete \class{tzinfo} classes are supplied by
the \module{datetime} module.  Instead, they provide a framework for
incorporating the level of detail an app may require.  The rules for
time adjustment across the world are more political than rational, and
there is no standard suitable for every app.

The \module{datetime} module exports the following constants:

\begin{datadesc}{MINYEAR}
  The smallest year number allowed in a \class{date},
  \class{datetime}, or \class{datetimetz} object.  \constant{MINYEAR}
  is \code{1}.
\end{datadesc}

\begin{datadesc}{MAXYEAR}
  The largest year number allowed in a \class{date}, \class{datetime},
  or \class{datetimetz} object.  \constant{MAXYEAR} is \code{9999}.
\end{datadesc}

\begin{seealso}
  \seemodule{calendar}{General calendar related functions.}
  \seemodule{time}{Time access and conversions.}
\end{seealso}

\subsection{Available Types}

\begin{classdesc*}{date}
  An idealized naive date, assuming the current Gregorian calendar
  always was, and always will be, in effect.
  Attributes: \member{year}, \member{month}, and \member{day}.
\end{classdesc*}

\begin{classdesc*}{time}
  An idealized naive time, independent of any particular day, assuming
  that every day has exactly 24*60*60 seconds (there is no notion
  of "leap seconds" here).
  Attributes: \member{hour}, \member{minute}, \member{second}, and
              \member{microsecond}
\end{classdesc*}

\begin{classdesc*}{datetime}
  A combination of a naive date and a naive time.
  Attributes: \member{year}, \member{month}, \member{day},
              \member{hour}, \member{minute}, \member{second},
              and \member{microsecond}.
\end{classdesc*}

\begin{classdesc*}{timedelta}
  A duration, expressing the difference between two \class{date},
  \class{time}, or \class{datetime} instances, to microsecond
  resolution.
\end{classdesc*}

\begin{classdesc*}{tzinfo}
  An abstract base class for time zone information objects.  These
  are used by the  \class{datetimetz}  and \class{timetz} classes to
  provided a customizable notion of time adjustment (for example, to
  account for time zone and/or daylight savings time).
\end{classdesc*}

\begin{classdesc*}{timetz}
  An aware subclass of \class{time}, supporting a customizable notion of
  time adjustment.
\end{classdesc*}

\begin{classdesc*}{datetimetz}
  An aware subclass of \class{datetime}, supporting a customizable notion of
  time adjustment.
\end{classdesc*}

Objects of these types are immutable.

Objects of the \class{date}, \class{datetime}, and \class{time} types
are always naive.

An object \var{d} of type \class{timetz} or \class{datetimetz} may be
naive or aware.  \var{d} is aware if \code{\var{d}.tzinfo} is not
\code{None}, and \code{\var{d}.tzinfo.utcoffset(\var{d})} does not return
\code{None}.  If \code{\var{d}.tzinfo} is \code{None}, or if
\code{\var{d}.tzinfo} is not \code{None} but
\code{\var{d}.tzinfo.utcoffset(\var{d})} returns \code{None}, \var{d}
is naive.

The distinction between naive and aware doesn't apply to
\code{timedelta} objects.

Subclass relationships:

\begin{verbatim}
object
    timedelta
    tzinfo
    time
        timetz
    date
        datetime
            datetimetz
\end{verbatim}

\subsection{\class{timedelta} Objects \label{datetime-timedelta}}

A \class{timedelta} object represents a duration, the difference
between two dates or times.

\begin{classdesc}{timedelta}{days=0, seconds=0, microseconds=0,
                             milliseconds=0, minutes=0, hours=0, weeks=0}

  All arguments are optional.  Arguments may be ints, longs, or floats,
  and may be positive or negative.

  Only \var{days}, \var{seconds} and \var{microseconds} are stored
  internally.  Arguments are converted to those units:

\begin{verbatim}
A millisecond is converted to 1000 microseconds.
A minute is converted to 60 seconds.
An hour is converted to 3600 seconds.
A week is converted to 7 days.
\end{verbatim}

  and days, seconds and microseconds are then normalized so that the
  representation is unique, with

\begin{itemize}
  \item \code{0 <= \var{microseconds} < 1000000}
  \item \code{0 <= \var{seconds} < 3600*24} (the number of seconds in one day)
  \item \code{-999999999 <= \var{days} <= 999999999}
\end{itemize}

  If any argument is a float, and there are fractional microseconds,
  the fractional microseconds left over from all arguments are combined
  and their sum is rounded to the nearest microsecond.  If no
  argument is a float, the conversion and normalization processes
  are exact (no information is lost).

  If the normalized value of days lies outside the indicated range,
  \exception{OverflowError} is raised.

  Note that normalization of negative values may be surprising at first.
  For example,

\begin{verbatim}
>>> d = timedelta(microseconds=-1)
>>> (d.days, d.seconds, d.microseconds)
(-1, 86399, 999999)
\end{verbatim}
\end{classdesc}

Class attributes are:

\begin{memberdesc}{min}
  The most negative \class{timedelta} object,
  \code{timedelta(-999999999)}.
\end{memberdesc}

\begin{memberdesc}{max}
  The most positive \class{timedelta} object,
  \code{timedelta(days=999999999, hours=23, minutes=59, seconds=59,
                  microseconds=999999)}.
\end{memberdesc}

\begin{memberdesc}{resolution}
  The smallest possible difference between non-equal
  \class{timedelta} objects, \code{timedelta(microseconds=1)}.
\end{memberdesc}

Note that, because of normalization, \code{timedelta.max} \textgreater
\code{-timedelta.min}.  \code{-timedelta.max} is not representable as
a \class{timedelta} object.

Instance attributes (read-only):

\begin{tableii}{c|l}{code}{Attribute}{Value}
  \lineii{days}{Between -999999999 and 999999999 inclusive}
  \lineii{seconds}{Between 0 and 86399 inclusive}
  \lineii{microseconds}{Between 0 and 999999 inclusive}
\end{tableii}

Supported operations:

\begin{tableiii}{c|l|c}{code}{Operation}{Result}{Notes}
  \lineiii{\var{t1} = \var{t2} + \var{t3}}
          {Sum of \var{t2} and \var{t3}.
           Afterwards \var{t1}-\var{t2} == \var{t3} and \var{t1}-\var{t3}
           == \var{t2} are true.}
          {(1)}
  \lineiii{\var{t1} = \var{t2} - \var{t3}}
          {Difference of \var{t2} and \var{t3}. Afterwards \var{t1} ==
           \var{t2} - \var{t3} and \var{t2} == \var{t1} + \var{t3} are
           true.}
          {(1)}
  \lineiii{\var{t1} = \var{t2} * \var{i} or \var{t1} = \var{i} * \var{t2}}
          {Delta multiplied by an integer or long.
           Afterwards \var{t1} // i == \var{t2} is true,
           provided \code{i != 0}.
           In general, \var{t1} * i == \var{t1} * (i-1) + \var{t1} is true.}
          {(1)}
  \lineiii{\var{t1} = \var{t2} // \var{i}}
          {The floor is computed and the remainder (if any) is thrown away.}
          {(3)}
  \lineiii{+\var{t1}}
          {Returns a \class{timedelta} object with the same value.}
          {(2)}
  \lineiii{-\var{t1}}
          {equivalent to \class{timedelta}(-\var{t1.days}, -\var{t1.seconds},
           -\var{t1.microseconds}),and to \var{t1}* -1.}
          {(1)(4)}
  \lineiii{abs(\var{t})}
          {equivalent to +\var{t} when \code{t.days >= 0}, and to
           -\var{t} when \code{t.days < 0}.
           overflow.}
          {(2)}
\end{tableiii}
\noindent
Notes:

\begin{description}
\item[(1)]
  This is exact, but may overflow.

\item[(2)]
  This is exact, and cannot overflow.

\item[(3)]
  Division by 0 raises \exception{ZeroDivisionError}.

\item[(4)]
  -\var{timedelta.max} is not representable as a \class{timedelta} object.
\end{description}

In addition to the operations listed above \class{timedelta} objects
support certain additions and subtractions with \class{date},
\class{datetime}, and \class{datimetz} objects (see below).

Comparisons of \class{timedelta} objects are supported with the
\class{timedelta} object representing the smaller duration considered
to be the smaller timedelta.

\class{timedelta} objects are hashable (usable as dictionary key),
support efficient pickling, and in Boolean contexts, a \class{timedelta}
object is considered to be true if and only if it isn't equal to
\code{timedelta(0)}.


\subsection{\class{date} Objects \label{datetime-date}}

A \class{date} object represents a date (year, month and day) in an idealized
calendar, the current Gregorian calendar indefinitely extended in both
directions.  January 1 of year 1 is called day number 1, January 2 of year
1 is called day number 2, and so on.  This matches the definition of the
"proleptic Gregorian" calendar in Dershowitz and Reingold's book
\citetitle{Calendrical Calculations}, where it's the base calendar for all
computations.  See the book for algorithms for converting between
proleptic Gregorian ordinals and many other calendar systems.

\begin{classdesc}{date}{year, month, day}
  All arguments are required.  Arguments may be ints or longs, in the
  following ranges:

  \begin{itemize}
    \item \code{MINYEAR <= \var{year} <= MAXYEAR}
    \item \code{1 <= \var{month} <= 12}
    \item \code{1 <= \var{day} <= number of days in the given month and year}
  \end{itemize}

  If an argument outside those ranges is given, \exception{ValueError}
  is raised.
\end{classdesc}

Other constructors, all class methods:

\begin{methoddesc}{today}{}
  Return the current local date.  This is equivalent to
  \code{date.fromtimestamp(time.time())}.
\end{methoddesc}

\begin{methoddesc}{fromtimestamp}{timestamp}
  Return the local date corresponding to the POSIX timestamp, such
  as is returned by \function{time.time()}.  This may raise
  \exception{ValueError}, if the timestamp is out of the range of
  values supported by the platform C \cfunction{localtime()}
  function.  It's common for this to be restricted to years from 1970
  through 2038.  Note that on non-POSIX systems that include leap
  seconds in their notion of a timestamp, leap seconds are ignored by
  \method{fromtimestamp()}.

\begin{methoddesc}{fromordinal}{ordinal}
  Return the date corresponding to the proleptic Gregorian ordinal,
  where January 1 of year 1 has ordinal 1.  \exception{ValueError} is
  raised unless \code{1 <= \var{ordinal} <= date.max.toordinal()}.
  For any date \var{d}, \code{date.fromordinal(\var{d}.toordinal()) ==
  \var{d}}.
\end{methoddesc}

Class attributes:

\begin{memberdesc}{min}
  The earliest representable date, \code{date(MINYEAR, 1, 1)}.
\end{memberdesc}

\begin{memberdesc}{max}
  The latest representable date, \code{date(MAXYEAR, 12, 31)}.
\end{memberdesc}

\begin{memberdesc}{resolution}
  The smallest possible difference between non-equal date
  objects, \code{timedelta(days=1)}.
\end{memberdesc}

Instance attributes (read-only):

\begin{memberdesc}{year}
  Between \constant{MINYEAR} and \constant{MAXYEAR} inclusive
\end{memberdesc}

\begin{memberdesc}{month}
  Between 1 and 12 inclusive.
\end{memberdesc}

\begin{memberdesc}{day}
  Between 1 and the number of days in the given month of the given
  year.
\end{memberdesc}

Supported operations:

\begin{itemize}
  \item
    date1 + timedelta -> date2
    timedelta + date1 -> date2
    date2 is timedelta.days days removed from the date1, moving forward
    in time if timedelta.days > 0, or backward if timedetla.days < 0.
    date2 - date1 == timedelta.days after.  timedelta.seconds and
    timedelta.microseconds are ignored.  \exception{OverflowError} is
    raised if date2.year would be smaller than \constant{MINYEAR} or
    larger than \constant{MAXYEAR}.

  \item
    date1 - timedelta -> date2
    Computes the date2 such that date2 + timedelta == date1.  This
    isn't quite equivalent to date1 + (-timedelta), because -timedelta
    in isolation can overflow in cases where date1 - timedelta does
    not.  timedelta.seconds and timedelta.microseconds are ignored.

  \item
    date1 - date2 -> timedelta
    This is exact, and cannot overflow.  timedelta.seconds and
    timedelta.microseconds are 0, and date2 + timedelta == date1
    after.

  \item
    comparison of date to date, where date1 is considered less than
    date2 when date1 precedes date2 in time.  In other words,
    date1 < date2 if and only if date1.toordinal() < date2.toordinal().

  \item
    hash, use as dict key

  \item
    efficient pickling

  \item
    in Boolean contexts, all \class{date} objects are considered to be true
\end{itemize}

Instance methods:

\begin{methoddesc}{replace}{year, month, day}
  Return a date with the same value, except for those fields given
  new values by whichever keyword arguments are specified.  For
  example, if \code{d == date(2002, 12, 31)}, then
  \code{d.replace(day=26) == date(2000, 12, 26)}.
\end{methoddesc}

\begin{methoddesc}{timetuple}{}
  Return a 9-element tuple of the form returned by
  \function{time.localtime()}.  The hours, minutes and seconds are
  0, and the DST flag is -1.
  \code{\var{d}.timetuple()} is equivalent to
      \code{(\var{d}.year, \var{d}.month, \var{d}.day,
             0, 0, 0,  \# h, m, s
             \var{d}.weekday(),  \# 0 is Monday
             \var{d}.toordinal() - date(\var{d}.year, 1, 1).toordinal() + 1,
             \# day of year
            -1)}
\end{methoddesc}

\begin{methoddesc}{toordinal}{}
  Return the proleptic Gregorian ordinal of the date, where January 1
  of year 1 has ordinal 1.  For any \class{date} object \var{d},
  \code{date.fromordinal(\var{d}.toordinal()) == \var{d}}.
\end{methoddesc}

\begin{methoddesc}{weekday}{}
  Return the day of the week as an integer, where Monday is 0 and
  Sunday is 6.  For example, date(2002, 12, 4).weekday() == 2, a
  Wednesday.
  See also \method{isoweekday()}.
\end{methoddesc}

\begin{methoddesc}{isoweekday}{}
  Return the day of the week as an integer, where Monday is 1 and
  Sunday is 7.  For example, date(2002, 12, 4).isoweekday() == 3, a
  Wednesday.
  See also \method{weekday()}, \method{isocalendar()}.
\end{methoddesc}

\begin{methoddesc}{isocalendar}{}
  Return a 3-tuple, (ISO year, ISO week number, ISO weekday).

  The ISO calendar is a widely used variant of the Gregorian calendar.
  See \url{http://www.phys.uu.nl/~vgent/calendar/isocalendar.htm}
  for a good explanation.

  The ISO year consists of 52 or 53 full weeks, and where a week starts
  on a Monday and ends on a Sunday.  The first week of an ISO year is
  the first (Gregorian) calendar week of a year containing a Thursday.
  This is called week number 1, and the ISO year of that Thursday is
  the same as its Gregorian year.

  For example, 2004 begins on a Thursday, so the first week of ISO
  year 2004 begins on Monday, 29 Dec 2003 and ends on Sunday, 4 Jan
  2004, so that

  date(2003, 12, 29).isocalendar() == (2004, 1, 1)
  date(2004, 1, 4).isocalendar() == (2004, 1, 7)
\end{methoddesc}

\begin{methoddesc}{isoformat}{}
  Return a string representing the date in ISO 8601 format,
  'YYYY-MM-DD'.  For example,
  date(2002, 12, 4).isoformat() == '2002-12-04'.
\end{methoddesc}

\begin{methoddesc}{__str__}{}
  For a date \var{d}, \code{str(\var{d})} is equivalent to
  \code{\var{d}.isoformat()}.
\end{methoddesc}

\begin{methoddesc}{ctime}{}
  Return a string representing the date, for example
  date(2002, 12, 4).ctime() == 'Wed Dec  4 00:00:00 2002'.
  \code{\var{d}.ctime()} is equivalent to
  \code{time.ctime(time.mktime(\var{d}.timetuple()))}
  on platforms where the native C \cfunction{ctime()} function
  (which \function{time.ctime()} invokes, but which
  \method{date.ctime()} does not invoke) conforms to the C standard.
\end{methoddesc}

\begin{methoddesc}{strftime}{format}
  Return a string representing the date, controlled by an explicit
  format string.  Format codes referring to hours, minutes or seconds
  will see 0 values.
  See the section on \method{strftime()} behavior.
\end{methoddesc}


\subsection{\class{datetime} Objects \label{datetime-datetime}}

A \class{datetime} object is a single object containing all the
information from a \class{date} object and a time object.  Like a
\class{date} object, \class{datetime} assumes the current Gregorian
calendar extended in both directions; like a time object,
\class{datetime} assumes there are exactly 3600*24 seconds in every
day.

\begin{classdesc}{datetime}{year, month, day,
                            hour=0, minute=0, second=0, microsecond=0}
  The year, month and day arguments are required.  Arguments may be
  ints or longs, in the following ranges:

  \begin{itemize}
    \item \code{\member{MINYEAR} <= \var{year} <= \member{MAXYEAR}}
    \item \code{1 <= \var{month} <= 12}
    \item \code{1 <= \var{day} <= number of days in the given month and year}
    \item \code{0 <= \var{hour} < 24}
    \item \code{0 <= \var{minute} < 60}
    \item \code{0 <= \var{second} < 60}
    \item \code{0 <= \var{microsecond} < 1000000}
  \end{itemize}

  If an argument outside those ranges is given, \exception{ValueError}
  is raised.
\end{classdesc}

Other constructors, all class methods:

\begin{methoddesc}{today}{}
  Return the current local datetime.  This is equivalent to
  \code{datetime.fromtimestamp(time.time())}.
  See also \method{now()}, \method{fromtimestamp()}.
\end{methoddesc}

\begin{methoddesc}{now}{}
  Return the current local datetime.  This is like \method{today()},
  but, if possible, supplies more precision than can be gotten from
  going through a \function{time.time()} timestamp (for example,
  this may be possible on platforms that supply the C
  \cfunction{gettimeofday()} function).
  See also \method{today()}, \method{utcnow()}.
\end{methoddesc}

\begin{methoddesc}{utcnow}{}
  Return the current UTC datetime.  This is like \method{now()}, but
  returns the current UTC date and time.
  See also \method{now()}.
\end{methoddesc}

\begin{methoddesc}{fromtimestamp}{timestamp}
  Return the local \class{datetime} corresponding to the \POSIX{}
  timestamp, such as is returned by \function{time.time()}.  This
  may raise \exception{ValueError}, if the timestamp is out of the
  range of values supported by the platform C
  \cfunction{localtime()} function.  It's common for this to be
  restricted to years in 1970 through 2038.
  Note that on non-POSIX systems that include leap seconds in their
  notion of a timestamp, leap seconds are ignored by
  \method{fromtimestamp()}, and then it's possible to have two timestamps
  differing by a second that yield identical \class{datetime} objects.
\end{methoddesc}
  See also \method{utcfromtimestamp()}.
\end{methoddesc}

\begin{methoddesc}{utcfromtimestamp}{timestamp}
  Return the UTC \class{datetime} corresponding to the \POSIX{}
  timestamp.  This may raise \exception{ValueError}, if the
  timestamp is out of the range of values supported by the platform
  C \cfunction{gmtime()} function.  It's common for this to be
  restricted to years in 1970 through 2038.
  See also \method{fromtimestamp()}.
\end{methoddesc}

\begin{methoddesc}{fromordinal}{ordinal}
  Return the \class{datetime} corresponding to the proleptic
  Gregorian ordinal, where January 1 of year 1 has ordinal 1.
  \exception{ValueError} is raised unless 1 <= ordinal <=
  datetime.max.toordinal().  The hour, minute, second and
  microsecond of the result are all 0.
\end{methoddesc}

\begin{methoddesc}{combine}{date, time}
  Return a new \class{datetime} object whose date components are
  equal to the given \class{date} object's, and whose time
  components are equal to the given time object's.  For any
  \class{datetime} object \var{d}, \code{\var{d} ==
  datetime.combine(\var{d}.date(), \var{d}.time())}.  If date is a
  \class{datetime} or \class{datetimetz} object, its time components
  are ignored.  If date is \class{datetimetz} object, its
  \member{tzinfo} component is also ignored.  If time is a
  \class{timetz} object, its \member{tzinfo} component is ignored.
\end{methoddesc}

Class attributes:

\begin{memberdesc}{min}
  The earliest representable \class{datetime},
  \code{datetime(MINYEAR, 1, 1)}.
\end{memberdesc}

\begin{memberdesc}{max}
  The latest representable \class{datetime},
  \code{datetime(MAXYEAR, 12, 31, 23, 59, 59, 999999)}.
\end{memberdesc}

\begin{memberdesc}{resolution}
  The smallest possible difference between non-equal \class{datetime}
  objects, \code{timedelta(microseconds=1)}.
\end{memberdesc}

Instance attributes (read-only):

\begin{memberdesc}{year}
  Between \constant{MINYEAR} and \constant{MAXYEAR} inclusive
\end{memberdesc}

\begin{memberdesc}{month}
  Between 1 and 12 inclusive
\end{memberdesc}

\begin{memberdesc}{day}
  Between 1 and the number of days in the given month of the given
  year.
\end{memberdesc}

\begin{memberdesc}{hour}
  In \code{range(24)}.
\end{memberdesc}

\begin{memberdesc}{minute}
  In \code{range(60)}.
\end{memberdesc}

\begin{memberdesc}{second}
  In \code{range(60)}.
\end{memberdesc}

\begin{memberdesc}{microsecond}
  In \code{range(1000000)}.
\end{memberdesc}

Supported operations:

\begin{itemize}
  \item
    datetime1 + timedelta -> datetime2
    timedelta + datetime1 -> datetime2
    datetime2 is a duration of timedelta removed from datetime1, moving
    forward in time if timedelta.days > 0, or backward if
    timedelta.days < 0.  datetime2 - datetime1 == timedelta after.
    \exception{OverflowError} is raised if datetime2.year would be
    smaller than \constant{MINYEAR} or larger than \constant{MAXYEAR}.

  \item
    datetime1 - timedelta -> datetime2
    Computes the datetime2 such that datetime2 + timedelta == datetime1.
    This isn't quite equivalent to datetime1 + (-timedelta), because
    -timedelta in isolation can overflow in cases where
    datetime1 - timedelta does not.

  \item
    datetime1 - datetime2 -> timedelta
    This is exact, and cannot overflow.
    datetime2 + timedelta == datetime1 after.

  \item
    comparison of \class{datetime} to datetime, where datetime1 is
    considered less than datetime2 when datetime1 precedes datetime2
    in time.

  \item
    hash, use as dict key

  \item
    efficient pickling

  \item
    in Boolean contexts, all \class{datetime} objects are considered
    to be true
\end{itemize}

Instance methods:

\begin{methoddesc}{date}{}
  Return \class{date} object with same year, month and day.
\end{methoddesc}

\begin{methoddesc}{time}{}
  Return time object with same hour, minute, second and microsecond.
\end{methoddesc}

\begin{methoddesc}{replace}{year=, month=, day=, hour=, minute=,
                            second=, microsecond=}
  Return a datetime with the same value, except for those fields given
  new values by whichever keyword arguments are specified.
\end{methoddesc}

\begin{methoddesc}{astimezone}{tz}
  Return a \class{datetimetz} with the same date and time fields, and
  with \member{tzinfo} member \var{tz}.  \var{tz} must be \code{None},
  or an instance of a \class{tzinfo} subclass.
\end{methoddesc}

\begin{methoddesc}{timetuple}{}
  Return a 9-element tuple of the form returned by
  \function{time.localtime()}.
  The DST flag is -1.   \code{\var{d}.timetuple()} is equivalent to
  \code{(\var{d}.year, \var{d}.month, \var{d}.day,
         \var{d}.hour, \var{d}.minute, \var{d}.second,
         \var{d}.weekday(),  \# 0 is Monday
         \var{d}.toordinal() - date(\var{d}.year, 1, 1).toordinal() + 1,
         \# day of year
         -1)}
\end{methoddesc}

\begin{methoddesc}{toordinal}{}
  Return the proleptic Gregorian ordinal of the date.  The same as
  \method{date.toordinal()}.
\end{methoddesc}

\begin{methoddesc}{weekday}{}
  Return the day of the week as an integer, where Monday is 0 and
  Sunday is 6.  The same as \method{date.weekday()}.
  See also \method{isoweekday()}.
\end{methoddesc}

\begin{methoddesc}{isoweekday}{}
  Return the day of the week as an integer, where Monday is 1 and
  Sunday is 7.  The same as \method{date.isoweekday()}.
  See also \method{weekday()}, \method{isocalendar()}.
\end{methoddesc}

\begin{methoddesc}{isocalendar}{}
  Return a 3-tuple, (ISO year, ISO week number, ISO weekday).  The
  same as \method{date.isocalendar()}.
\end{methoddesc}

\begin{methoddesc}{isoformat}{sep='T'}
  Return a string representing the date and time in ISO 8601 format,
      YYYY-MM-DDTHH:MM:SS.mmmmmm
  or, if self.microsecond is 0,
      YYYY-MM-DDTHH:MM:SS
  The optional argument \var{sep} (default \code{'T'}) is a
  one-character separator, placed between the date and time portions
  of the result.  For example,
      datetime(2002, 12, 4, 1, 2, 3, 4).isoformat(' ') ==
      '2002-12-04 01:02:03.000004'
\end{methoddesc}

\begin{methoddesc}{__str__}{}
  For a \class{datetime} instance \var{d}, \code{str(\var{d})} is
  equivalent to \code{\var{d}.isoformat(' ')}.
\end{methoddesc}

\begin{methoddesc}{ctime}{}
  Return a string representing the date, for example
  datetime(2002, 12, 4, 20, 30, 40).ctime() == 'Wed Dec  4 20:30:40 2002'.
  \code{d.ctime()} is equivalent to
  \code{time.ctime(time.mktime(d.timetuple()))} on platforms where
  the native C \cfunction{ctime()} function (which
  \function{time.ctime()} invokes, but which
  \method{datetime.ctime()} does not invoke) conforms to the C
  standard.
\end{methoddesc}

\begin{methoddesc}{strftime}{format}
  Return a string representing the date and time, controlled by an
  explicit format string.  See the section on \method{strftime()}
  behavior.
\end{methoddesc}


\subsection{\class{time} Objects \label{datetime-time}}

A \class{time} object represents an idealized time of day, independent
of day and timezone.

\begin{classdesc}{time}{hour=0, minute=0, second=0, microsecond=0}
  All arguments are optional.  They may be ints or longs, in the
  following ranges:

\begin{itemize}
  \item \code{0 <= \var{hour} < 24}
  \item \code{0 <= \var{minute} < 60}
  \item \code{0 <= \var{second} < 60}
  \item \code{0 <= \var{microsecond} < 1000000}
\end{itemize}

  If an argument outside those ranges is given, \exception{ValueError}
  is raised.
\end{classdesc}

Class attributes:

\begin{memberdesc}{min}
  The earliest representable \class{time}, \code{time(0, 0, 0, 0)}.
\end{memberdesc}

\begin{memberdesc}{max}
  The latest representable \class{time}, \code{time(23, 59, 59, 999999)}.
\end{memberdesc}

\begin{memberdesc}{resolution}
  The smallest possible difference between non-equal \class{time}
  objects, \code{timedelta(microseconds=1)}, although note that
  arithmetic on \class{time} objects is not supported.
\end{memberdesc}

Instance attributes (read-only):

\begin{memberdesc}{hour}
  In \code{range(24)}.
\end{memberdesc}

\begin{memberdesc}{minute}
  In \code{range(60)}.
\end{memberdesc}

\begin{memberdesc}{second}
  In \code{range(60)}.
\end{memberdesc}

\begin{memberdesc}{microsecond}
  In \code{range(1000000)}.
\end{memberdesc}

Supported operations:

\begin{itemize}
  \item
    comparison of time to time, where time1 is considered
    less than time2 when time1 precedes time2 in time.

  \item
    hash, use as dict key

  \item
    efficient pickling

  \item
    in Boolean contexts, a time object is considered to be true
    if and only if it isn't equal to \code{time(0)}
\end{itemize}

Instance methods:

\begin{methoddesc}{replace}{hour=, minute=, second=, microsecond=}
  Return a time with the same value, except for those fields given
  new values by whichever keyword arguments are specified.
\end{methoddesc}

\begin{methoddesc}{isoformat}{}
  Return a string representing the time in ISO 8601 format,
      HH:MM:SS.mmmmmm
  or, if self.microsecond is 0
      HH:MM:SS
\end{methoddesc}

\begin{methoddesc}{__str__}{}
  For a time \var{t}, \code{str(\var{t})} is equivalent to
  \code{\var{t}.isoformat()}.
\end{methoddesc}

\begin{methoddesc}{strftime}{format}
  Return a string representing the time, controlled by an explicit
  format string.  See the section on \method{strftime()} behavior.
\end{methoddesc}


\subsection{\class{tzinfo} Objects \label{datetime-tzinfo}}

\class{tzinfo} is an abstract base clase, meaning that this class
should not be instantiated directly.  You need to derive a concrete
subclass, and (at least) supply implementations of the standard
\class{tzinfo} methods needed by the \class{datetime} methods you
use.  The \module{datetime} module does not supply any concrete
subclasses of \class{tzinfo}.

An instance of (a concrete subclass of) \class{tzinfo} can be passed
to the constructors for \class{datetimetz} and \class{timetz} objects.
The latter objects view their fields as being in local time, and the
\class{tzinfo} object supports methods revealing offset of local time
from UTC, the name of the time zone, and DST offset, all relative to a
date or time object passed to them.

Special requirement for pickling:  A \class{tzinfo} subclass must have an
\method{__init__} method that can be called with no arguments, else it
can be pickled but possibly not unpickled again.  This is a technical
requirement that may be relaxed in the future.

A concrete subclass of \class{tzinfo} may need to implement the
following methods.  Exactly which methods are needed depends on the
uses made of aware \module{datetime} objects.  If in doubt, simply
implement all of them.

\begin{methoddesc}{utcoffset}{self, dt}
  Return offset of local time from UTC, in minutes east of UTC.  If
  local time is west of UTC, this should be negative.  Note that this
  is intended to be the total offset from UTC; for example, if a
  \class{tzinfo} object represents both time zone and DST adjustments,
  \method{utcoffset()} should return their sum.  If the UTC offset
  isn't known, return \code{None}.  Else the value returned must be
  a \class{timedelta} object specifying a whole number of minutes in the
  range -1439 to 1439 inclusive (1440 = 24*60; the magnitude of the offset
  must be less than one day).  Most implementations of
  \method{utcoffset()} will probably look like one of these two:

\begin{verbatim}
    return CONSTANT                 # fixed-offset class
    return CONSTANT + self.dst(dt)  # daylight-aware class
\end{verbatim}

    If \method{utcoffset()} does not return \code{None},
    \method{dst()} should not return \code{None} either.
\end{methoddesc}


\begin{methoddesc}{dst}{self, dt}
  Return the daylight savings time (DST) adjustment, in minutes east of
  UTC, or \code{None} if DST information isn't known.  Return \code{0} if
  DST is not in effect.
  If DST is in effect, return the offset as a
  \class{timedelta} object (see \method{utcoffset()} for details).
  Note that DST offset, if applicable, has
  already been added to the UTC offset returned by
  \method{utcoffset()}, so there's no need to consult \method{dst()}
  unless you're interested in displaying DST info separately.  For
  example, \method{datetimetz.timetuple()} calls its \member{tzinfo}
  member's \method{dst()} method to determine how the
  \member{tm_isdst} flag should be set, and
  \method{datetimetz.astimezone()} calls \method{dst()} to account for
  DST changes when crossing time zones.

  An instance \var{tz} of a \class{tzinfo} subclass that models both
  standard and daylight times must be consistent in this sense:

      \code{tz.utcoffset(dt) - tz.dst(dt)}

  must return the same result for every \class{datetimetz} \var{dt}
  with \code{dt.tzinfo==tz}  For sane \class{tzinfo} subclasses, this
  expression yields the time zone's "standard offset", which should not
  depend on the date or the time, but only on geographic location.  The
  implementation of \method{datetimetz.astimezone()} relies on this, but
  cannot detect violations; it's the programmer's responsibility to
  ensure it.

\begin{methoddesc}{tzname}{self, dt}
  Return the timezone name corresponding to the \class{datetime} represented
  by \var{dt}, as a string.  Nothing about string names is defined by the
  \module{datetime} module, and there's no requirement that it mean anything
  in particular.  For example, "GMT", "UTC", "-500", "-5:00", "EDT",
  "US/Eastern", "America/New York" are all valid replies.  Return
  \code{None} if a string name isn't known.  Note that this is a method
  rather than a fixed string primarily because some \class{tzinfo} objects
  will wish to return different names depending on the specific value
  of \var{dt} passed, especially if the \class{tzinfo} class is
  accounting for daylight time.
\end{methoddesc}

\end{methoddesc}

These methods are called by a \class{datetimetz} or \class{timetz} object,
in response to their methods of the same names.  A \class{datetimetz}
object passes itself as the argument, and a \class{timetz} object passes
\code{None} as the argument.  A \class{tzinfo} subclass's methods should
therefore be prepared to accept a \var{dt} argument of \code{None}, or of
class \class{datetimetz}.

When \code{None} is passed, it's up to the class designer to decide the
best response.  For example, returning \code{None} is appropriate if the
class wishes to say that timetz objects don't participate in the
\class{tzinfo} protocol.  In other applications, it may be more useful
for \code{utcoffset(None)} to return the standard UTC offset.

When a \class{datetimetz} object is passed in response to a
\class{datetimetz} method, \code{dt.tzinfo} is the same object as
\var{self}.  \class{tzinfo} methods can rely on this, unless
user code calls \class{tzinfo} methods directly.  The intent is that
the \class{tzinfo} methods interpret \var{dt} as being in local time,
and not need to worry about objects in other timezones.

Example \class{tzinfo} classes:

\verbatiminput{tzinfo-examples.py}

Note that there are unavoidable subtleties twice per year in a tzinfo
subclass accounting for both standard and daylight time, at the DST
transition points.  For concreteness, consider US Eastern (UTC -0500),
where EDT begins the minute after 1:59 (EST) on the first Sunday in
April, and ends the minute after 1:59 (EDT) on the last Sunday in October:

\begin{verbatim}
    UTC   3:MM  4:MM  5:MM  6:MM  7:MM  8:MM
    EST  22:MM 23:MM  0:MM  1:MM  2:MM  3:MM
    EDT  23:MM  0:MM  1:MM  2:MM  3:MM  4:MM

  start  22:MM 23:MM  0:MM  1:MM  3:MM  4:MM

    end  23:MM  0:MM  1:MM  1:MM  2:MM  3:MM
\end{verbatim}

When DST starts (the "start" line), the local wall clock leaps from 1:59
to 3:00.  A wall time of the form 2:MM doesn't really make sense on that
day, so \code{astimezone(Eastern)} won't deliver a result with
\code{hour==2} on the
day DST begins.  How an Eastern instance chooses to interpret 2:MM on
that day is its business.  The example Eastern implementation above
chose to
consider it as a time in EDT, simply because it "looks like it's
after 2:00", and so synonymous with the EST 1:MM times on that day.
Your Eastern class may wish, for example, to raise an exception instead
when it sees a 2:MM time on the day EDT begins.

When DST ends (the "end" line), there's a potentially worse problem:
there's an hour that can't be spelled unambiguously in local wall time, the
hour beginning at the moment DST ends.  In this example, that's times of
the form 6:MM UTC on the day daylight time ends.  The local wall clock
leaps from 1:59 (daylight time) back to 1:00 (standard time) again.
1:MM is taken as daylight time (it's "before 2:00"), so maps to 5:MM UTC.
2:MM is taken as standard time (it's "after 2:00"), so maps to 7:MM UTC.
There is no local time that maps to 6:MM UTC on this day.

Just as the wall clock does, \code{astimezone(Eastern)} maps both UTC
hours 5:MM
and 6:MM to Eastern hour 1:MM on this day.  However, this result is
ambiguous (there's no way for Eastern to know which repetition of 1:MM
is intended).  Applications that can't bear such ambiguity
should avoid using hybrid tzinfo classes; there are no
ambiguities when using UTC, or any other fixed-offset tzinfo subclass
(such as a class representing only EST (fixed offset -5 hours), or only
EDT (fixed offset -4 hours)).


\subsection{\class{timetz} Objects \label{datetime-timetz}}

A time object represents a (local) time of day, independent of any
particular day, and subject to adjustment via a \class{tzinfo} object.

Constructor:

\begin{classdesc}{time}{hour=0, minute=0, second=0, microsecond=0,
                        tzinfo=None}
  All arguments are optional.  \var{tzinfo} may be \code{None}, or
  an instance of a \class{tzinfo} subclass.  The remaining arguments
  may be ints or longs, in the following ranges:

  \begin{itemize}
    \item \code{0 <= \var{hour} < 24}
    \item \code{0 <= \var{minute} < 60}
    \item \code{0 <= \var{second} < 60}
    \item \code{0 <= \var{microsecond} < 1000000}.
  \end{itemize}

  If an argument outside those ranges is given,
  \exception{ValueError} is raised.
\end{classdesc}

Class attributes:

\begin{memberdesc}{min}
  The earliest representable time, \code{timetz(0, 0, 0, 0)}.
\end{memberdesc}

\begin{memberdesc}{max}
  The latest representable time, \code{timetz(23, 59, 59, 999999)}.
\end{memberdesc}

\begin{memberdesc}{resolution}
  The smallest possible difference between non-equal \class{timetz}
  objects, \code{timedelta(microseconds=1)}, although note that
  arithmetic on \class{timetz} objects is not supported.
\end{memberdesc}

Instance attributes (read-only):

\begin{memberdesc}{hour}
  In \code{range(24)}.
\end{memberdesc}

\begin{memberdesc}{minute}
  In \code{range(60)}.
\end{memberdesc}

\begin{memberdesc}{second}
  In \code{range(60)}.
\end{memberdesc}

\begin{memberdesc}{microsecond}
  In \code{range(1000000)}.
\end{memberdesc}

\begin{memberdesc}{tzinfo}
  The object passed as the tzinfo argument to the \class{timetz}
  constructor, or \code{None} if none was passed.
\end{memberdesc}

Supported operations:

\begin{itemize}
  \item
    comparison of \class{timetz} to \class{time} or \class{timetz},
    where \var{a} is considered less than \var{b} when \var{a} precedes
    \var{b} in time.  If one comparand is naive and the other is aware,
    \exception{TypeError} is raised.  If both comparands are aware, and
    have the same \member{tzinfo} member, the common \member{tzinfo}
    member is ignored and the base times are compared.  If both
    comparands are aware and have different \member{tzinfo} members,
    the comparands are first adjusted by subtracting their UTC offsets
    (obtained from \code{self.utcoffset()}).

  \item
    hash, use as dict key

  \item
    pickling

  \item
    in Boolean contexts, a \class{timetz} object is considered to be
    true if and only if, after converting it to minutes and
    subtracting \method{utcoffset()} (or \code{0} if that's
    \code{None}), the result is non-zero.
\end{itemize}

Instance methods:

\begin{methoddesc}{replace}(hour=, minute=, second=, microsecond=, tzinfo=)
  Return a \class{timetz} with the same value, except for those fields given
  new values by whichever keyword arguments are specified.  Note that
  \code{tzinfo=None} can be specified to create a naive \class{timetz} from an
  aware \class{timetz}.
\end{methoddesc}

\begin{methoddesc}{isoformat}{}
  Return a string representing the time in ISO 8601 format,
      HH:MM:SS.mmmmmm
  or, if self.microsecond is 0,
      HH:MM:SS
  If \method{utcoffset()} does not return \code{None}, a 6-character
  string is appended, giving the UTC offset in (signed) hours and
  minutes:
      HH:MM:SS.mmmmmm+HH:MM
  or, if self.microsecond is 0,
      HH:MM:SS+HH:MM
\end{methoddesc}

\begin{methoddesc}{__str__}{}
  For a \class{timetz} \var{t}, \code{str(\var{t})} is equivalent to
  \code{\var{t}.isoformat()}.
\end{methoddesc}

\begin{methoddesc}{strftime}{format}
  Return a string representing the time, controlled by an explicit
  format string.  See the section on \method{strftime()} behavior.
\end{methoddesc}

\begin{methoddesc}{utcoffset}{}
  If \member{tzinfo} is \code{None}, returns \code{None}, else
  \code{tzinfo.utcoffset(self)} converted to a \class{timedelta}
  object.
\end{methoddesc}

\begin{methoddesc}{tzname}{}
  If \member{tzinfo} is \code{None}, returns \code{None}, else
  \code{tzinfo.tzname(self)}.
\end{methoddesc}

\begin{methoddesc}{dst}{}
  If \member{tzinfo} is \code{None}, returns \code{None}, else
  \code{tzinfo.dst(self)} converted to a \class{timedelta} object.
\end{methoddesc}



\subsection{ \class{datetimetz} Objects \label{datetime-datetimetz}}

\begin{notice}[warning]
  I think this is \emph{still} missing some methods from the
  Python implementation.
\end{notice}

A \class{datetimetz} object is a single object containing all the information
from a \class{date} object and a \class{timetz} object.

Constructor:

\begin{classdesc}{datetimetz}{year, month, day,
                              hour=0, minute=0, second=0, microsecond=0,
                              tzinfo=None}
  The year, month and day arguments are required.  \var{tzinfo} may
  be \code{None}, or an instance of a \class{tzinfo} subclass.  The
  remaining arguments may be ints or longs, in the following ranges:

  \begin{itemize}
    \item \code{MINYEAR <= \var{year} <= MAXYEAR}
    \item \code{1 <= \var{month} <= 12}
    \item \code{1 <= \var{day} <= number of days in the given month and year}
    \item \code{0 <= \var{hour} < 24}
    \item \code{0 <= \var{minute} < 60}
    \item \code{0 <= \var{second} < 60}
    \item \code{0 <= \var{microsecond} < 1000000}
  \end{itemize}

  If an argument outside those ranges is given,
  \exception{ValueError} is raised.
\end{classdesc}

Other constructors (class methods):

\begin{funcdesc}{today}{}
\methodline{utcnow}{}
\methodline{utcfromtimestamp}{timestamp}
\methodline{fromordinal}{ordinal}
  These are the same as the \class{datetime} class methods of the
  same names, except that they construct a \class{datetimetz}
  object, with tzinfo \code{None}.
\end{funcdesc}

\begin{funcdesc}{now}{\optional{tzinfo=None}}
\methodline{fromtimestamp}{timestamp\optional{, tzinfo=None}}
  These are the same as the \class{datetime} class methods of the
  same names, except that they accept an additional, optional tzinfo
  argument, and construct a \class{datetimetz} object with that
  \class{tzinfo} object attached.
\end{funcdesc}

\begin{funcdesc}{combine}{date, time}
  This is the same as \method{datetime.combine()}, except that it
  constructs a \class{datetimetz} object, and, if the time object is
  of type timetz, the \class{datetimetz} object has the same
  \class{tzinfo} object as the time object.
\end{funcdesc}

Class attributes:

\begin{memberdesc}{min}
  The earliest representable \class{datetimetz},
  \code{datetimetz(MINYEAR, 1, 1)}.
\end{memberdesc}

\begin{memberdesc}{max}
  The latest representable \class{datetime},
  \code{datetimetz(MAXYEAR, 12, 31, 23, 59, 59, 999999)}.
\end{memberdesc}

\begin{memberdesc}{resolution}
  The smallest possible difference between non-equal \class{datetimetz}
  objects, \code{timedelta(microseconds=1)}.
\end{memberdesc}

Instance attributes, all read-only:

\begin{memberdesc}{year}
  Between \constant{MINYEAR} and \constant{MAXYEAR}, inclusive.
\end{memberdesc}

\begin{memberdesc}{month}
  Between 1 and 12 inclusive
\end{memberdesc}

\begin{memberdesc}{day}
  Between 1 and the number of days in the given month of the given
  year.
\end{memberdesc}

\begin{memberdesc}{hour}
  In \code{range(24)}.
\end{memberdesc}

\begin{memberdesc}{minute}
  In \code{range(60)}.
\end{memberdesc}

\begin{memberdesc}{second}
  In \code{range(60)}.
\end{memberdesc}

\begin{memberdesc}{microsecond}
  In \code{range(1000000)}.
\end{memberdesc}

\begin{memberdesc}{tzinfo}
  The object passed as the \var{tzinfo} argument to the
  \class{datetimetz} constructor, or \code{None} if none was passed.
\end{memberdesc}

Supported operations:

\begin{itemize}
  \item
    datetimetz1 + timedelta -> datetimetz2
    timedelta + datetimetz1 -> datetimetz2

    The same as addition of \class{datetime} objects, except that
    datetimetz2.tzinfo is set to datetimetz1.tzinfo.

  \item
    datetimetz1 - timedelta -> datetimetz2

    The same as addition of \class{datetime} objects, except that
    datetimetz2.tzinfo is set to datetimetz1.tzinfo.

  \item
    aware_datetimetz1 - aware_datetimetz2 -> timedelta
    naive_datetimetz1 - naive_datetimetz2 -> timedelta
    naive_datetimetz1 - datetime2 -> timedelta
    datetime1 - naive_datetimetz2 -> timedelta

    Subtraction of a \class{datetime} or \class{datetimetz}, from a
    \class{datetime} or \class{datetimetz}, is defined only if both
    operands are naive, or if both are aware.  If one is aware and the
    other is naive, \exception{TypeError} is raised.

    If both are naive, or both are aware and have the same \member{tzinfo}
    member, subtraction acts as for \class{datetime} subtraction.

    If both are aware and have different \member{tzinfo} members,
    \code{a-b} acts as if \var{a} and \var{b} were first converted to UTC
    datetimes (by subtracting \code{a.utcoffset()} minutes from \var{a},
    and \code{b.utcoffset()} minutes from \var{b}), and then doing
    \class{datetime} subtraction, except that the implementation never
    overflows.

  \item
    comparison of \class{datetimetz} to \class{datetime} or
    \class{datetimetz}, where \var{a} is considered less than \var{b}
    when \var{a} precedes \var{b} in time.  If one comparand is naive and
    the other is aware, \exception{TypeError} is raised.  If both
    comparands are aware, and have the same \member{tzinfo} member,
    the common \member{tzinfo} member is ignored and the base datetimes
    are compared.  If both comparands are aware and have different
    \member{tzinfo} members, the comparands are first adjusted by
    subtracting their UTC offsets (obtained from \code{self.utcoffset()}).

  \item
    hash, use as dict key

  \item
    efficient pickling

  \item
    in Boolean contexts, all \class{datetimetz} objects are considered to be
    true
\end{itemize}

Instance methods:

\begin{methoddesc}{date}{}
\methodline{time}{}
\methodline{toordinal}{}
\methodline{weekday}{}
\methodline{isoweekday}{}
\methodline{isocalendar}{}
\methodline{ctime}{}
\methodline{__str__}{}
\methodline{strftime}{format}
  These are the same as the \class{datetime} methods of the same names.
\end{methoddesc}

\begin{methoddesc}{timetz}{}
  Return \class{timetz} object with same hour, minute, second, microsecond,
  and tzinfo.
\end{methoddesc}

\begin{methoddesc}{replace}{year=, month=, day=, hour=, minute=, second=,
                            microsecond=, tzinfo=}
  Return a datetimetz with the same value, except for those fields given
  new values by whichever keyword arguments are specified.  Note that
  \code{tzinfo=None} can be specified to create a naive datetimetz from
  an aware datetimetz.
\end{methoddesc}

\begin{methoddesc}{astimezone}{tz}
  Return a \class{datetimetz} object with new \membar{tzinfo} member
  \var{tz}.
  \var{tz} must be \code{None}, or an instance of a \class{tzinfo} subclass.
  If \var{tz} is \code{None}, \var{self} is naive,
  \code{tz.utcoffset(self)} returns \code{None},
  or \code{self.tzinfo}\ is \var{tz},
  \code{self.astimezone(tz)} is equivalent to
  \code{self.replace(tzinfo=tz)}:  a new timezone object is attached
  without any conversion of date or time fields.  Else \code{self.tzinfo}
  and \var{tz} must implement the \method{utcoffset()} and \method{dst()}
  \class{tzinfo} methods, and the date and time fields are adjusted so
  that the result is local time in time zone \var{tz}, representing the
  same UTC time as \var{self}:  after \code{astz = dt.astimezone(tz)},
  \code{astz - astz.utcoffset()} will usually have the same date and time
  members as \code{dt - dt.utcoffset()}.  The discussion of class
  \class{tzinfo} explains the cases at Daylight Saving Time
  transition boundaries where this cannot be achieved (an issue only if
  \var{tz} models both standard and daylight time).
\end{methoddesc}

\begin{methoddesc}{utcoffset}{}
  If \member{tzinfo} is \code{None}, returns \code{None}, else
  \code{tzinfo.utcoffset(self)} converted to a \class{timedelta}
  object.
\end{methoddesc}

\begin{methoddesc}{tzname}{}
  If \member{tzinfo} is \code{None}, returns \code{None}, else
  returns \code{tzinfo.tzname(self)}.
\end{methoddesc}

\begin{methoddesc}{dst}{}
  If \member{tzinfo} is \code{None}, returns \code{None}, else
  \code{tzinfo.dst(self)} converted to a \class{timedelta}
  object.
\end{methoddesc}

\begin{methoddesc}{timetuple}{}
  Like \function{datetime.timetuple()}, but sets the
  \member{tm_isdst} flag according to the \method{dst()} method:  if
  \method{dst()} returns \code{None}, \member{tm_isdst} is set to
  \code{-1}; else if \method{dst()} returns a non-zero value,
  \member{tm_isdst} is set to \code{1}; else \code{tm_isdst} is set
  to \code{0}.
\end{methoddesc}

\begin{methoddesc}{utctimetuple}{}
  If \class{datetimetz} instance \var{d} is naive, this is the same as
  \code{\var{d}.timetuple()} except that \member{tm_isdst} is forced to 0
  regardless of what \code{d.dst()} returns.  DST is never in effect
  for a UTC time.

  If \var{d} is aware, \var{d} is normalized to UTC time, by subtracting
  \code{\var{d}.utcoffset()} minutes, and a timetuple for the
  normalized time is returned.  \member{tm_isdst} is forced to 0.
  Note that the result's \member{tm_year} field may be
  \constant{MINYEAR}-1 or \constant{MAXYEAR}+1, if \var{d}.year was
  \code{MINYEAR} or \code{MAXYEAR} and UTC adjustment spills over a
  year boundary.
\end{methoddesc}

\begin{methoddesc}{isoformat}{sep='T'}
  Return a string representing the date and time in ISO 8601 format,
      YYYY-MM-DDTHH:MM:SS.mmmmmm
  or, if \member{microsecond} is 0,
      YYYY-MM-DDTHH:MM:SS

  If \method{utcoffset()} does not return \code{None}, a 6-character
  string is appended, giving the UTC offset in (signed) hours and
  minutes:
      YYYY-MM-DDTHH:MM:SS.mmmmmm+HH:MM
  or, if \member{microsecond} is 0
      YYYY-MM-DDTHH:MM:SS+HH:MM

  The optional argument \var{sep} (default \code{'T'}) is a
  one-character separator, placed between the date and time portions
  of the result.  For example,

\begin{verbatim}
>>> from datetime import *
>>> class TZ(tzinfo):
...     def utcoffset(self, dt): return timedelta(minutes=-399)
...
>>> datetimetz(2002, 12, 25, tzinfo=TZ()).isoformat(' ')
'2002-12-25 00:00:00-06:39'
\end{verbatim}
\end{methoddesc}

\code{str(\var{d})} is equivalent to \code{\var{d}.isoformat(' ')}.


\subsection{\method{strftime()} Behavior}

\class{date}, \class{datetime}, \class{datetimetz}, \class{time},
and \class{timetz} objects all support a \code{strftime(\var{format})}
method, to create a string representing the time under the control of
an explicit format string.  Broadly speaking,
\code{d.strftime(fmt)}
acts like the \refmodule{time} module's
\code{time.strftime(fmt, d.timetuple())}
although not all objects support a \method{timetuple()} method.

For \class{time} and \class{timetz} objects, the format codes for
year, month, and day should not be used, as time objects have no such
values.  If they're used anyway, \code{1900} is substituted for the
year, and \code{0} for the month and day.

For \class{date} objects, the format codes for hours, minutes, and
seconds should not be used, as \class{date} objects have no such
values.  If they're used anyway, \code{0} is substituted for them.

For a naive object, the \code{\%z} and \code{\%Z} format codes are
replaced by empty strings.

For an aware object:

\begin{itemize}
  \item[\code{\%z}]
    \method{utcoffset()} is transformed into a 5-character string of
    the form +HHMM or -HHMM, where HH is a 2-digit string giving the
    number of UTC offset hours, and MM is a 2-digit string giving the
    number of UTC offset minutes.  For example, if
    \method{utcoffset()} returns \code{timedelta(hours=-3, minutes=-30)},
    \code{\%z} is replaced with the string \code{'-0330'}.

  \item[\code{\%Z}]
    If \method{tzname()} returns \code{None}, \code{\%Z} is replaced
    by an empty string.  Otherwise \code{\%Z} is replaced by the returned
    value, which must be a string.
\end{itemize}

The full set of format codes supported varies across platforms,
because Python calls the platform C library's \function{strftime()}
function, and platform variations are common.  The documentation for
Python's \refmodule{time} module lists the format codes that the C
standard (1989 version) requires, and those work on all platforms
with a standard C implementation.  Note that the 1999 version of the
C standard added additional format codes.

The exact range of years for which \method{strftime()} works also
varies across platforms.  Regardless of platform, years before 1900
cannot be used.


\begin{comment}

\subsection{C API}

Struct typedefs:

    PyDateTime_Date
    PyDateTime_DateTime
    PyDateTime_DateTimeTZ
    PyDateTime_Time
    PyDateTime_TimeTZ
    PyDateTime_Delta
    PyDateTime_TZInfo

Type-check macros:

    PyDate_Check(op)
    PyDate_CheckExact(op)

    PyDateTime_Check(op)
    PyDateTime_CheckExact(op)

    PyDateTimeTZ_Check(op)
    PyDateTimeTZ_CheckExact(op)

    PyTime_Check(op)
    PyTime_CheckExact(op)

    PyTimeTZ_Check(op)
    PyTimeTZ_CheckExact(op)

    PyDelta_Check(op)
    PyDelta_CheckExact(op)

    PyTZInfo_Check(op)
    PyTZInfo_CheckExact(op)

Accessor macros:

All objects are immutable, so accessors are read-only.  All macros
return ints:

    For \class{date}, \class{datetime}, and \class{datetimetz} instances:
        PyDateTime_GET_YEAR(o)
        PyDateTime_GET_MONTH(o)
        PyDateTime_GET_DAY(o)

    For \class{datetime} and \class{datetimetz} instances:
        PyDateTime_DATE_GET_HOUR(o)
        PyDateTime_DATE_GET_MINUTE(o)
        PyDateTime_DATE_GET_SECOND(o)
        PyDateTime_DATE_GET_MICROSECOND(o)

    For \class{time} and \class{timetz} instances:
        PyDateTime_TIME_GET_HOUR(o)
        PyDateTime_TIME_GET_MINUTE(o)
        PyDateTime_TIME_GET_SECOND(o)
        PyDateTime_TIME_GET_MICROSECOND(o)

\end{comment}
