\section{\module{xml.etree.ElementTree} --- The ElementTree XML API}
\declaremodule{standard}{xml.etree.ElementTree}
\moduleauthor{Fredrik Lundh}{fredrik@pythonware.com}
\modulesynopsis{Implementation of the ElementTree API.}

\versionadded{2.5}

The Element type is a flexible container object, designed to store
hierarchical data structures in memory. The type can be described as a
cross between a list and a dictionary.

Each element has a number of properties associated with it:

\begin{itemize}
  \item a tag which is a string identifying what kind of data
        this element represents (the element type, in other words).
  \item a number of attributes, stored in a Python dictionary.
  \item a text string.
  \item an optional tail string.
  \item a number of child elements, stored in a Python sequence
\end{itemize}

To create an element instance, use the Element or SubElement factory
functions.

The \class{ElementTree} class can be used to wrap an element
structure, and convert it from and to XML.

A C implementation of this API is available as
\module{xml.etree.cElementTree}.


\subsection{Functions\label{elementtree-functions}}

\begin{funcdesc}{Comment}{\optional{text}}
Comment element factory.  This factory function creates a special
element that will be serialized as an XML comment.
The comment string can be either an 8-bit ASCII string or a Unicode
string.
\var{text} is a string containing the comment string.
Returns an element instance representing a comment.
\end{funcdesc}

\begin{funcdesc}{dump}{elem}
Writes an element tree or element structure to sys.stdout.  This
function should be used for debugging only.

The exact output format is implementation dependent.  In this
version, it's written as an ordinary XML file.

\var{elem} is an element tree or an individual element.
\end{funcdesc}

\begin{funcdesc}{Element}{tag\optional{, attrib}\optional{, **extra}}
Element factory.  This function returns an object implementing the
standard Element interface.  The exact class or type of that object
is implementation dependent, but it will always be compatible with
the {\_}ElementInterface class in this module.

The element name, attribute names, and attribute values can be
either 8-bit ASCII strings or Unicode strings.
\var{tag} is the element name.
\var{attrib} is an optional dictionary, containing element attributes.
\var{extra} contains additional attributes, given as keyword arguments.
Returns an element instance.
\end{funcdesc}

\begin{funcdesc}{fromstring}{text}
Parses an XML section from a string constant.  Same as XML.
\var{text} is a string containing XML data.
Returns an Element instance.
\end{funcdesc}

\begin{funcdesc}{iselement}{element}
Checks if an object appears to be a valid element object.
\var{element} is an element instance.
Returns a true value if this is an element object.
\end{funcdesc}

\begin{funcdesc}{iterparse}{source\optional{, events}}
Parses an XML section into an element tree incrementally, and reports
what's going on to the user.
\var{source} is a filename or file object containing XML data.
\var{events} is a list of events to report back.  If omitted, only ``end''
events are reported.
Returns an iterator providing \code{(\var{event}, \var{elem})} pairs.
\end{funcdesc}

\begin{funcdesc}{parse}{source\optional{, parser}}
Parses an XML section into an element tree.
\var{source} is a filename or file object containing XML data.
\var{parser} is an optional parser instance.  If not given, the
standard XMLTreeBuilder parser is used.
Returns an ElementTree instance.
\end{funcdesc}

\begin{funcdesc}{ProcessingInstruction}{target\optional{, text}}
PI element factory.  This factory function creates a special element
that will be serialized as an XML processing instruction.
\var{target} is a string containing the PI target.
\var{text} is a string containing the PI contents, if given.
Returns an element instance, representing a processing instruction.
\end{funcdesc}

\begin{funcdesc}{SubElement}{parent, tag\optional{,
                             attrib\optional{,  **extra}}}
Subelement factory.  This function creates an element instance, and
appends it to an existing element.

The element name, attribute names, and attribute values can be
either 8-bit ASCII strings or Unicode strings.
\var{parent} is the parent element.
\var{tag} is the subelement name.
\var{attrib} is an optional dictionary, containing element attributes.
\var{extra} contains additional attributes, given as keyword arguments.
Returns an element instance.
\end{funcdesc}

\begin{funcdesc}{tostring}{element\optional{, encoding}}
Generates a string representation of an XML element, including all
subelements.
\var{element} is an Element instance.
\var{encoding} is the output encoding (default is US-ASCII).
Returns an encoded string containing the XML data.
\end{funcdesc}

\begin{funcdesc}{XML}{text}
Parses an XML section from a string constant.  This function can
be used to embed ``XML literals'' in Python code.
\var{text} is a string containing XML data.
Returns an Element instance.
\end{funcdesc}

\begin{funcdesc}{XMLID}{text}
Parses an XML section from a string constant, and also returns
a dictionary which maps from element id:s to elements.
\var{text} is a string containing XML data.
Returns a tuple containing an Element instance and a dictionary.
\end{funcdesc}


\subsection{The Element Interface\label{elementtree-element-interface}}

Element objects returned by Element or SubElement have the 
following methods and attributes.

\begin{memberdesc}{tag}
A string identifying what kind of data this element represents
(the element type, in other words).
\end{memberdesc}

\begin{memberdesc}{text}
The \var{text} attribute can be used to hold additional data
associated with the element.
As the name implies this attribute is usually a string but may be any
application-specific object.
If the element is created from an XML file the attribute will contain
any text found between the element tags.
\end{memberdesc}

\begin{memberdesc}{tail}
The \var{tail} attribute can be used to hold additional data
associated with the element.
This attribute is usually a string but may be any application-specific object.
If the element is created from an XML file the attribute will contain
any text found after the element's end tag and before the next tag.
\end{memberdesc}

\begin{memberdesc}{attrib}
A dictionary containing the element's attributes.
Note that while the \var{attrib} value is always a real mutable Python
dictionary, an ElementTree implementation may choose to use another
internal representation, and create the dictionary only if someone
asks for it. To take advantage of such implementations, use the
dictionary methods below whenever possible.
\end{memberdesc}

The following dictionary-like methods work on the element attributes.

\begin{methoddesc}{clear}{}
Resets an element.  This function removes all subelements, clears
all attributes, and sets the text and tail attributes to None.
\end{methoddesc}

\begin{methoddesc}{get}{key\optional{, default=None}}
Gets the element attribute named \var{key}.

Returns the attribute value, or \var{default} if the
attribute was not found.
\end{methoddesc}

\begin{methoddesc}{items}{}
Returns the element attributes as a sequence of (name, value) pairs.
The attributes are returned in an arbitrary order.
\end{methoddesc}

\begin{methoddesc}{keys}{}
Returns the elements attribute names as a list.
The names are returned in an arbitrary order.
\end{methoddesc}

\begin{methoddesc}{set}{key, value}
Set the attribute \var{key} on the element to \var{value}.  
\end{methoddesc}

The following methods work on the element's children (subelements).

\begin{methoddesc}{append}{subelement}
Adds the element \var{subelement} to the end of this elements internal list
of subelements.
\end{methoddesc}

\begin{methoddesc}{find}{match}
Finds the first subelement matching \var{match}. 
\var{match} may be a tag name or path.
Returns an element instance or \code{None}.
\end{methoddesc}

\begin{methoddesc}{findall}{match}
Finds all subelements matching \var{match}. 
\var{match} may be a tag name or path.
Returns an iterable yielding all matching elements in document order.
\end{methoddesc}

\begin{methoddesc}{findtext}{condition\optional{, default=None}}
Finds text for the first subelement matching \var{condition}. 
\var{condition} may be a tag name or path.
Returns the text content of the first matching element, or
\var{default} if no element was found.  Note that if the
matching element has no text content an empty string is returned.
\end{methoddesc}

\begin{methoddesc}{getchildren}{}
Returns all subelements.  The elements are returned in document order.
\end{methoddesc}

\begin{methoddesc}{getiterator}{\optional{tag=None}}
Creates a tree iterator with the current element as the root.  
The iterator iterates over this element and all elements below it 
that match the given tag. If tag
is \code{None} or \code{'*'} then all elements are iterated over.
Returns an iterable that provides element objects in document (depth first)
order.
\end{methoddesc}

\begin{methoddesc}{insert}{index, element}
Inserts a subelement at the given position in this element.
\end{methoddesc}

\begin{methoddesc}{makeelement}{tag, attrib}
Creates a new element object of the same type as this element.
Do not call this method, use the SubElement factory function instead.
\end{methoddesc}

\begin{methoddesc}{remove}{subelement}
Removes \var{subelement} from the element.  
Unlike the findXXX methods this method compares elements based on 
the instance identity, not on tag value or contents.
\end{methoddesc}

Element objects also support the following sequence type methods for
working with subelements: \method{__delitem__()},
\method{__getitem__()}, \method{__setitem__()}, \method{__len__()}.

Caution: Because Element objects do not define a
\method{__nonzero__()} method, elements with no subelements will test
as \code{False}.

\begin{verbatim}
element = root.find('foo')

if not element: # careful!
    print "element not found, or element has no subelements"

if element is None:
    print "element not found"
\end{verbatim}


\subsection{ElementTree Objects\label{elementtree-elementtree-objects}}

\begin{classdesc}{ElementTree}{\optional{element,} \optional{file}}
ElementTree wrapper class.  This class represents an entire element
hierarchy, and adds some extra support for serialization to and from
standard XML.

\var{element} is the root element.
The tree is initialized with the contents of the XML \var{file} if given.
\end{classdesc}

\begin{methoddesc}{_setroot}{element}
Replaces the root element for this tree.  This discards the
current contents of the tree, and replaces it with the given
element.  Use with care.
\var{element} is an element instance.
\end{methoddesc}

\begin{methoddesc}{find}{path}
Finds the first toplevel element with given tag.
Same as getroot().find(path).
\var{path} is the element to look for.
Returns the first matching element, or \code{None} if no element was found.
\end{methoddesc}

\begin{methoddesc}{findall}{path}
Finds all toplevel elements with the given tag.
Same as getroot().findall(path).
\var{path} is the element to look for.
Returns a list or iterator containing all matching elements,
in document order.
\end{methoddesc}

\begin{methoddesc}{findtext}{path\optional{, default}}
Finds the element text for the first toplevel element with given
tag.  Same as getroot().findtext(path).
\var{path} is the toplevel element to look for.
\var{default} is the value to return if the element was not found.
Returns the text content of the first matching element, or the
default value no element was found.  Note that if the element
has is found, but has no text content, this method returns an
empty string.
\end{methoddesc}

\begin{methoddesc}{getiterator}{\optional{tag}}
Creates and returns a tree iterator for the root element.  The iterator loops
over all elements in this tree, in section order.
\var{tag} is the tag to look for (default is to return all elements)
\end{methoddesc}

\begin{methoddesc}{getroot}{}
Returns the root element for this tree.
\end{methoddesc}

\begin{methoddesc}{parse}{source\optional{, parser}}
Loads an external XML section into this element tree.
\var{source} is a file name or file object.
\var{parser} is an optional parser instance.  If not given, the
standard XMLTreeBuilder parser is used.
Returns the section root element.
\end{methoddesc}

\begin{methoddesc}{write}{file\optional{, encoding}}
Writes the element tree to a file, as XML.
\var{file} is a file name, or a file object opened for writing.
\var{encoding} is the output encoding (default is US-ASCII).
\end{methoddesc}


\subsection{QName Objects\label{elementtree-qname-objects}}

\begin{classdesc}{QName}{text_or_uri\optional{, tag}}
QName wrapper.  This can be used to wrap a QName attribute value, in
order to get proper namespace handling on output.
\var{text_or_uri} is a string containing the QName value,
in the form {\{}uri{\}}local, or, if the tag argument is given,
the URI part of a QName.
If \var{tag} is given, the first argument is interpreted as
an URI, and this argument is interpreted as a local name.
\class{QName} instances are opaque.
\end{classdesc}


\subsection{TreeBuilder Objects\label{elementtree-treebuilder-objects}}

\begin{classdesc}{TreeBuilder}{\optional{element_factory}}
Generic element structure builder.  This builder converts a sequence
of start, data, and end method calls to a well-formed element structure.
You can use this class to build an element structure using a custom XML
parser, or a parser for some other XML-like format.
The \var{element_factory} is called to create new Element instances when
given.
\end{classdesc}

\begin{methoddesc}{close}{}
Flushes the parser buffers, and returns the toplevel documen
element.
Returns an Element instance.
\end{methoddesc}

\begin{methoddesc}{data}{data}
Adds text to the current element.
\var{data} is a string.  This should be either an 8-bit string
containing ASCII text, or a Unicode string.
\end{methoddesc}

\begin{methoddesc}{end}{tag}
Closes the current element.
\var{tag} is the element name.
Returns the closed element.
\end{methoddesc}

\begin{methoddesc}{start}{tag, attrs}
Opens a new element.
\var{tag} is the element name.
\var{attrs} is a dictionary containing element attributes.
Returns the opened element.
\end{methoddesc}


\subsection{XMLTreeBuilder Objects\label{elementtree-xmltreebuilder-objects}}

\begin{classdesc}{XMLTreeBuilder}{\optional{html,} \optional{target}}
Element structure builder for XML source data, based on the
expat parser.
\var{html} are predefined HTML entities.  This flag is not supported
by the current implementation.
\var{target} is the target object.  If omitted, the builder uses an
instance of the standard TreeBuilder class.
\end{classdesc}

\begin{methoddesc}{close}{}
Finishes feeding data to the parser.
Returns an element structure.
\end{methoddesc}

\begin{methoddesc}{doctype}{name, pubid, system}
Handles a doctype declaration.
\var{name} is the doctype name.
\var{pubid} is the public identifier.
\var{system} is the system identifier.
\end{methoddesc}

\begin{methoddesc}{feed}{data}
Feeds data to the parser.
\var{data} is encoded data.
\end{methoddesc}
