\section{\module{fcntl} ---
         The \function{fcntl()} and \function{ioctl()} system calls}

\declaremodule{builtin}{fcntl}
  \platform{Unix}
\modulesynopsis{The \function{fcntl()} and \function{ioctl()} system calls.}
\sectionauthor{Jaap Vermeulen}{}

\indexii{UNIX@\UNIX}{file control}
\indexii{UNIX@\UNIX}{I/O control}

This module performs file control and I/O control on file descriptors.
It is an interface to the \cfunction{fcntl()} and \cfunction{ioctl()}
\UNIX{} routines.

All functions in this module take a file descriptor \var{fd} as their
first argument.  This can be an integer file descriptor, such as
returned by \code{sys.stdin.fileno()}, or a file object, such as
\code{sys.stdin} itself, which provides a \method{fileno()} which
returns a genuine file descriptor.

The module defines the following functions:


\begin{funcdesc}{fcntl}{fd, op\optional{, arg}}
  Perform the requested operation on file descriptor \var{fd} (file
  objects providing a \method{fileno()} method are accepted as well).
  The operation is defined by \var{op} and is operating system
  dependent.  These codes are also found in the \module{fcntl}
  module. The argument \var{arg} is optional, and defaults to the
  integer value \code{0}.  When present, it can either be an integer
  value, or a string.  With the argument missing or an integer value,
  the return value of this function is the integer return value of the
  C \cfunction{fcntl()} call.  When the argument is a string it
  represents a binary structure, e.g.\ created by
  \function{\refmodule{struct}.pack()}. The binary data is copied to a buffer
  whose address is passed to the C \cfunction{fcntl()} call.  The
  return value after a successful call is the contents of the buffer,
  converted to a string object.  The length of the returned string
  will be the same as the length of the \var{arg} argument.  This is
  limited to 1024 bytes.  If the information returned in the buffer by
  the operating system is larger than 1024 bytes, this is most likely
  to result in a segmentation violation or a more subtle data
  corruption.

  If the \cfunction{fcntl()} fails, an \exception{IOError} is
  raised.
\end{funcdesc}

\begin{funcdesc}{ioctl}{fd, op\optional{, arg\optional{, mutate_flag}}}
  This function is identical to the \function{fcntl()} function,
  except that the operations are typically defined in the library
  module \refmodule{termios} and the argument handling is even more
  complicated.
  
  The parameter \var{arg} can be one of an integer, absent (treated
  identically to the integer \code{0}), an object supporting the
  read-only buffer interface (most likely a plain Python string) or an
  object supporting the read-write buffer interface.
  
  In all but the last case, behaviour is as for the \function{fcntl()}
  function.
  
  If a mutable buffer is passed, then the behaviour is determined by
  the value of the \var{mutate_flag} parameter.
  
  If it is false, the buffer's mutability is ignored and behaviour is
  as for a read-only buffer, except that the 1024 byte limit mentioned
  above is avoided -- so long as the buffer you pass is longer than
  what the operating system wants to put there, things should work.
  
  If \var{mutate_flag} is true, then the buffer is (in effect) passed
  to the underlying \function{ioctl()} system call, the latter's
  return code is passed back to the calling Python, and the buffer's
  new contents reflect the action of the \function{ioctl()}.  This is a
  slight simplification, because if the supplied buffer is less than
  1024 bytes long it is first copied into a static buffer 1024 bytes
  long which is then passed to \function{ioctl()} and copied back into
  the supplied buffer.
  
  If \var{mutate_flag} is not supplied, then in 2.3 it defaults to
  false.  This is planned to change over the next few Python versions:
  in 2.4 failing to supply \var{mutate_flag} will get a warning but
  the same behavior and in versions later than 2.5 it will default to
  true.

  An example:

\begin{verbatim}
>>> import array, fcntl, struct, termios, os
>>> os.getpgrp()
13341
>>> struct.unpack('h', fcntl.ioctl(0, termios.TIOCGPGRP, "  "))[0]
13341
>>> buf = array.array('h', [0])
>>> fcntl.ioctl(0, termios.TIOCGPGRP, buf, 1)
0
>>> buf
array('h', [13341])
\end{verbatim}
\end{funcdesc}

\begin{funcdesc}{flock}{fd, op}
Perform the lock operation \var{op} on file descriptor \var{fd} (file
  objects providing a \method{fileno()} method are accepted as well).
See the \UNIX{} manual \manpage{flock}{3} for details.  (On some
systems, this function is emulated using \cfunction{fcntl()}.)
\end{funcdesc}

\begin{funcdesc}{lockf}{fd, operation,
    \optional{length, \optional{start, \optional{whence}}}}
This is essentially a wrapper around the \function{fcntl()} locking
calls.  \var{fd} is the file descriptor of the file to lock or unlock,
and \var{operation} is one of the following values:

\begin{itemize}
\item \constant{LOCK_UN} -- unlock
\item \constant{LOCK_SH} -- acquire a shared lock
\item \constant{LOCK_EX} -- acquire an exclusive lock
\end{itemize}

When \var{operation} is \constant{LOCK_SH} or \constant{LOCK_EX}, it
can also be bit-wise OR'd with \constant{LOCK_NB} to avoid blocking on
lock acquisition.  If \constant{LOCK_NB} is used and the lock cannot
be acquired, an \exception{IOError} will be raised and the exception
will have an \var{errno} attribute set to \constant{EACCES} or
\constant{EAGAIN} (depending on the operating system; for portability,
check for both values).  On at least some systems, \constant{LOCK_EX}
can only be used if the file descriptor refers to a file opened for
writing.

\var{length} is the number of bytes to lock, \var{start} is the byte
offset at which the lock starts, relative to \var{whence}, and
\var{whence} is as with \function{fileobj.seek()}, specifically:

\begin{itemize}
\item \constant{0} -- relative to the start of the file
      (\constant{SEEK_SET})
\item \constant{1} -- relative to the current buffer position
      (\constant{SEEK_CUR})
\item \constant{2} -- relative to the end of the file
      (\constant{SEEK_END})
\end{itemize}

The default for \var{start} is 0, which means to start at the
beginning of the file.  The default for \var{length} is 0 which means
to lock to the end of the file.  The default for \var{whence} is also
0.
\end{funcdesc}

Examples (all on a SVR4 compliant system):

\begin{verbatim}
import struct, fcntl, os

f = open(...)
rv = fcntl.fcntl(f, fcntl.F_SETFL, os.O_NDELAY)

lockdata = struct.pack('hhllhh', fcntl.F_WRLCK, 0, 0, 0, 0, 0)
rv = fcntl.fcntl(f, fcntl.F_SETLKW, lockdata)
\end{verbatim}

Note that in the first example the return value variable \var{rv} will
hold an integer value; in the second example it will hold a string
value.  The structure lay-out for the \var{lockdata} variable is
system dependent --- therefore using the \function{flock()} call may be
better.

\begin{seealso}
  \seemodule{os}{The \function{os.open()} function supports locking flags
                 and is available on a wider variety of platforms than
                 the \function{lockf()} and \function{flock()}
                 functions, providing a more platform-independent file
                 locking facility.}
\end{seealso}
