\section{\module{imaplib} ---
         IMAP4 protocol client}

\declaremodule{standard}{imaplib}
\modulesynopsis{IMAP4 protocol client (requires sockets).}
\moduleauthor{Piers Lauder}{piers@communitysolutions.com.au}
\sectionauthor{Piers Lauder}{piers@communitysolutions.com.au}

% Based on HTML documentation by Piers Lauder
% <piers@communitysolutions.com.au>;
% converted by Fred L. Drake, Jr. <fdrake@acm.org>.
% Revised by ESR, January 2000.
% Changes for IMAP4_SSL by Tino Lange <Tino.Lange@isg.de>, March 2002 
% Changes for IMAP4_stream by Piers Lauder
% <piers@communitysolutions.com.au>, November 2002

\indexii{IMAP4}{protocol}
\indexii{IMAP4_SSL}{protocol}
\indexii{IMAP4_stream}{protocol}

This module defines three classes, \class{IMAP4}, \class{IMAP4_SSL}
and \class{IMAP4_stream}, which encapsulate a
connection to an IMAP4 server and implement a large subset of the
IMAP4rev1 client protocol as defined in \rfc{2060}. It is backward
compatible with IMAP4 (\rfc{1730}) servers, but note that the
\samp{STATUS} command is not supported in IMAP4.

Three classes are provided by the \module{imaplib} module,
\class{IMAP4} is the base class:

\begin{classdesc}{IMAP4}{\optional{host\optional{, port}}}
This class implements the actual IMAP4 protocol.  The connection is
created and protocol version (IMAP4 or IMAP4rev1) is determined when
the instance is initialized.
If \var{host} is not specified, \code{''} (the local host) is used.
If \var{port} is omitted, the standard IMAP4 port (143) is used.
\end{classdesc}

Three exceptions are defined as attributes of the \class{IMAP4} class:

\begin{excdesc}{IMAP4.error}
Exception raised on any errors.  The reason for the exception is
passed to the constructor as a string.
\end{excdesc}

\begin{excdesc}{IMAP4.abort}
IMAP4 server errors cause this exception to be raised.  This is a
sub-class of \exception{IMAP4.error}.  Note that closing the instance
and instantiating a new one will usually allow recovery from this
exception.
\end{excdesc}

\begin{excdesc}{IMAP4.readonly}
This exception is raised when a writable mailbox has its status
changed by the server.  This is a sub-class of
\exception{IMAP4.error}.  Some other client now has write permission,
and the mailbox will need to be re-opened to re-obtain write
permission.
\end{excdesc}

There's also a subclass for secure connections:

\begin{classdesc}{IMAP4_SSL}{\optional{host\optional{, port\optional{,
                             keyfile\optional{, certfile}}}}}
This is a subclass derived from \class{IMAP4} that connects over an
SSL encrypted socket (to use this class you need a socket module that
was compiled with SSL support).  If \var{host} is not specified,
\code{''} (the local host) is used.  If \var{port} is omitted, the
standard IMAP4-over-SSL port (993) is used.  \var{keyfile} and
\var{certfile} are also optional - they can contain a PEM formatted
private key and certificate chain file for the SSL connection.
\end{classdesc}

The second subclass allows for connections created by a child process:

\begin{classdesc}{IMAP4_stream}{command}
This is a subclass derived from \class{IMAP4} that connects
to the \code{stdin/stdout} file descriptors created by passing
\var{command} to \code{os.popen2()}.
\versionadded{2.3}
\end{classdesc}

The following utility functions are defined:

\begin{funcdesc}{Internaldate2tuple}{datestr}
  Converts an IMAP4 INTERNALDATE string to Coordinated Universal
  Time. Returns a \refmodule{time} module tuple.
\end{funcdesc}

\begin{funcdesc}{Int2AP}{num}
  Converts an integer into a string representation using characters
  from the set [\code{A} .. \code{P}].
\end{funcdesc}

\begin{funcdesc}{ParseFlags}{flagstr}
  Converts an IMAP4 \samp{FLAGS} response to a tuple of individual
  flags.
\end{funcdesc}

\begin{funcdesc}{Time2Internaldate}{date_time}
  Converts a \refmodule{time} module tuple to an IMAP4
  \samp{INTERNALDATE} representation.  Returns a string in the form:
  \code{"DD-Mmm-YYYY HH:MM:SS +HHMM"} (including double-quotes).
\end{funcdesc}


Note that IMAP4 message numbers change as the mailbox changes; in
particular, after an \samp{EXPUNGE} command performs deletions the
remaining messages are renumbered. So it is highly advisable to use
UIDs instead, with the UID command.

At the end of the module, there is a test section that contains a more
extensive example of usage.

\begin{seealso}
  \seetext{Documents describing the protocol, and sources and binaries 
           for servers implementing it, can all be found at the
           University of Washington's \emph{IMAP Information Center}
           (\url{http://www.cac.washington.edu/imap/}).}
\end{seealso}


\subsection{IMAP4 Objects \label{imap4-objects}}

All IMAP4rev1 commands are represented by methods of the same name,
either upper-case or lower-case.

All arguments to commands are converted to strings, except for
\samp{AUTHENTICATE}, and the last argument to \samp{APPEND} which is
passed as an IMAP4 literal.  If necessary (the string contains IMAP4
protocol-sensitive characters and isn't enclosed with either
parentheses or double quotes) each string is quoted. However, the
\var{password} argument to the \samp{LOGIN} command is always quoted.
If you want to avoid having an argument string quoted
(eg: the \var{flags} argument to \samp{STORE}) then enclose the string in
parentheses (eg: \code{r'(\e Deleted)'}).

Each command returns a tuple: \code{(\var{type}, [\var{data},
...])} where \var{type} is usually \code{'OK'} or \code{'NO'},
and \var{data} is either the text from the command response, or
mandated results from the command. Each \var{data}
is either a string, or a tuple. If a tuple, then the first part
is the header of the response, and the second part contains
the data (ie: 'literal' value).

The \var{message_set} options to commands below is a string specifying one
or more messages to be acted upon.  It may be a simple message number
(\code{'1'}), a range of message numbers (\code{'2:4'}), or a group of
non-contiguous ranges separated by commas (\code{'1:3,6:9'}).  A range
can contain an asterisk to indicate an infinite upper bound
(\code{'3:*'}).

An \class{IMAP4} instance has the following methods:


\begin{methoddesc}{append}{mailbox, flags, date_time, message}
  Append \var{message} to named mailbox. 
\end{methoddesc}

\begin{methoddesc}{authenticate}{mechanism, authobject}
  Authenticate command --- requires response processing.

  \var{mechanism} specifies which authentication mechanism is to be
  used - it should appear in the instance variable \code{capabilities}
  in the form \code{AUTH=mechanism}.

  \var{authobject} must be a callable object:

\begin{verbatim}
data = authobject(response)
\end{verbatim}

  It will be called to process server continuation responses.
  It should return \code{data} that will be encoded and sent to server.
  It should return \code{None} if the client abort response \samp{*} should
  be sent instead.
\end{methoddesc}

\begin{methoddesc}{check}{}
  Checkpoint mailbox on server. 
\end{methoddesc}

\begin{methoddesc}{close}{}
  Close currently selected mailbox. Deleted messages are removed from
  writable mailbox. This is the recommended command before
  \samp{LOGOUT}.
\end{methoddesc}

\begin{methoddesc}{copy}{message_set, new_mailbox}
  Copy \var{message_set} messages onto end of \var{new_mailbox}. 
\end{methoddesc}

\begin{methoddesc}{create}{mailbox}
  Create new mailbox named \var{mailbox}.
\end{methoddesc}

\begin{methoddesc}{delete}{mailbox}
  Delete old mailbox named \var{mailbox}.
\end{methoddesc}

\begin{methoddesc}{deleteacl}{mailbox, who}
  Delete the ACLs (remove any rights) set for who on mailbox.
\versionadded{2.4}
\end{methoddesc}

\begin{methoddesc}{expunge}{}
  Permanently remove deleted items from selected mailbox. Generates an
  \samp{EXPUNGE} response for each deleted message. Returned data
  contains a list of \samp{EXPUNGE} message numbers in order
  received.
\end{methoddesc}

\begin{methoddesc}{fetch}{message_set, message_parts}
  Fetch (parts of) messages.  \var{message_parts} should be
  a string of message part names enclosed within parentheses,
  eg: \samp{"(UID BODY[TEXT])"}.  Returned data are tuples
  of message part envelope and data.
\end{methoddesc}

\begin{methoddesc}{getacl}{mailbox}
  Get the \samp{ACL}s for \var{mailbox}.
  The method is non-standard, but is supported by the \samp{Cyrus} server.
\end{methoddesc}

\begin{methoddesc}{getquota}{root}
  Get the \samp{quota} \var{root}'s resource usage and limits.
  This method is part of the IMAP4 QUOTA extension defined in rfc2087.
\versionadded{2.3}
\end{methoddesc}

\begin{methoddesc}{getquotaroot}{mailbox}
  Get the list of \samp{quota} \samp{roots} for the named \var{mailbox}.
  This method is part of the IMAP4 QUOTA extension defined in rfc2087.
\versionadded{2.3}
\end{methoddesc}

\begin{methoddesc}{list}{\optional{directory\optional{, pattern}}}
  List mailbox names in \var{directory} matching
  \var{pattern}.  \var{directory} defaults to the top-level mail
  folder, and \var{pattern} defaults to match anything.  Returned data
  contains a list of \samp{LIST} responses.
\end{methoddesc}

\begin{methoddesc}{login}{user, password}
  Identify the client using a plaintext password.
  The \var{password} will be quoted.
\end{methoddesc}

\begin{methoddesc}{login_cram_md5}{user, password}
  Force use of \samp{CRAM-MD5} authentication when identifying the
  client to protect the password.  Will only work if the server
  \samp{CAPABILITY} response includes the phrase \samp{AUTH=CRAM-MD5}.
\versionadded{2.3}
\end{methoddesc}

\begin{methoddesc}{logout}{}
  Shutdown connection to server. Returns server \samp{BYE} response.
\end{methoddesc}

\begin{methoddesc}{lsub}{\optional{directory\optional{, pattern}}}
  List subscribed mailbox names in directory matching pattern.
  \var{directory} defaults to the top level directory and
  \var{pattern} defaults to match any mailbox.
  Returned data are tuples of message part envelope and data.
\end{methoddesc}

\begin{methoddesc}{myrights}{mailbox}
  Show my ACLs for a mailbox (i.e. the rights that I have on mailbox).
\versionadded{2.4}
\end{methoddesc}

\begin{methoddesc}{namespace}{}
  Returns IMAP namespaces as defined in RFC2342.
\versionadded{2.3}
\end{methoddesc}

\begin{methoddesc}{noop}{}
  Send \samp{NOOP} to server.
\end{methoddesc}

\begin{methoddesc}{open}{host, port}
  Opens socket to \var{port} at \var{host}.
  The connection objects established by this method
  will be used in the \code{read}, \code{readline}, \code{send}, and
  \code{shutdown} methods.
  You may override this method.
\end{methoddesc}

\begin{methoddesc}{partial}{message_num, message_part, start, length}
  Fetch truncated part of a message.
  Returned data is a tuple of message part envelope and data.
\end{methoddesc}

\begin{methoddesc}{proxyauth}{user}
  Assume authentication as \var{user}.
  Allows an authorised administrator to proxy into any user's mailbox.
\versionadded{2.3}
\end{methoddesc}

\begin{methoddesc}{read}{size}
  Reads \var{size} bytes from the remote server.
  You may override this method.
\end{methoddesc}

\begin{methoddesc}{readline}{}
  Reads one line from the remote server.
  You may override this method.
\end{methoddesc}

\begin{methoddesc}{recent}{}
  Prompt server for an update. Returned data is \code{None} if no new
  messages, else value of \samp{RECENT} response.
\end{methoddesc}

\begin{methoddesc}{rename}{oldmailbox, newmailbox}
  Rename mailbox named \var{oldmailbox} to \var{newmailbox}.
\end{methoddesc}

\begin{methoddesc}{response}{code}
  Return data for response \var{code} if received, or
  \code{None}. Returns the given code, instead of the usual type.
\end{methoddesc}

\begin{methoddesc}{search}{charset, criterion\optional{, ...}}
  Search mailbox for matching messages.  Returned data contains a space
  separated list of matching message numbers.  \var{charset} may be
  \code{None}, in which case no \samp{CHARSET} will be specified in the
  request to the server.  The IMAP protocol requires that at least one
  criterion be specified; an exception will be raised when the server
  returns an error.

  Example:

\begin{verbatim}
# M is a connected IMAP4 instance...
msgnums = M.search(None, 'FROM', '"LDJ"')

# or:
msgnums = M.search(None, '(FROM "LDJ")')
\end{verbatim}
\end{methoddesc}

\begin{methoddesc}{select}{\optional{mailbox\optional{, readonly}}}
  Select a mailbox. Returned data is the count of messages in
  \var{mailbox} (\samp{EXISTS} response).  The default \var{mailbox}
  is \code{'INBOX'}.  If the \var{readonly} flag is set, modifications
  to the mailbox are not allowed.
\end{methoddesc}

\begin{methoddesc}{send}{data}
  Sends \code{data} to the remote server.
  You may override this method.
\end{methoddesc}

\begin{methoddesc}{setacl}{mailbox, who, what}
  Set an \samp{ACL} for \var{mailbox}.
  The method is non-standard, but is supported by the \samp{Cyrus} server.
\end{methoddesc}

\begin{methoddesc}{setquota}{root, limits}
  Set the \samp{quota} \var{root}'s resource \var{limits}.
  This method is part of the IMAP4 QUOTA extension defined in rfc2087.
\versionadded{2.3}
\end{methoddesc}

\begin{methoddesc}{shutdown}{}
  Close connection established in \code{open}.
  You may override this method.
\end{methoddesc}

\begin{methoddesc}{socket}{}
  Returns socket instance used to connect to server.
\end{methoddesc}

\begin{methoddesc}{sort}{sort_criteria, charset, search_criterion\optional{, ...}}
  The \code{sort} command is a variant of \code{search} with sorting
  semantics for the results.  Returned data contains a space separated
  list of matching message numbers.

  Sort has two arguments before the \var{search_criterion}
  argument(s); a parenthesized list of \var{sort_criteria}, and the
  searching \var{charset}.  Note that unlike \code{search}, the
  searching \var{charset} argument is mandatory.  There is also a
  \code{uid sort} command which corresponds to \code{sort} the way
  that \code{uid search} corresponds to \code{search}.  The
  \code{sort} command first searches the mailbox for messages that
  match the given searching criteria using the charset argument for
  the interpretation of strings in the searching criteria.  It then
  returns the numbers of matching messages.

  This is an \samp{IMAP4rev1} extension command.
\end{methoddesc}

\begin{methoddesc}{status}{mailbox, names}
  Request named status conditions for \var{mailbox}. 
\end{methoddesc}

\begin{methoddesc}{store}{message_set, command, flag_list}
  Alters flag dispositions for messages in mailbox.  \var{command} is
  specified by section 6.4.6 of \rfc{2060} as being one of "FLAGS", "+FLAGS",
  or "-FLAGS", optionally with a suffix of ".SILENT".

  For example, to set the delete flag on all messages:

\begin{verbatim}
typ, data = M.search(None, 'ALL')
for num in data[0].split():
   M.store(num, '+FLAGS', '\\Deleted')
M.expunge()
\end{verbatim}
\end{methoddesc}

\begin{methoddesc}{subscribe}{mailbox}
  Subscribe to new mailbox.
\end{methoddesc}

\begin{methoddesc}{thread}{threading_algorithm, charset,
                           search_criterion\optional{, ...}}
  The \code{thread} command is a variant of \code{search} with
  threading semantics for the results.  Returned data contains a space
  separated list of thread members.

  Thread members consist of zero or more messages numbers, delimited
  by spaces, indicating successive parent and child.

  Thread has two arguments before the \var{search_criterion}
  argument(s); a \var{threading_algorithm}, and the searching
  \var{charset}.  Note that unlike \code{search}, the searching
  \var{charset} argument is mandatory.  There is also a \code{uid
  thread} command which corresponds to \code{thread} the way that
  \code{uid search} corresponds to \code{search}.  The \code{thread}
  command first searches the mailbox for messages that match the given
  searching criteria using the charset argument for the interpretation
  of strings in the searching criteria. It then returns the matching
  messages threaded according to the specified threading algorithm.

  This is an \samp{IMAP4rev1} extension command. \versionadded{2.4}
\end{methoddesc}

\begin{methoddesc}{uid}{command, arg\optional{, ...}}
  Execute command args with messages identified by UID, rather than
  message number.  Returns response appropriate to command.  At least
  one argument must be supplied; if none are provided, the server will
  return an error and an exception will be raised.
\end{methoddesc}

\begin{methoddesc}{unsubscribe}{mailbox}
  Unsubscribe from old mailbox.
\end{methoddesc}

\begin{methoddesc}{xatom}{name\optional{, arg\optional{, ...}}}
  Allow simple extension commands notified by server in
  \samp{CAPABILITY} response.
\end{methoddesc}


Instances of \class{IMAP4_SSL} have just one additional method:

\begin{methoddesc}{ssl}{}
  Returns SSLObject instance used for the secure connection with the server.
\end{methoddesc}


The following attributes are defined on instances of \class{IMAP4}:


\begin{memberdesc}{PROTOCOL_VERSION}
The most recent supported protocol in the
\samp{CAPABILITY} response from the server.
\end{memberdesc}

\begin{memberdesc}{debug}
Integer value to control debugging output.  The initialize value is
taken from the module variable \code{Debug}.  Values greater than
three trace each command.
\end{memberdesc}


\subsection{IMAP4 Example \label{imap4-example}}

Here is a minimal example (without error checking) that opens a
mailbox and retrieves and prints all messages:

\begin{verbatim}
import getpass, imaplib

M = imaplib.IMAP4()
M.login(getpass.getuser(), getpass.getpass())
M.select()
typ, data = M.search(None, 'ALL')
for num in data[0].split():
    typ, data = M.fetch(num, '(RFC822)')
    print 'Message %s\n%s\n' % (num, data[0][1])
M.close()
M.logout()
\end{verbatim}
