\section{\module{mmap} ---
Memory-mapped file support}

\declaremodule{builtin}{mmap}
\modulesynopsis{Interface to memory-mapped files for \UNIX\ and Windows.}

Memory-mapped file objects behave like both strings and like
file objects.  Unlike normal string objects, however, these are
mutable.  You can use mmap objects in most places where strings
are expected; for example, you can use the \module{re} module to
search through a memory-mapped file.  Since they're mutable, you can
change a single character by doing \code{obj[\var{index}] = 'a'}, or
change a substring by assigning to a slice:
\code{obj[\var{i1}:\var{i2}] = '...'}.  You can also read and write
data starting at the current file position, and \method{seek()}
through the file to different positions.

A memory-mapped file is created by the \function{mmap()} function,
which is different on \UNIX{} and on Windows.  In either case you must
provide a file descriptor for a file opened for update.
If you wish to map an existing Python file object, use its
\method{fileno()} method to obtain the correct value for the
\var{fileno} parameter.  Otherwise, you can open the file using the
\function{os.open()} function, which returns a file descriptor
directly (the file still needs to be closed when done).

For both the \UNIX{} and Windows versions of the function,
\var{access} may be specified as an optional keyword parameter.
\var{access} accepts one of three values: \constant{ACCESS_READ},
\constant{ACCESS_WRITE}, or \constant{ACCESS_COPY} to specify
readonly, write-through or copy-on-write memory respectively.
\var{access} can be used on both \UNIX{} and Windows.  If
\var{access} is not specified, Windows mmap returns a write-through
mapping.  The initial memory values for all three access types are
taken from the specified file.  Assignment to an
\constant{ACCESS_READ} memory map raises a \exception{TypeError}
exception.  Assignment to an \constant{ACCESS_WRITE} memory map
affects both memory and the underlying file.  Assignment to an
\constant{ACCESS_COPY} memory map affects memory but does not update
the underlying file.

\begin{funcdesc}{mmap}{fileno, length\optional{, tagname\optional{, access}}}
  \strong{(Windows version)} Maps \var{length} bytes from the file
  specified by the file handle \var{fileno}, and returns a mmap
  object.  If \var{length} is larger than the current size of the file,
  the file is extended to contain \var{length} bytes.  If \var{length}
  is \code{0}, the maximum length of the map is the current size
  of the file, except that if the file is empty Windows raises an
  exception (you cannot create an empty mapping on Windows).

  \var{tagname}, if specified and not \code{None}, is a string giving
  a tag name for the mapping.  Windows allows you to have many
  different mappings against the same file.  If you specify the name
  of an existing tag, that tag is opened, otherwise a new tag of this
  name is created.  If this parameter is omitted or \code{None}, the
  mapping is created without a name.  Avoiding the use of the tag
  parameter will assist in keeping your code portable between \UNIX{}
  and Windows.
\end{funcdesc}

\begin{funcdescni}{mmap}{fileno, length\optional{, flags\optional{,
                         prot\optional{, access}}}}
  \strong{(\UNIX{} version)} Maps \var{length} bytes from the file
  specified by the file descriptor \var{fileno}, and returns a mmap
  object.

  \var{flags} specifies the nature of the mapping.
  \constant{MAP_PRIVATE} creates a private copy-on-write mapping, so
  changes to the contents of the mmap object will be private to this
  process, and \constant{MAP_SHARED} creates a mapping that's shared
  with all other processes mapping the same areas of the file.  The
  default value is \constant{MAP_SHARED}.

  \var{prot}, if specified, gives the desired memory protection; the
  two most useful values are \constant{PROT_READ} and
  \constant{PROT_WRITE}, to specify that the pages may be read or
  written.  \var{prot} defaults to \constant{PROT_READ | PROT_WRITE}.

  \var{access} may be specified in lieu of \var{flags} and \var{prot}
  as an optional keyword parameter.  It is an error to specify both
  \var{flags}, \var{prot} and \var{access}.  See the description of
  \var{access} above for information on how to use this parameter.
\end{funcdescni}


Memory-mapped file objects support the following methods:


\begin{methoddesc}{close}{}
  Close the file.  Subsequent calls to other methods of the object
  will result in an exception being raised.
\end{methoddesc}

\begin{methoddesc}{find}{string\optional{, start}}
  Returns the lowest index in the object where the substring
  \var{string} is found.  Returns \code{-1} on failure.  \var{start}
  is the index at which the search begins, and defaults to zero.
\end{methoddesc}

\begin{methoddesc}{flush}{\optional{offset, size}}
  Flushes changes made to the in-memory copy of a file back to disk.
  Without use of this call there is no guarantee that changes are
  written back before the object is destroyed.  If \var{offset} and
  \var{size} are specified, only changes to the given range of bytes
  will be flushed to disk; otherwise, the whole extent of the mapping
  is flushed.
\end{methoddesc}

\begin{methoddesc}{move}{\var{dest}, \var{src}, \var{count}}
  Copy the \var{count} bytes starting at offset \var{src} to the
  destination index \var{dest}.  If the mmap was created with
  \constant{ACCESS_READ}, then calls to move will throw a
  \exception{TypeError} exception.
\end{methoddesc}

\begin{methoddesc}{read}{\var{num}}
  Return a string containing up to \var{num} bytes starting from the
  current file position; the file position is updated to point after the
  bytes that were returned.
\end{methoddesc}

\begin{methoddesc}{read_byte}{}
  Returns a string of length 1 containing the character at the current
  file position, and advances the file position by 1.
\end{methoddesc}

\begin{methoddesc}{readline}{}
  Returns a single line, starting at the current file position and up to
  the next newline.
\end{methoddesc}

\begin{methoddesc}{resize}{\var{newsize}}
  If the mmap was created with \constant{ACCESS_READ} or
  \constant{ACCESS_COPY}, resizing the map will throw a \exception{TypeError} exception.
\end{methoddesc}

\begin{methoddesc}{seek}{pos\optional{, whence}}
  Set the file's current position.  \var{whence} argument is optional
  and defaults to \code{0} (absolute file positioning); other values
  are \code{1} (seek relative to the current position) and \code{2}
  (seek relative to the file's end).
\end{methoddesc}

\begin{methoddesc}{size}{}
  Return the length of the file, which can be larger than the size of
  the memory-mapped area.
\end{methoddesc}

\begin{methoddesc}{tell}{}
  Returns the current position of the file pointer.
\end{methoddesc}

\begin{methoddesc}{write}{\var{string}}
  Write the bytes in \var{string} into memory at the current position
  of the file pointer; the file position is updated to point after the
  bytes that were written. If the mmap was created with
  \constant{ACCESS_READ}, then writing to it will throw a
  \exception{TypeError} exception.
\end{methoddesc}

\begin{methoddesc}{write_byte}{\var{byte}}
  Write the single-character string \var{byte} into memory at the
  current position of the file pointer; the file position is advanced
  by \code{1}.If the mmap was created with \constant{ACCESS_READ},
  then writing to it will throw a \exception{TypeError} exception.
\end{methoddesc}
