\section{\module{popen2} ---
         Subprocesses with accessible I/O streams}

\declaremodule{standard}{popen2}
\modulesynopsis{Subprocesses with accessible standard I/O streams.}
\sectionauthor{Drew Csillag}{drew_csillag@geocities.com}


This module allows you to spawn processes and connect to their
input/output/error pipes and obtain their return codes under
\UNIX{} and Windows.

The \module{subprocess} module provides more powerful facilities for
spawning new processes and retrieving their results.  Using the
\module{subprocess} module is preferable to using the \module{popen2}
module.

The primary interface offered by this module is a trio of factory
functions.  For each of these, if \var{bufsize} is specified, 
it specifies the buffer size for the I/O pipes.  \var{mode}, if
provided, should be the string \code{'b'} or \code{'t'}; on Windows
this is needed to determine whether the file objects should be opened
in binary or text mode.  The default value for \var{mode} is
\code{'t'}.

On \UNIX, \var{cmd} may be a sequence, in which case arguments will be passed
directly to the program without shell intervention (as with
\function{os.spawnv()}). If \var{cmd} is a string it will be passed to the
shell (as with \function{os.system()}).

The only way to retrieve the return codes for the child processes is
by using the \method{poll()} or \method{wait()} methods on the
\class{Popen3} and \class{Popen4} classes; these are only available on
\UNIX.  This information is not available when using the
\function{popen2()}, \function{popen3()}, and \function{popen4()}
functions, or the equivalent functions in the \refmodule{os} module.
(Note that the tuples returned by the \refmodule{os} module's functions
are in a different order from the ones returned by the \module{popen2}
module.)

\begin{funcdesc}{popen2}{cmd\optional{, bufsize\optional{, mode}}}
Executes \var{cmd} as a sub-process.  Returns the file objects
\code{(\var{child_stdout}, \var{child_stdin})}.
\end{funcdesc}

\begin{funcdesc}{popen3}{cmd\optional{, bufsize\optional{, mode}}}
Executes \var{cmd} as a sub-process.  Returns the file objects
\code{(\var{child_stdout}, \var{child_stdin}, \var{child_stderr})}.
\end{funcdesc}

\begin{funcdesc}{popen4}{cmd\optional{, bufsize\optional{, mode}}}
Executes \var{cmd} as a sub-process.  Returns the file objects
\code{(\var{child_stdout_and_stderr}, \var{child_stdin})}.
\versionadded{2.0}
\end{funcdesc}


On \UNIX, a class defining the objects returned by the factory
functions is also available.  These are not used for the Windows
implementation, and are not available on that platform.

\begin{classdesc}{Popen3}{cmd\optional{, capturestderr\optional{, bufsize}}}
This class represents a child process.  Normally, \class{Popen3}
instances are created using the \function{popen2()} and
\function{popen3()} factory functions described above.

If not using one of the helper functions to create \class{Popen3}
objects, the parameter \var{cmd} is the shell command to execute in a
sub-process.  The \var{capturestderr} flag, if true, specifies that
the object should capture standard error output of the child process.
The default is false.  If the \var{bufsize} parameter is specified, it
specifies the size of the I/O buffers to/from the child process.
\end{classdesc}

\begin{classdesc}{Popen4}{cmd\optional{, bufsize}}
Similar to \class{Popen3}, but always captures standard error into the
same file object as standard output.  These are typically created
using \function{popen4()}.
\versionadded{2.0}
\end{classdesc}

\subsection{Popen3 and Popen4 Objects \label{popen3-objects}}

Instances of the \class{Popen3} and \class{Popen4} classes have the
following methods:

\begin{methoddesc}{poll}{}
Returns \code{-1} if child process hasn't completed yet, or its status 
code (see \method{wait()}) otherwise.
\end{methoddesc}

\begin{methoddesc}{wait}{}
Waits for and returns the status code of the child process.  The
status code encodes both the return code of the process and
information about whether it exited using the \cfunction{exit()}
system call or died due to a signal.  Functions to help interpret the
status code are defined in the \refmodule{os} module; see section
\ref{os-process} for the \function{W\var{*}()} family of functions.
\end{methoddesc}


The following attributes are also available: 

\begin{memberdesc}{fromchild}
A file object that provides output from the child process.  For
\class{Popen4} instances, this will provide both the standard output
and standard error streams.
\end{memberdesc}

\begin{memberdesc}{tochild}
A file object that provides input to the child process.
\end{memberdesc}

\begin{memberdesc}{childerr}
A file object that provides error output from the child process, if
\var{capturestderr} was true for the constructor, otherwise
\code{None}.  This will always be \code{None} for \class{Popen4}
instances.
\end{memberdesc}

\begin{memberdesc}{pid}
The process ID of the child process.
\end{memberdesc}


\subsection{Flow Control Issues \label{popen2-flow-control}}

Any time you are working with any form of inter-process communication,
control flow needs to be carefully thought out.  This remains the case
with the file objects provided by this module (or the \refmodule{os}
module equivalents).

% Example explanation and suggested work-arounds substantially stolen
% from Martin von Löwis:
% http://mail.python.org/pipermail/python-dev/2000-September/009460.html

When reading output from a child process that writes a lot of data to
standard error while the parent is reading from the child's standard
output, a deadlock can occur.  A similar situation can occur with other
combinations of reads and writes.  The essential factors are that more
than \constant{_PC_PIPE_BUF} bytes are being written by one process in
a blocking fashion, while the other process is reading from the other
process, also in a blocking fashion.

There are several ways to deal with this situation.

The simplest application change, in many cases, will be to follow this
model in the parent process:

\begin{verbatim}
import popen2

r, w, e = popen2.popen3('python slave.py')
e.readlines()
r.readlines()
r.close()
e.close()
w.close()
\end{verbatim}

with code like this in the child:

\begin{verbatim}
import os
import sys

# note that each of these print statements
# writes a single long string

print >>sys.stderr, 400 * 'this is a test\n'
os.close(sys.stderr.fileno())
print >>sys.stdout, 400 * 'this is another test\n'
\end{verbatim}

In particular, note that \code{sys.stderr} must be closed after
writing all data, or \method{readlines()} won't return.  Also note
that \function{os.close()} must be used, as \code{sys.stderr.close()}
won't close \code{stderr} (otherwise assigning to \code{sys.stderr}
will silently close it, so no further errors can be printed).

Applications which need to support a more general approach should
integrate I/O over pipes with their \function{select()} loops, or use
separate threads to read each of the individual files provided by
whichever \function{popen*()} function or \class{Popen*} class was
used.

\begin{seealso}
  \seemodule{subprocess}{Module for spawning and managing subprocesses.}
\end{seealso}
