\section{\module{sys} ---
         System-specific parameters and functions}

\declaremodule{builtin}{sys}
\modulesynopsis{Access system-specific parameters and functions.}

This module provides access to some variables used or maintained by the
interpreter and to functions that interact strongly with the interpreter.
It is always available.


\begin{datadesc}{argv}
  The list of command line arguments passed to a Python script.
  \code{argv[0]} is the script name (it is operating system
  dependent whether this is a full pathname or not).
  If the command was executed using the \programopt{-c} command line
  option to the interpreter, \code{argv[0]} is set to the string
  \code{'-c'}.
  If no script name was passed to the Python interpreter,
  \code{argv} has zero length.
\end{datadesc}

\begin{datadesc}{byteorder}
  An indicator of the native byte order.  This will have the value
  \code{'big'} on big-endian (most-signigicant byte first) platforms,
  and \code{'little'} on little-endian (least-significant byte first)
  platforms.
  \versionadded{2.0}
\end{datadesc}

\begin{datadesc}{builtin_module_names}
  A tuple of strings giving the names of all modules that are compiled
  into this Python interpreter.  (This information is not available in
  any other way --- \code{modules.keys()} only lists the imported
  modules.)
\end{datadesc}

\begin{datadesc}{copyright}
A string containing the copyright pertaining to the Python interpreter.
\end{datadesc}

\begin{datadesc}{dllhandle}
Integer specifying the handle of the Python DLL.
Availability: Windows.
\end{datadesc}

\begin{funcdesc}{exc_info}{}
This function returns a tuple of three values that give information
about the exception that is currently being handled.  The information
returned is specific both to the current thread and to the current
stack frame.  If the current stack frame is not handling an exception,
the information is taken from the calling stack frame, or its caller,
and so on until a stack frame is found that is handling an exception.
Here, ``handling an exception'' is defined as ``executing or having
executed an except clause.''  For any stack frame, only
information about the most recently handled exception is accessible.

If no exception is being handled anywhere on the stack, a tuple
containing three \code{None} values is returned.  Otherwise, the
values returned are
\code{(\var{type}, \var{value}, \var{traceback})}.
Their meaning is: \var{type} gets the exception type of the exception
being handled (a string or class object); \var{value} gets the
exception parameter (its \dfn{associated value} or the second argument
to \keyword{raise}, which is always a class instance if the exception
type is a class object); \var{traceback} gets a traceback object (see
the Reference Manual) which encapsulates the call stack at the point
where the exception originally occurred.
\obindex{traceback}

\strong{Warning:} assigning the \var{traceback} return value to a
local variable in a function that is handling an exception will cause
a circular reference. This will prevent anything referenced by a local
variable in the same function or by the traceback from being garbage
collected.  Since most functions don't need access to the traceback,
the best solution is to use something like
\code{type, value = sys.exc_info()[:2]}
to extract only the exception type and value.  If you do need the
traceback, make sure to delete it after use (best done with a
\keyword{try} ... \keyword{finally} statement) or to call
\function{exc_info()} in a function that does not itself handle an
exception.
\end{funcdesc}

\begin{datadesc}{exc_type}
\dataline{exc_value}
\dataline{exc_traceback}
\deprecated {1.5}
            {Use \function{exc_info()} instead.}
Since they are global variables, they are not specific to the current
thread, so their use is not safe in a multi-threaded program.  When no
exception is being handled, \code{exc_type} is set to \code{None} and
the other two are undefined.
\end{datadesc}

\begin{datadesc}{exec_prefix}
A string giving the site-specific directory prefix where the
platform-dependent Python files are installed; by default, this is
also \code{'/usr/local'}.  This can be set at build time with the
\longprogramopt{exec-prefix} argument to the
\program{configure} script.  Specifically, all configuration files
(e.g. the \file{config.h} header file) are installed in the directory
\code{exec_prefix + '/lib/python\var{version}/config'}, and shared
library modules are installed in \code{exec_prefix +
'/lib/python\var{version}/lib-dynload'}, where \var{version} is equal
to \code{version[:3]}.
\end{datadesc}

\begin{datadesc}{executable}
A string giving the name of the executable binary for the Python
interpreter, on systems where this makes sense.
\end{datadesc}

\begin{funcdesc}{exit}{\optional{arg}}
Exit from Python.  This is implemented by raising the
\exception{SystemExit} exception, so cleanup actions specified by
finally clauses of \keyword{try} statements are honored, and it is
possible to intercept the exit attempt at an outer level.  The
optional argument \var{arg} can be an integer giving the exit status
(defaulting to zero), or another type of object.  If it is an integer,
zero is considered ``successful termination'' and any nonzero value is
considered ``abnormal termination'' by shells and the like.  Most
systems require it to be in the range 0-127, and produce undefined
results otherwise.  Some systems have a convention for assigning
specific meanings to specific exit codes, but these are generally
underdeveloped; Unix programs generally use 2 for command line syntax
errors and 1 for all other kind of errors.  If another type of object
is passed, \code{None} is equivalent to passing zero, and any other
object is printed to \code{sys.stderr} and results in an exit code of
1.  In particular, \code{sys.exit("some error message")} is a quick
way to exit a program when an error occurs.
\end{funcdesc}

\begin{datadesc}{exitfunc}
  This value is not actually defined by the module, but can be set by
  the user (or by a program) to specify a clean-up action at program
  exit.  When set, it should be a parameterless function.  This function
  will be called when the interpreter exits.  Only one function may be
  installed in this way; to allow multiple functions which will be called
  at termination, use the \refmodule{atexit} module.  Note: the exit function
  is not called when the program is killed by a signal, when a Python
  fatal internal error is detected, or when \code{os._exit()} is called.
\end{datadesc}

\begin{funcdesc}{getrefcount}{object}
Return the reference count of the \var{object}.  The count returned is
generally one higher than you might expect, because it includes the
(temporary) reference as an argument to \function{getrefcount()}.
\end{funcdesc}

\begin{funcdesc}{getrecursionlimit}{}
Return the current value of the recursion limit, the maximum depth of
the Python interpreter stack.  This limit prevents infinite recursion
from causing an overflow of the C stack and crashing Python.  It can
be set by \function{setrecursionlimit()}.
\end{funcdesc}

\begin{datadesc}{hexversion}
The version number encoded as a single integer.  This is guaranteed to
increase with each version, including proper support for
non-production releases.  For example, to test that the Python
interpreter is at least version 1.5.2, use:

\begin{verbatim}
if sys.hexversion >= 0x010502F0:
    # use some advanced feature
    ...
else:
    # use an alternative implementation or warn the user
    ...
\end{verbatim}

This is called \samp{hexversion} since it only really looks meaningful
when viewed as the result of passing it to the built-in
\function{hex()} function.  The \code{version_info} value may be used
for a more human-friendly encoding of the same information.
\versionadded{1.5.2}
\end{datadesc}

\begin{datadesc}{last_type}
\dataline{last_value}
\dataline{last_traceback}
These three variables are not always defined; they are set when an
exception is not handled and the interpreter prints an error message
and a stack traceback.  Their intended use is to allow an interactive
user to import a debugger module and engage in post-mortem debugging
without having to re-execute the command that caused the error.
(Typical use is \samp{import pdb; pdb.pm()} to enter the post-mortem
debugger; see the chapter ``The Python Debugger'' for more
information.)
\refstmodindex{pdb}

The meaning of the variables is the same
as that of the return values from \function{exc_info()} above.
(Since there is only one interactive thread, thread-safety is not a
concern for these variables, unlike for \code{exc_type} etc.)
\end{datadesc}

\begin{datadesc}{maxint}
The largest positive integer supported by Python's regular integer
type.  This is at least 2**31-1.  The largest negative integer is
\code{-maxint-1} -- the asymmetry results from the use of 2's
complement binary arithmetic.
\end{datadesc}

\begin{datadesc}{modules}
  This is a dictionary that maps module names to modules which have
  already been loaded.  This can be manipulated to force reloading of
  modules and other tricks.  Note that removing a module from this
  dictionary is \emph{not} the same as calling
  \function{reload()}\bifuncindex{reload} on the corresponding module
  object.
\end{datadesc}

\begin{datadesc}{path}
\indexiii{module}{search}{path}
  A list of strings that specifies the search path for modules.
  Initialized from the environment variable \envvar{PYTHONPATH}, or an
  installation-dependent default.  

The first item of this list, \code{path[0]}, is the 
directory containing the script that was used to invoke the Python 
interpreter.  If the script directory is not available (e.g.  if the 
interpreter is invoked interactively or if the script is read from 
standard input), \code{path[0]} is the empty string, which directs 
Python to search modules in the current directory first.  Notice that 
the script directory is inserted \emph{before} the entries inserted as 
a result of \envvar{PYTHONPATH}.  
\end{datadesc}

\begin{datadesc}{platform}
This string contains a platform identifier, e.g. \code{'sunos5'} or
\code{'linux1'}.  This can be used to append platform-specific
components to \code{path}, for instance. 
\end{datadesc}

\begin{datadesc}{prefix}
A string giving the site-specific directory prefix where the platform
independent Python files are installed; by default, this is the string
\code{'/usr/local'}.  This can be set at build time with the
\longprogramopt{prefix} argument to the
\program{configure} script.  The main collection of Python library
modules is installed in the directory \code{prefix +
'/lib/python\var{version}'} while the platform independent header
files (all except \file{config.h}) are stored in \code{prefix +
'/include/python\var{version}'}, where \var{version} is equal to
\code{version[:3]}.
\end{datadesc}

\begin{datadesc}{ps1}
\dataline{ps2}
\index{interpreter prompts}
\index{prompts, interpreter}
  Strings specifying the primary and secondary prompt of the
  interpreter.  These are only defined if the interpreter is in
  interactive mode.  Their initial values in this case are
  \code{'>>> '} and \code{'... '}.  If a non-string object is assigned
  to either variable, its \function{str()} is re-evaluated each time
  the interpreter prepares to read a new interactive command; this can
  be used to implement a dynamic prompt.
\end{datadesc}

\begin{funcdesc}{setcheckinterval}{interval}
Set the interpreter's ``check interval''.  This integer value
determines how often the interpreter checks for periodic things such
as thread switches and signal handlers.  The default is \code{10}, meaning
the check is performed every 10 Python virtual instructions.  Setting
it to a larger value may increase performance for programs using
threads.  Setting it to a value \code{<=} 0 checks every virtual instruction,
maximizing responsiveness as well as overhead.
\end{funcdesc}

\begin{funcdesc}{setprofile}{profilefunc}
  Set the system's profile function, which allows you to implement a
  Python source code profiler in Python.  See the chapter on the
  Python Profiler.  The system's profile function
  is called similarly to the system's trace function (see
  \function{settrace()}), but it isn't called for each executed line of
  code (only on call and return and when an exception occurs).  Also,
  its return value is not used, so it can just return \code{None}.
\end{funcdesc}
\index{profile function}
\index{profiler}

\begin{funcdesc}{setrecursionlimit}{limit}
Set the maximum depth of the Python interpreter stack to \var{limit}.
This limit prevents infinite recursion from causing an overflow of the
C stack and crashing Python.  

The highest possible limit is platform-dependent.  A user may need to
set the limit higher when she has a program that requires deep
recursion and a platform that supports a higher limit.  This should be
done with care, because a too-high limit can lead to a crash.
\end{funcdesc}

\begin{funcdesc}{settrace}{tracefunc}
  Set the system's trace function, which allows you to implement a
  Python source code debugger in Python.  See section ``How It Works''
  in the chapter on the Python Debugger.
\end{funcdesc}
\index{trace function}
\index{debugger}

\begin{datadesc}{stdin}
\dataline{stdout}
\dataline{stderr}
  File objects corresponding to the interpreter's standard input,
  output and error streams.  \code{stdin} is used for all
  interpreter input except for scripts but including calls to
  \function{input()}\bifuncindex{input} and
  \function{raw_input()}\bifuncindex{raw_input}.  \code{stdout} is used
  for the output of \keyword{print} and expression statements and for the
  prompts of \function{input()} and \function{raw_input()}.  The interpreter's
  own prompts and (almost all of) its error messages go to
  \code{stderr}.  \code{stdout} and \code{stderr} needn't
  be built-in file objects: any object is acceptable as long as it has
  a \method{write()} method that takes a string argument.  (Changing these
  objects doesn't affect the standard I/O streams of processes
  executed by \function{os.popen()}, \function{os.system()} or the
  \function{exec*()} family of functions in the \refmodule{os} module.)
\refstmodindex{os}
\end{datadesc}

\begin{datadesc}{__stdin__}
\dataline{__stdout__}
\dataline{__stderr__}
These objects contain the original values of \code{stdin},
\code{stderr} and \code{stdout} at the start of the program.  They are 
used during finalization, and could be useful to restore the actual
files to known working file objects in case they have been overwritten
with a broken object.
\end{datadesc}

\begin{datadesc}{tracebacklimit}
When this variable is set to an integer value, it determines the
maximum number of levels of traceback information printed when an
unhandled exception occurs.  The default is \code{1000}.  When set to
0 or less, all traceback information is suppressed and only the
exception type and value are printed.
\end{datadesc}

\begin{datadesc}{version}
A string containing the version number of the Python interpreter plus
additional information on the build number and compiler used.  It has
a value of the form \code{'\var{version} (\#\var{build_number},
\var{build_date}, \var{build_time}) [\var{compiler}]'}.  The first
three characters are used to identify the version in the installation
directories (where appropriate on each platform).  An example:

\begin{verbatim}
>>> import sys
>>> sys.version
'1.5.2 (#0 Apr 13 1999, 10:51:12) [MSC 32 bit (Intel)]'
\end{verbatim}
\end{datadesc}

\begin{datadesc}{version_info}
A tuple containing the five components of the version number:
\var{major}, \var{minor}, \var{micro}, \var{releaselevel}, and
\var{serial}.  All values except \var{releaselevel} are integers; the
release level is \code{'alpha'}, \code{'beta'},
\code{'candidate'}, or \code{'final'}.  The \code{version_info} value
corresponding to the Python version 2.0 is
\code{(2, 0, 0, 'final', 0)}.
\versionadded{2.0}
\end{datadesc}

\begin{datadesc}{winver}
The version number used to form registry keys on Windows platforms.
This is stored as string resource 1000 in the Python DLL.  The value
is normally the first three characters of \constant{version}.  It is
provided in the \module{sys} module for informational purposes;
modifying this value has no effect on the registry keys used by
Python.
Availability: Windows.
\end{datadesc}
