\section{Built-in Module \sectcode{termios}}
\label{module-termios}
\bimodindex{termios}
\indexii{Posix}{I/O control}
\indexii{tty}{I/O control}

\renewcommand{\indexsubitem}{(in module termios)}

This module provides an interface to the Posix calls for tty I/O
control.  For a complete description of these calls, see the Posix or
\UNIX{} manual pages.  It is only available for those \UNIX{} versions
that support Posix \code{termios} style tty I/O control (and then
only if configured at installation time).

All functions in this module take a file descriptor \var{fd} as their
first argument.  This must be an integer file descriptor, such as
returned by \code{sys.stdin.fileno()}.

This module should be used in conjunction with the \code{TERMIOS}
module, which defines the relevant symbolic constants (see the next
section).

The module defines the following functions:

\begin{funcdesc}{tcgetattr}{fd}
Return a list containing the tty attributes for file descriptor
\var{fd}, as follows: \code{[\var{iflag}, \var{oflag}, \var{cflag},
\var{lflag}, \var{ispeed}, \var{ospeed}, \var{cc}]} where \var{cc} is
a list of the tty special characters (each a string of length 1,
except the items with indices \code{VMIN} and \code{VTIME}, which are
integers when these fields are defined).  The interpretation of the
flags and the speeds as well as the indexing in the \var{cc} array
must be done using the symbolic constants defined in the
\code{TERMIOS} module.
\end{funcdesc}

\begin{funcdesc}{tcsetattr}{fd\, when\, attributes}
Set the tty attributes for file descriptor \var{fd} from the
\var{attributes}, which is a list like the one returned by
\code{tcgetattr()}.  The \var{when} argument determines when the
attributes are changed: \code{TERMIOS.TCSANOW} to change immediately,
\code{TERMIOS.TCSADRAIN} to change after transmitting all queued
output, or \code{TERMIOS.TCSAFLUSH} to change after transmitting all
queued output and discarding all queued input.
\end{funcdesc}

\begin{funcdesc}{tcsendbreak}{fd\, duration}
Send a break on file descriptor \var{fd}.  A zero \var{duration} sends
a break for 0.25--0.5 seconds; a nonzero \var{duration} has a system
dependent meaning.
\end{funcdesc}

\begin{funcdesc}{tcdrain}{fd}
Wait until all output written to file descriptor \var{fd} has been
transmitted.
\end{funcdesc}

\begin{funcdesc}{tcflush}{fd\, queue}
Discard queued data on file descriptor \var{fd}.  The \var{queue}
selector specifies which queue: \code{TERMIOS.TCIFLUSH} for the input
queue, \code{TERMIOS.TCOFLUSH} for the output queue, or
\code{TERMIOS.TCIOFLUSH} for both queues.
\end{funcdesc}

\begin{funcdesc}{tcflow}{fd\, action}
Suspend or resume input or output on file descriptor \var{fd}.  The
\var{action} argument can be \code{TERMIOS.TCOOFF} to suspend output,
\code{TERMIOS.TCOON} to restart output, \code{TERMIOS.TCIOFF} to
suspend input, or \code{TERMIOS.TCION} to restart input.
\end{funcdesc}

\subsection{Example}
\nodename{termios Example}

Here's a function that prompts for a password with echoing turned off.
Note the technique using a separate \code{termios.tcgetattr()} call
and a \code{try \ldots{} finally} statement to ensure that the old tty
attributes are restored exactly no matter what happens:

\bcode\begin{verbatim}
def getpass(prompt = "Password: "):
    import termios, TERMIOS, sys
    fd = sys.stdin.fileno()
    old = termios.tcgetattr(fd)
    new = termios.tcgetattr(fd)
    new[3] = new[3] & ~TERMIOS.ECHO          # lflags
    try:
        termios.tcsetattr(fd, TERMIOS.TCSADRAIN, new)
        passwd = raw_input(prompt)
    finally:
        termios.tcsetattr(fd, TERMIOS.TCSADRAIN, old)
    return passwd
\end{verbatim}\ecode
%
\section{Standard Module \sectcode{TERMIOS}}
\stmodindex{TERMIOS}
\indexii{Posix}{I/O control}
\indexii{tty}{I/O control}

\renewcommand{\indexsubitem}{(in module TERMIOS)}

This module defines the symbolic constants required to use the
\code{termios} module (see the previous section).  See the Posix or
\UNIX{} manual pages (or the source) for a list of those constants.
\refbimodindex{termios}

Note: this module resides in a system-dependent subdirectory of the
Python library directory.  You may have to generate it for your
particular system using the script \file{Tools/scripts/h2py.py}.
