\section{\module{test} ---
         Regression tests package for Python}

\declaremodule{standard}{test}
\sectionauthor{Brett Cannon}{brett@python.org}
\modulesynopsis{Regression tests package containing the testing suite
                for Python.}


The \module{test} package contains all regression tests for Python as
well as the modules \module{test.test_support} and
\module{test.regrtest}.  \module{test.test_support} is used to enhance
your tests while \module{test.regrtest} drives the testing suite.

Each module in the \module{test} package whose name starts with
\samp{test_} is a testing suite for a specific module or feature.
All new tests should be written using the \refmodule{unittest} or
\refmodule{doctest} module.  Some older tests are
written using a ``traditional'' testing style that compares output
printed to \code{sys.stdout}; this style of test is considered
deprecated.

\begin{seealso}
\seemodule{unittest}{Writing PyUnit regression tests.}
\seemodule{doctest}{Tests embedded in documentation strings.}
\end{seealso}


\subsection{Writing Unit Tests for the \module{test} package%
            \label{writing-tests}}

It is preferred that tests that use the \refmodule{unittest} module
follow a few guidelines.
One is to name the test module by starting it with \samp{test_} and end it with
the name of the module being tested.
The test methods in the test module should start with \samp{test_} and end with
a description of what the method is testing.
This is needed so that the methods are recognized by the test driver as
test methods.
Also, no documentation string for the method should be included.
A comment (such as
\samp{\# Tests function returns only True or False}) should be used to provide
documentation for test methods.
This is done because documentation strings get printed out if they exist and
thus what test is being run is not stated.

A basic boilerplate is often used:

\begin{verbatim}
import unittest
from test import test_support

class MyTestCase1(unittest.TestCase):

    # Only use setUp() and tearDown() if necessary

    def setUp(self):
        ... code to execute in preparation for tests ...

    def tearDown(self):
        ... code to execute to clean up after tests ...

    def test_feature_one(self):
        # Test feature one.
        ... testing code ...

    def test_feature_two(self):
        # Test feature two.
        ... testing code ...

    ... more test methods ...

class MyTestCase2(unittest.TestCase):
    ... same structure as MyTestCase1 ...

... more test classes ...

def test_main():
    test_support.run_unittest(MyTestCase1,
                              MyTestCase2,
                              ... list other tests ...
                             )

if __name__ == '__main__':
    test_main()
\end{verbatim}

This boilerplate code allows the testing suite to be run by
\module{test.regrtest} as well as on its own as a script.

The goal for regression testing is to try to break code.
This leads to a few guidelines to be followed:

\begin{itemize}
\item The testing suite should exercise all classes, functions, and
      constants.
      This includes not just the external API that is to be presented to the
      outside world but also "private" code.
\item Whitebox testing (examining the code being tested when the tests are
      being written) is preferred.
      Blackbox testing (testing only the published user interface) is not
      complete enough to make sure all boundary and edge cases are tested.
\item Make sure all possible values are tested including invalid ones.
      This makes sure that not only all valid values are acceptable but also
      that improper values are handled correctly.
\item Exhaust as many code paths as possible.
      Test where branching occurs and thus tailor input to make sure as many
      different paths through the code are taken.
\item Add an explicit test for any bugs discovered for the tested code.
      This will make sure that the error does not crop up again if the code is
      changed in the future.
\item Make sure to clean up after your tests (such as close and remove all
      temporary files).
\item If a test is dependent on a specific condition of the operating system
      then verify the condition already exists before attempting the test.
\item Import as few modules as possible and do it as soon as possible.
      This minimizes external dependencies of tests and also minimizes possible
      anomalous behavior from side-effects of importing a module.
\item Try to maximize code reuse.
      On occasion, tests will vary by something as small as what type
      of input is used.
      Minimize code duplication by subclassing a basic test class with a class
      that specifies the input:
\begin{verbatim}
class TestFuncAcceptsSequences(unittest.TestCase):

    func = mySuperWhammyFunction

    def test_func(self):
        self.func(self.arg)

class AcceptLists(TestFuncAcceptsSequences):
    arg = [1,2,3]

class AcceptStrings(TestFuncAcceptsSequences):
    arg = 'abc'

class AcceptTuples(TestFuncAcceptsSequences):
    arg = (1,2,3)
\end{verbatim}
\end{itemize}

\begin{seealso}
\seetitle{Test Driven Development}
         {A book by Kent Beck on writing tests before code.}
\end{seealso}


\subsection{Running tests using \module{test.regrtest} \label{regrtest}}

\module{test.regrtest} can be used as a script to drive Python's
regression test suite.
Running the script by itself automatically starts running all
regression tests in the \module{test} package.
It does this by finding all modules in the package whose name starts with
\samp{test_}, importing them, and executing the function
\function{test_main()} if present.
The names of tests to execute may also be passed to the script.
Specifying a single regression test (\program{python regrtest.py}
\programopt{test_spam.py}) will minimize output and only print whether
the test passed or failed and thus minimize output.

Running \module{test.regrtest} directly allows what resources are
available for tests to use to be set.
You do this by using the \programopt{-u} command-line option.
Run \program{python regrtest.py} \programopt{-uall} to turn on all
resources; specifying \programopt{all} as an option for
\programopt{-u} enables all possible resources.
If all but one resource is desired (a more common case), a
comma-separated list of resources that are not desired may be listed after
\programopt{all}.
The command \program{python regrtest.py}
\programopt{-uall,-audio,-largefile} will run \module{test.regrtest}
with all resources except the \programopt{audio} and
\programopt{largefile} resources.
For a list of all resources and more command-line options, run
\program{python regrtest.py} \programopt{-h}.

Some other ways to execute the regression tests depend on what platform the
tests are being executed on.
On \UNIX{}, you can run \program{make} \programopt{test} at the
top-level directory where Python was built.
On Windows, executing \program{rt.bat} from your \file{PCBuild}
directory will run all regression tests.


\section{\module{test.test_support} ---
         Utility functions for tests}

\declaremodule[test.testsupport]{standard}{test.test_support}
\modulesynopsis{Support for Python regression tests.}

The \module{test.test_support} module provides support for Python's
regression tests.

This module defines the following exceptions:

\begin{excdesc}{TestFailed}
Exception to be raised when a test fails. This is deprecated in favor
of \module{unittest}-based tests and \class{unittest.TestCase}'s
assertion methods.
\end{excdesc}

\begin{excdesc}{TestSkipped}
Subclass of \exception{TestFailed}.
Raised when a test is skipped.
This occurs when a needed resource (such as a network connection) is not
available at the time of testing.
\end{excdesc}

\begin{excdesc}{ResourceDenied}
Subclass of \exception{TestSkipped}.
Raised when a resource (such as a network connection) is not available.
Raised by the \function{requires()} function.
\end{excdesc}


The \module{test.test_support} module defines the following constants:

\begin{datadesc}{verbose}
\constant{True} when verbose output is enabled.
Should be checked when more detailed information is desired about a running
test.
\var{verbose} is set by \module{test.regrtest}.
\end{datadesc}

\begin{datadesc}{have_unicode}
\constant{True} when Unicode support is available.
\end{datadesc}

\begin{datadesc}{is_jython}
\constant{True} if the running interpreter is Jython.
\end{datadesc}

\begin{datadesc}{TESTFN}
Set to the path that a temporary file may be created at.
Any temporary that is created should be closed and unlinked (removed).
\end{datadesc}


The \module{test.test_support} module defines the following functions:

\begin{funcdesc}{forget}{module_name}
Removes the module named \var{module_name} from \code{sys.modules} and deletes
any byte-compiled files of the module.
\end{funcdesc}

\begin{funcdesc}{is_resource_enabled}{resource}
Returns \constant{True} if \var{resource} is enabled and available.
The list of available resources is only set when \module{test.regrtest}
is executing the tests.
\end{funcdesc}

\begin{funcdesc}{requires}{resource\optional{, msg}}
Raises \exception{ResourceDenied} if \var{resource} is not available.
\var{msg} is the argument to \exception{ResourceDenied} if it is raised.
Always returns true if called by a function whose \code{__name__} is
\code{'__main__'}.
Used when tests are executed by \module{test.regrtest}.
\end{funcdesc}

\begin{funcdesc}{findfile}{filename}
Return the path to the file named \var{filename}.
If no match is found \var{filename} is returned.
This does not equal a failure since it could be the path to the file.
\end{funcdesc}

\begin{funcdesc}{guard_warnings_filter}{}
Returns a context manager that guards the \module{warnings} module's
filter settings.
\versionadded{2.6}
\end{funcdesc}

\begin{funcdesc}{run_unittest}{*classes}
Execute \class{unittest.TestCase} subclasses passed to the function.
The function scans the classes for methods starting with the prefix
\samp{test_} and executes the tests individually.

It is also legal to pass strings as parameters; these should be keys in
\code{sys.modules}. Each associated module will be scanned by
\code{unittest.TestLoader.loadTestsFromModule()}. This is usually seen in
the following \function{test_main()} function:

\begin{verbatim}
def test_main():
    test_support.run_unittest(__name__)
\end{verbatim}

This will run all tests defined in the named module.
\end{funcdesc}

The \module{test.test_support} module defines the following classes:

\begin{classdesc}{TransientResource}{exc\optional{, **kwargs}}
Instances are a  context manager that raises \class{ResourceDenied} if the
specified exception type is raised.  Any keyword arguments are treated as
attribute/value pairs to be compared against any exception raised within the
\code{with} statement.  Only if all pairs match properly against attributes on
the exception is \class{ResourceDenied} raised.
\versionadded{2.6}
\end{classdesc}

\begin{classdesc}{EnvironmentVarGuard}{}
Class used to temporarily set or unset environment variables.  Instances can be
used as a context manager.
\versionadded{2.6}
\end{classdesc}

\begin{methoddesc}{set}{envvar, value}
Temporarily set the environment variable \code{envvar} to the value of
\code{value}.
\end{methoddesc}

\begin{methoddesc}{unset}{envvar}
Temporarily unset the environment variable \code{envvar}.
\end{methoddesc}

