\section{\module{unicodedata} ---
         Unicode Database}

\declaremodule{standard}{unicodedata}
\modulesynopsis{Access the Unicode Database.}
\moduleauthor{Marc-Andre Lemburg}{mal@lemburg.com}
\sectionauthor{Marc-Andre Lemburg}{mal@lemburg.com}
\sectionauthor{Martin v. L\"owis}{martin@v.loewis.de}

\index{Unicode}
\index{character}
\indexii{Unicode}{database}

This module provides access to the Unicode Character Database which
defines character properties for all Unicode characters. The data in
this database is based on the \file{UnicodeData.txt} file version
3.2.0 which is publically available from \url{ftp://ftp.unicode.org/}.

The module uses the same names and symbols as defined by the
UnicodeData File Format 3.2.0 (see
\url{http://www.unicode.org/Public/UNIDATA/UnicodeData.html}).  It
defines the following functions:

\begin{funcdesc}{lookup}{name}
  Look up character by name.  If a character with the
  given name is found, return the corresponding Unicode
  character.  If not found, \exception{KeyError} is raised.
\end{funcdesc}

\begin{funcdesc}{name}{unichr\optional{, default}}
  Returns the name assigned to the Unicode character
  \var{unichr} as a string. If no name is defined,
  \var{default} is returned, or, if not given,
  \exception{ValueError} is raised.
\end{funcdesc}

\begin{funcdesc}{decimal}{unichr\optional{, default}}
  Returns the decimal value assigned to the Unicode character
  \var{unichr} as integer. If no such value is defined,
  \var{default} is returned, or, if not given,
  \exception{ValueError} is raised.
\end{funcdesc}

\begin{funcdesc}{digit}{unichr\optional{, default}}
  Returns the digit value assigned to the Unicode character
  \var{unichr} as integer. If no such value is defined,
  \var{default} is returned, or, if not given,
  \exception{ValueError} is raised.
\end{funcdesc}

\begin{funcdesc}{numeric}{unichr\optional{, default}}
  Returns the numeric value assigned to the Unicode character
  \var{unichr} as float. If no such value is defined, \var{default} is
  returned, or, if not given, \exception{ValueError} is raised.
\end{funcdesc}

\begin{funcdesc}{category}{unichr}
  Returns the general category assigned to the Unicode character
  \var{unichr} as string.
\end{funcdesc}

\begin{funcdesc}{bidirectional}{unichr}
  Returns the bidirectional category assigned to the Unicode character
  \var{unichr} as string. If no such value is defined, an empty string
  is returned.
\end{funcdesc}

\begin{funcdesc}{combining}{unichr}
  Returns the canonical combining class assigned to the Unicode
  character \var{unichr} as integer. Returns \code{0} if no combining
  class is defined.
\end{funcdesc}

\begin{funcdesc}{east_asian_width}{unichr}
  Returns the east asian width assigned to the Unicode character
  \var{unichr} as string.
\versionadded{2.4}
\end{funcdesc}

\begin{funcdesc}{mirrored}{unichr}
  Returns the mirrored property assigned to the Unicode character
  \var{unichr} as integer. Returns \code{1} if the character has been
  identified as a ``mirrored'' character in bidirectional text,
  \code{0} otherwise.
\end{funcdesc}

\begin{funcdesc}{decomposition}{unichr}
  Returns the character decomposition mapping assigned to the Unicode
  character \var{unichr} as string. An empty string is returned in case
  no such mapping is defined.
\end{funcdesc}

\begin{funcdesc}{normalize}{form, unistr}

Return the normal form \var{form} for the Unicode string \var{unistr}.
Valid values for \var{form} are 'NFC', 'NFKC', 'NFD', and 'NFKD'.

The Unicode standard defines various normalization forms of a Unicode
string, based on the definition of canonical equivalence and
compatibility equivalence. In Unicode, several characters can be
expressed in various way. For example, the character U+00C7 (LATIN
CAPITAL LETTER C WITH CEDILLA) can also be expressed as the sequence
U+0043 (LATIN CAPITAL LETTER C) U+0327 (COMBINING CEDILLA).

For each character, there are two normal forms: normal form C and
normal form D. Normal form D (NFD) is also known as canonical
decomposition, and translates each character into its decomposed form.
Normal form C (NFC) first applies a canonical decomposition, then
composes pre-combined characters again.

In addition to these two forms, there two additional normal forms
based on compatibility equivalence. In Unicode, certain characters are
supported which normally would be unified with other characters. For
example, U+2160 (ROMAN NUMERAL ONE) is really the same thing as U+0049
(LATIN CAPITAL LETTER I). However, it is supported in Unicode for
compatibility with existing character sets (e.g. gb2312).

The normal form KD (NFKD) will apply the compatibility decomposition,
i.e. replace all compatibility characters with their equivalents. The
normal form KC (NFKC) first applies the compatibility decomposition,
followed by the canonical composition.

\versionadded{2.3}
\end{funcdesc}

In addition, the module exposes the following constant:

\begin{datadesc}{unidata_version}
The version of the Unicode database used in this module.

\versionadded{2.3}
\end{datadesc}
