\chapter{Abstract Objects Layer \label{abstract}}

The functions in this chapter interact with Python objects regardless
of their type, or with wide classes of object types (e.g. all
numerical types, or all sequence types).  When used on object types
for which they do not apply, they will raise a Python exception.


\section{Object Protocol \label{object}}

\begin{cfuncdesc}{int}{PyObject_Print}{PyObject *o, FILE *fp, int flags}
  Print an object \var{o}, on file \var{fp}.  Returns \code{-1} on
  error.  The flags argument is used to enable certain printing
  options.  The only option currently supported is
  \constant{Py_PRINT_RAW}; if given, the \function{str()} of the
  object is written instead of the \function{repr()}.
\end{cfuncdesc}

\begin{cfuncdesc}{int}{PyObject_HasAttrString}{PyObject *o, char *attr_name}
  Returns \code{1} if \var{o} has the attribute \var{attr_name}, and
  \code{0} otherwise.  This is equivalent to the Python expression
  \samp{hasattr(\var{o}, \var{attr_name})}.  This function always
  succeeds.
\end{cfuncdesc}

\begin{cfuncdesc}{PyObject*}{PyObject_GetAttrString}{PyObject *o,
                                                     char *attr_name}
  Retrieve an attribute named \var{attr_name} from object \var{o}.
  Returns the attribute value on success, or \NULL{} on failure.
  This is the equivalent of the Python expression
  \samp{\var{o}.\var{attr_name}}.
\end{cfuncdesc}


\begin{cfuncdesc}{int}{PyObject_HasAttr}{PyObject *o, PyObject *attr_name}
  Returns \code{1} if \var{o} has the attribute \var{attr_name}, and
  \code{0} otherwise.  This is equivalent to the Python expression
  \samp{hasattr(\var{o}, \var{attr_name})}.  This function always
  succeeds.
\end{cfuncdesc}


\begin{cfuncdesc}{PyObject*}{PyObject_GetAttr}{PyObject *o,
                                               PyObject *attr_name}
  Retrieve an attribute named \var{attr_name} from object \var{o}.
  Returns the attribute value on success, or \NULL{} on failure.  This
  is the equivalent of the Python expression
  \samp{\var{o}.\var{attr_name}}.
\end{cfuncdesc}


\begin{cfuncdesc}{int}{PyObject_SetAttrString}{PyObject *o,
                                               char *attr_name, PyObject *v}
  Set the value of the attribute named \var{attr_name}, for object
  \var{o}, to the value \var{v}. Returns \code{-1} on failure.  This
  is the equivalent of the Python statement
  \samp{\var{o}.\var{attr_name} = \var{v}}.
\end{cfuncdesc}


\begin{cfuncdesc}{int}{PyObject_SetAttr}{PyObject *o,
                                         PyObject *attr_name, PyObject *v}
  Set the value of the attribute named \var{attr_name}, for object
  \var{o}, to the value \var{v}. Returns \code{-1} on failure.  This
  is the equivalent of the Python statement
  \samp{\var{o}.\var{attr_name} = \var{v}}.
\end{cfuncdesc}


\begin{cfuncdesc}{int}{PyObject_DelAttrString}{PyObject *o, char *attr_name}
  Delete attribute named \var{attr_name}, for object \var{o}. Returns
  \code{-1} on failure.  This is the equivalent of the Python
  statement: \samp{del \var{o}.\var{attr_name}}.
\end{cfuncdesc}


\begin{cfuncdesc}{int}{PyObject_DelAttr}{PyObject *o, PyObject *attr_name}
  Delete attribute named \var{attr_name}, for object \var{o}. Returns
  \code{-1} on failure.  This is the equivalent of the Python
  statement \samp{del \var{o}.\var{attr_name}}.
\end{cfuncdesc}


\begin{cfuncdesc}{PyObject*}{PyObject_RichCompare}{PyObject *o1,
                                                   PyObject *o2, int opid}
  Compare the values of \var{o1} and \var{o2} using the operation
  specified by \var{opid}, which must be one of
  \constant{Py_LT},
  \constant{Py_LE},
  \constant{Py_EQ},
  \constant{Py_NE},
  \constant{Py_GT}, or
  \constant{Py_GE}, corresponding to
  \code{<},
  \code{<=},
  \code{==},
  \code{!=},
  \code{>}, or
  \code{>=} respectively. This is the equivalent of the Python expression
  \samp{\var{o1} op \var{o2}}, where \code{op} is the operator
  corresponding to \var{opid}. Returns the value of the comparison on
  success, or \NULL{} on failure.
\end{cfuncdesc}

\begin{cfuncdesc}{int}{PyObject_RichCompareBool}{PyObject *o1,
                                                 PyObject *o2, int opid}
  Compare the values of \var{o1} and \var{o2} using the operation
  specified by \var{opid}, which must be one of
  \constant{Py_LT},
  \constant{Py_LE},
  \constant{Py_EQ},
  \constant{Py_NE},
  \constant{Py_GT}, or
  \constant{Py_GE}, corresponding to
  \code{<},
  \code{<=},
  \code{==},
  \code{!=},
  \code{>}, or
  \code{>=} respectively. Returns \code{-1} on error, \code{0} if the
  result is false, \code{1} otherwise. This is the equivalent of the
  Python expression \samp{\var{o1} op \var{o2}}, where
  \code{op} is the operator corresponding to \var{opid}.
\end{cfuncdesc}

\begin{cfuncdesc}{int}{PyObject_Cmp}{PyObject *o1, PyObject *o2, int *result}
  Compare the values of \var{o1} and \var{o2} using a routine provided
  by \var{o1}, if one exists, otherwise with a routine provided by
  \var{o2}.  The result of the comparison is returned in
  \var{result}.  Returns \code{-1} on failure.  This is the equivalent
  of the Python statement\bifuncindex{cmp} \samp{\var{result} =
  cmp(\var{o1}, \var{o2})}.
\end{cfuncdesc}


\begin{cfuncdesc}{int}{PyObject_Compare}{PyObject *o1, PyObject *o2}
  Compare the values of \var{o1} and \var{o2} using a routine provided
  by \var{o1}, if one exists, otherwise with a routine provided by
  \var{o2}.  Returns the result of the comparison on success.  On
  error, the value returned is undefined; use
  \cfunction{PyErr_Occurred()} to detect an error.  This is equivalent
  to the Python expression\bifuncindex{cmp} \samp{cmp(\var{o1},
  \var{o2})}.
\end{cfuncdesc}


\begin{cfuncdesc}{PyObject*}{PyObject_Repr}{PyObject *o}
  Compute a string representation of object \var{o}.  Returns the
  string representation on success, \NULL{} on failure.  This is the
  equivalent of the Python expression \samp{repr(\var{o})}.  Called by
  the \function{repr()}\bifuncindex{repr} built-in function and by
  reverse quotes.
\end{cfuncdesc}


\begin{cfuncdesc}{PyObject*}{PyObject_Str}{PyObject *o}
  Compute a string representation of object \var{o}.  Returns the
  string representation on success, \NULL{} on failure.  This is the
  equivalent of the Python expression \samp{str(\var{o})}.  Called by
  the \function{str()}\bifuncindex{str} built-in function and by the
  \keyword{print} statement.
\end{cfuncdesc}


\begin{cfuncdesc}{PyObject*}{PyObject_Unicode}{PyObject *o}
  Compute a Unicode string representation of object \var{o}.  Returns
  the Unicode string representation on success, \NULL{} on failure.
  This is the equivalent of the Python expression
  \samp{unicode(\var{o})}.  Called by the
  \function{unicode()}\bifuncindex{unicode} built-in function.
\end{cfuncdesc}

\begin{cfuncdesc}{int}{PyObject_IsInstance}{PyObject *inst, PyObject *cls}
  Returns \code{1} if \var{inst} is an instance of the class \var{cls}
  or a subclass of \var{cls}, or \code{0} if not.  On error, returns
  \code{-1} and sets an exception.  If \var{cls} is a type object
  rather than a class object, \cfunction{PyObject_IsInstance()}
  returns \code{1} if \var{inst} is of type \var{cls}.  If \var{cls}
  is a tuple, the check will be done against every entry in \var{cls}.
  The result will be \code{1} when at least one of the checks returns
  \code{1}, otherwise it will be \code{0}. If \var{inst} is not a class
  instance and \var{cls} is neither a type object, nor a class object,
  nor a tuple, \var{inst} must have a \member{__class__} attribute
  --- the class relationship of the value of that attribute with
  \var{cls} will be used to determine the result of this function.
  \versionadded{2.1}
  \versionchanged[Support for a tuple as the second argument added]{2.2}
\end{cfuncdesc}

Subclass determination is done in a fairly straightforward way, but
includes a wrinkle that implementors of extensions to the class system
may want to be aware of.  If \class{A} and \class{B} are class
objects, \class{B} is a subclass of \class{A} if it inherits from
\class{A} either directly or indirectly.  If either is not a class
object, a more general mechanism is used to determine the class
relationship of the two objects.  When testing if \var{B} is a
subclass of \var{A}, if \var{A} is \var{B},
\cfunction{PyObject_IsSubclass()} returns true.  If \var{A} and
\var{B} are different objects, \var{B}'s \member{__bases__} attribute
is searched in a depth-first fashion for \var{A} --- the presence of
the \member{__bases__} attribute is considered sufficient for this
determination.

\begin{cfuncdesc}{int}{PyObject_IsSubclass}{PyObject *derived,
                                            PyObject *cls}
  Returns \code{1} if the class \var{derived} is identical to or
  derived from the class \var{cls}, otherwise returns \code{0}.  In
  case of an error, returns \code{-1}. If \var{cls}
  is a tuple, the check will be done against every entry in \var{cls}.
  The result will be \code{1} when at least one of the checks returns
  \code{1}, otherwise it will be \code{0}. If either \var{derived} or
  \var{cls} is not an actual class object (or tuple), this function
  uses the generic algorithm described above.
  \versionadded{2.1}
  \versionchanged[Older versions of Python did not support a tuple
                  as the second argument]{2.3}
\end{cfuncdesc}


\begin{cfuncdesc}{int}{PyCallable_Check}{PyObject *o}
  Determine if the object \var{o} is callable.  Return \code{1} if the
  object is callable and \code{0} otherwise.  This function always
  succeeds.
\end{cfuncdesc}


\begin{cfuncdesc}{PyObject*}{PyObject_Call}{PyObject *callable_object,
                                            PyObject *args,
                                            PyObject *kw}
  Call a callable Python object \var{callable_object}, with arguments
  given by the tuple \var{args}, and named arguments given by the
  dictionary \var{kw}. If no named arguments are needed, \var{kw} may
  be \NULL{}. \var{args} must not be \NULL{}, use an empty tuple if
  no arguments are needed. Returns the result of the call on success,
  or \NULL{} on failure.  This is the equivalent of the Python
  expression \samp{apply(\var{callable_object}, \var{args}, \var{kw})}
  or \samp{\var{callable_object}(*\var{args}, **\var{kw})}.
  \bifuncindex{apply}
  \versionadded{2.2}
\end{cfuncdesc}


\begin{cfuncdesc}{PyObject*}{PyObject_CallObject}{PyObject *callable_object,
                                                  PyObject *args}
  Call a callable Python object \var{callable_object}, with arguments
  given by the tuple \var{args}.  If no arguments are needed, then
  \var{args} may be \NULL.  Returns the result of the call on
  success, or \NULL{} on failure.  This is the equivalent of the
  Python expression \samp{apply(\var{callable_object}, \var{args})} or
  \samp{\var{callable_object}(*\var{args})}.
  \bifuncindex{apply}
\end{cfuncdesc}

\begin{cfuncdesc}{PyObject*}{PyObject_CallFunction}{PyObject *callable,
                                                    char *format, \moreargs}
  Call a callable Python object \var{callable}, with a variable
  number of C arguments.  The C arguments are described using a
  \cfunction{Py_BuildValue()} style format string.  The format may be
  \NULL, indicating that no arguments are provided.  Returns the
  result of the call on success, or \NULL{} on failure.  This is the
  equivalent of the Python expression \samp{apply(\var{callable},
  \var{args})} or \samp{\var{callable}(*\var{args})}.
  \bifuncindex{apply}
\end{cfuncdesc}


\begin{cfuncdesc}{PyObject*}{PyObject_CallMethod}{PyObject *o,
                                                  char *method, char *format,
                                                  \moreargs}
  Call the method named \var{method} of object \var{o} with a variable
  number of C arguments.  The C arguments are described by a
  \cfunction{Py_BuildValue()} format string that should 
  produce a tuple.  The format may be \NULL,
  indicating that no arguments are provided. Returns the result of the
  call on success, or \NULL{} on failure.  This is the equivalent of
  the Python expression \samp{\var{o}.\var{method}(\var{args})}.
\end{cfuncdesc}


\begin{cfuncdesc}{PyObject*}{PyObject_CallFunctionObjArgs}{PyObject *callable,
                                                           \moreargs,
                                                           \code{NULL}}
  Call a callable Python object \var{callable}, with a variable
  number of \ctype{PyObject*} arguments.  The arguments are provided
  as a variable number of parameters followed by \NULL.
  Returns the result of the call on success, or \NULL{} on failure.
  \versionadded{2.2}
\end{cfuncdesc}


\begin{cfuncdesc}{PyObject*}{PyObject_CallMethodObjArgs}{PyObject *o,
                                                         PyObject *name,
                                                         \moreargs,
                                                         \code{NULL}}
  Calls a method of the object \var{o}, where the name of the method
  is given as a Python string object in \var{name}.  It is called with
  a variable number of \ctype{PyObject*} arguments.  The arguments are
  provided as a variable number of parameters followed by \NULL.
  Returns the result of the call on success, or \NULL{} on failure.
  \versionadded{2.2}
\end{cfuncdesc}


\begin{cfuncdesc}{int}{PyObject_Hash}{PyObject *o}
  Compute and return the hash value of an object \var{o}.  On failure,
  return \code{-1}.  This is the equivalent of the Python expression
  \samp{hash(\var{o})}.\bifuncindex{hash}
\end{cfuncdesc}


\begin{cfuncdesc}{int}{PyObject_IsTrue}{PyObject *o}
  Returns \code{1} if the object \var{o} is considered to be true, and
  \code{0} otherwise.  This is equivalent to the Python expression
  \samp{not not \var{o}}.  On failure, return \code{-1}. 
\end{cfuncdesc}


\begin{cfuncdesc}{int}{PyObject_Not}{PyObject *o}
  Returns \code{0} if the object \var{o} is considered to be true, and
  \code{1} otherwise.  This is equivalent to the Python expression
  \samp{not \var{o}}.  On failure, return \code{-1}. 
\end{cfuncdesc}


\begin{cfuncdesc}{PyObject*}{PyObject_Type}{PyObject *o}
  When \var{o} is non-\NULL, returns a type object corresponding to
  the object type of object \var{o}. On failure, raises
  \exception{SystemError} and returns \NULL.  This is equivalent to
  the Python expression \code{type(\var{o})}.\bifuncindex{type}
  This function increments the reference count of the return value.
  There's really no reason to use this function instead of the
  common expression \code{\var{o}->ob_type}, which returns a pointer
  of type \ctype{PyTypeObject*}, except when the incremented reference
  count is needed.
\end{cfuncdesc}

\begin{cfuncdesc}{int}{PyObject_TypeCheck}{PyObject *o, PyTypeObject *type}
  Return true if the object \var{o} is of type \var{type} or a subtype
  of \var{type}.  Both parameters must be non-\NULL.
  \versionadded{2.2}
\end{cfuncdesc}

\begin{cfuncdesc}{int}{PyObject_Length}{PyObject *o}
\cfuncline{int}{PyObject_Size}{PyObject *o}
  Return the length of object \var{o}.  If the object \var{o} provides
  either the sequence and mapping protocols, the sequence length is
  returned.  On error, \code{-1} is returned.  This is the equivalent
  to the Python expression \samp{len(\var{o})}.\bifuncindex{len}
\end{cfuncdesc}


\begin{cfuncdesc}{PyObject*}{PyObject_GetItem}{PyObject *o, PyObject *key}
  Return element of \var{o} corresponding to the object \var{key} or
  \NULL{} on failure.  This is the equivalent of the Python expression
  \samp{\var{o}[\var{key}]}.
\end{cfuncdesc}


\begin{cfuncdesc}{int}{PyObject_SetItem}{PyObject *o,
                                         PyObject *key, PyObject *v}
  Map the object \var{key} to the value \var{v}.  Returns \code{-1} on
  failure.  This is the equivalent of the Python statement
  \samp{\var{o}[\var{key}] = \var{v}}.
\end{cfuncdesc}


\begin{cfuncdesc}{int}{PyObject_DelItem}{PyObject *o, PyObject *key}
  Delete the mapping for \var{key} from \var{o}.  Returns \code{-1} on
  failure. This is the equivalent of the Python statement \samp{del
  \var{o}[\var{key}]}.
\end{cfuncdesc}

\begin{cfuncdesc}{int}{PyObject_AsFileDescriptor}{PyObject *o}
  Derives a file-descriptor from a Python object.  If the object is an
  integer or long integer, its value is returned.  If not, the
  object's \method{fileno()} method is called if it exists; the method
  must return an integer or long integer, which is returned as the
  file descriptor value.  Returns \code{-1} on failure.
\end{cfuncdesc}

\begin{cfuncdesc}{PyObject*}{PyObject_Dir}{PyObject *o}
  This is equivalent to the Python expression \samp{dir(\var{o})},
  returning a (possibly empty) list of strings appropriate for the
  object argument, or \NULL{} if there was an error.  If the argument
  is \NULL, this is like the Python \samp{dir()}, returning the names
  of the current locals; in this case, if no execution frame is active
  then \NULL{} is returned but \cfunction{PyErr_Occurred()} will
  return false.
\end{cfuncdesc}

\begin{cfuncdesc}{PyObject*}{PyObject_GetIter}{PyObject *o}
  This is equivalent to the Python expression \samp{iter(\var{o})}.
  It returns a new iterator for the object argument, or the object 
  itself if the object is already an iterator.  Raises
  \exception{TypeError} and returns \NULL{} if the object cannot be
  iterated.
\end{cfuncdesc}


\section{Number Protocol \label{number}}

\begin{cfuncdesc}{int}{PyNumber_Check}{PyObject *o}
  Returns \code{1} if the object \var{o} provides numeric protocols,
  and false otherwise.  This function always succeeds.
\end{cfuncdesc}


\begin{cfuncdesc}{PyObject*}{PyNumber_Add}{PyObject *o1, PyObject *o2}
  Returns the result of adding \var{o1} and \var{o2}, or \NULL{} on
  failure.  This is the equivalent of the Python expression
  \samp{\var{o1} + \var{o2}}.
\end{cfuncdesc}


\begin{cfuncdesc}{PyObject*}{PyNumber_Subtract}{PyObject *o1, PyObject *o2}
  Returns the result of subtracting \var{o2} from \var{o1}, or \NULL{}
  on failure.  This is the equivalent of the Python expression
  \samp{\var{o1} - \var{o2}}.
\end{cfuncdesc}


\begin{cfuncdesc}{PyObject*}{PyNumber_Multiply}{PyObject *o1, PyObject *o2}
  Returns the result of multiplying \var{o1} and \var{o2}, or \NULL{}
  on failure.  This is the equivalent of the Python expression
  \samp{\var{o1} * \var{o2}}.
\end{cfuncdesc}


\begin{cfuncdesc}{PyObject*}{PyNumber_Divide}{PyObject *o1, PyObject *o2}
  Returns the result of dividing \var{o1} by \var{o2}, or \NULL{} on
  failure.  This is the equivalent of the Python expression
  \samp{\var{o1} / \var{o2}}.
\end{cfuncdesc}


\begin{cfuncdesc}{PyObject*}{PyNumber_FloorDivide}{PyObject *o1, PyObject *o2}
  Return the floor of \var{o1} divided by \var{o2}, or \NULL{} on
  failure.  This is equivalent to the ``classic'' division of
  integers.
  \versionadded{2.2}
\end{cfuncdesc}


\begin{cfuncdesc}{PyObject*}{PyNumber_TrueDivide}{PyObject *o1, PyObject *o2}
  Return a reasonable approximation for the mathematical value of
  \var{o1} divided by \var{o2}, or \NULL{} on failure.  The return
  value is ``approximate'' because binary floating point numbers are
  approximate; it is not possible to represent all real numbers in
  base two.  This function can return a floating point value when
  passed two integers.
  \versionadded{2.2}
\end{cfuncdesc}


\begin{cfuncdesc}{PyObject*}{PyNumber_Remainder}{PyObject *o1, PyObject *o2}
  Returns the remainder of dividing \var{o1} by \var{o2}, or \NULL{}
  on failure.  This is the equivalent of the Python expression
  \samp{\var{o1} \%\ \var{o2}}.
\end{cfuncdesc}


\begin{cfuncdesc}{PyObject*}{PyNumber_Divmod}{PyObject *o1, PyObject *o2}
  See the built-in function \function{divmod()}\bifuncindex{divmod}.
  Returns \NULL{} on failure.  This is the equivalent of the Python
  expression \samp{divmod(\var{o1}, \var{o2})}.
\end{cfuncdesc}


\begin{cfuncdesc}{PyObject*}{PyNumber_Power}{PyObject *o1,
                                             PyObject *o2, PyObject *o3}
  See the built-in function \function{pow()}\bifuncindex{pow}.
  Returns \NULL{} on failure.  This is the equivalent of the Python
  expression \samp{pow(\var{o1}, \var{o2}, \var{o3})}, where \var{o3}
  is optional.  If \var{o3} is to be ignored, pass \cdata{Py_None} in
  its place (passing \NULL{} for \var{o3} would cause an illegal
  memory access).
\end{cfuncdesc}


\begin{cfuncdesc}{PyObject*}{PyNumber_Negative}{PyObject *o}
  Returns the negation of \var{o} on success, or \NULL{} on failure.
  This is the equivalent of the Python expression \samp{-\var{o}}.
\end{cfuncdesc}


\begin{cfuncdesc}{PyObject*}{PyNumber_Positive}{PyObject *o}
  Returns \var{o} on success, or \NULL{} on failure.  This is the
  equivalent of the Python expression \samp{+\var{o}}.
\end{cfuncdesc}


\begin{cfuncdesc}{PyObject*}{PyNumber_Absolute}{PyObject *o}
  Returns the absolute value of \var{o}, or \NULL{} on failure.  This
  is the equivalent of the Python expression \samp{abs(\var{o})}.
  \bifuncindex{abs}
\end{cfuncdesc}


\begin{cfuncdesc}{PyObject*}{PyNumber_Invert}{PyObject *o}
  Returns the bitwise negation of \var{o} on success, or \NULL{} on
  failure.  This is the equivalent of the Python expression
  \samp{\~\var{o}}.
\end{cfuncdesc}


\begin{cfuncdesc}{PyObject*}{PyNumber_Lshift}{PyObject *o1, PyObject *o2}
  Returns the result of left shifting \var{o1} by \var{o2} on success,
  or \NULL{} on failure.  This is the equivalent of the Python
  expression \samp{\var{o1} <\code{<} \var{o2}}.
\end{cfuncdesc}


\begin{cfuncdesc}{PyObject*}{PyNumber_Rshift}{PyObject *o1, PyObject *o2}
  Returns the result of right shifting \var{o1} by \var{o2} on
  success, or \NULL{} on failure.  This is the equivalent of the
  Python expression \samp{\var{o1} >\code{>} \var{o2}}.
\end{cfuncdesc}


\begin{cfuncdesc}{PyObject*}{PyNumber_And}{PyObject *o1, PyObject *o2}
  Returns the ``bitwise and'' of \var{o1} and \var{o2} on success and
  \NULL{} on failure. This is the equivalent of the Python expression
  \samp{\var{o1} \&\ \var{o2}}.
\end{cfuncdesc}


\begin{cfuncdesc}{PyObject*}{PyNumber_Xor}{PyObject *o1, PyObject *o2}
  Returns the ``bitwise exclusive or'' of \var{o1} by \var{o2} on
  success, or \NULL{} on failure.  This is the equivalent of the
  Python expression \samp{\var{o1} \textasciicircum{} \var{o2}}.
\end{cfuncdesc}

\begin{cfuncdesc}{PyObject*}{PyNumber_Or}{PyObject *o1, PyObject *o2}
  Returns the ``bitwise or'' of \var{o1} and \var{o2} on success, or
  \NULL{} on failure.  This is the equivalent of the Python expression
  \samp{\var{o1} | \var{o2}}.
\end{cfuncdesc}


\begin{cfuncdesc}{PyObject*}{PyNumber_InPlaceAdd}{PyObject *o1, PyObject *o2}
  Returns the result of adding \var{o1} and \var{o2}, or \NULL{} on
  failure.  The operation is done \emph{in-place} when \var{o1}
  supports it.  This is the equivalent of the Python statement
  \samp{\var{o1} += \var{o2}}.
\end{cfuncdesc}


\begin{cfuncdesc}{PyObject*}{PyNumber_InPlaceSubtract}{PyObject *o1,
                                                       PyObject *o2}
  Returns the result of subtracting \var{o2} from \var{o1}, or \NULL{}
  on failure.  The operation is done \emph{in-place} when \var{o1}
  supports it.  This is the equivalent of the Python statement
  \samp{\var{o1} -= \var{o2}}.
\end{cfuncdesc}


\begin{cfuncdesc}{PyObject*}{PyNumber_InPlaceMultiply}{PyObject *o1,
                                                       PyObject *o2}
  Returns the result of multiplying \var{o1} and \var{o2}, or \NULL{}
  on failure.  The operation is done \emph{in-place} when \var{o1}
  supports it.  This is the equivalent of the Python statement
  \samp{\var{o1} *= \var{o2}}.
\end{cfuncdesc}


\begin{cfuncdesc}{PyObject*}{PyNumber_InPlaceDivide}{PyObject *o1,
                                                     PyObject *o2}
  Returns the result of dividing \var{o1} by \var{o2}, or \NULL{} on
  failure.  The operation is done \emph{in-place} when \var{o1}
  supports it. This is the equivalent of the Python statement
  \samp{\var{o1} /= \var{o2}}.
\end{cfuncdesc}


\begin{cfuncdesc}{PyObject*}{PyNumber_InPlaceFloorDivide}{PyObject *o1,
                                                          PyObject *o2}
  Returns the mathematical floor of dividing \var{o1} by \var{o2}, or
  \NULL{} on failure.  The operation is done \emph{in-place} when
  \var{o1} supports it.  This is the equivalent of the Python
  statement \samp{\var{o1} //= \var{o2}}.
  \versionadded{2.2}
\end{cfuncdesc}


\begin{cfuncdesc}{PyObject*}{PyNumber_InPlaceTrueDivide}{PyObject *o1,
                                                         PyObject *o2}
  Return a reasonable approximation for the mathematical value of
  \var{o1} divided by \var{o2}, or \NULL{} on failure.  The return
  value is ``approximate'' because binary floating point numbers are
  approximate; it is not possible to represent all real numbers in
  base two.  This function can return a floating point value when
  passed two integers.  The operation is done \emph{in-place} when
  \var{o1} supports it.
  \versionadded{2.2}
\end{cfuncdesc}


\begin{cfuncdesc}{PyObject*}{PyNumber_InPlaceRemainder}{PyObject *o1,
                                                        PyObject *o2}
  Returns the remainder of dividing \var{o1} by \var{o2}, or \NULL{}
  on failure.  The operation is done \emph{in-place} when \var{o1}
  supports it.  This is the equivalent of the Python statement
  \samp{\var{o1} \%= \var{o2}}.
\end{cfuncdesc}


\begin{cfuncdesc}{PyObject*}{PyNumber_InPlacePower}{PyObject *o1,
                                                    PyObject *o2, PyObject *o3}
  See the built-in function \function{pow()}.\bifuncindex{pow}
  Returns \NULL{} on failure.  The operation is done \emph{in-place}
  when \var{o1} supports it.  This is the equivalent of the Python
  statement \samp{\var{o1} **= \var{o2}} when o3 is \cdata{Py_None},
  or an in-place variant of \samp{pow(\var{o1}, \var{o2}, \var{o3})}
  otherwise. If \var{o3} is to be ignored, pass \cdata{Py_None} in its
  place (passing \NULL{} for \var{o3} would cause an illegal memory
  access).
\end{cfuncdesc}

\begin{cfuncdesc}{PyObject*}{PyNumber_InPlaceLshift}{PyObject *o1,
                                                     PyObject *o2}
  Returns the result of left shifting \var{o1} by \var{o2} on success,
  or \NULL{} on failure.  The operation is done \emph{in-place} when
  \var{o1} supports it.  This is the equivalent of the Python
  statement \samp{\var{o1} <\code{<=} \var{o2}}.
\end{cfuncdesc}


\begin{cfuncdesc}{PyObject*}{PyNumber_InPlaceRshift}{PyObject *o1,
                                                     PyObject *o2}
  Returns the result of right shifting \var{o1} by \var{o2} on
  success, or \NULL{} on failure.  The operation is done
  \emph{in-place} when \var{o1} supports it.  This is the equivalent
  of the Python statement \samp{\var{o1} >\code{>=} \var{o2}}.
\end{cfuncdesc}


\begin{cfuncdesc}{PyObject*}{PyNumber_InPlaceAnd}{PyObject *o1, PyObject *o2}
  Returns the ``bitwise and'' of \var{o1} and \var{o2} on success and
  \NULL{} on failure. The operation is done \emph{in-place} when
  \var{o1} supports it.  This is the equivalent of the Python
  statement \samp{\var{o1} \&= \var{o2}}.
\end{cfuncdesc}


\begin{cfuncdesc}{PyObject*}{PyNumber_InPlaceXor}{PyObject *o1, PyObject *o2}
  Returns the ``bitwise exclusive or'' of \var{o1} by \var{o2} on
  success, or \NULL{} on failure.  The operation is done
  \emph{in-place} when \var{o1} supports it.  This is the equivalent
  of the Python statement \samp{\var{o1} \textasciicircum= \var{o2}}.
\end{cfuncdesc}

\begin{cfuncdesc}{PyObject*}{PyNumber_InPlaceOr}{PyObject *o1, PyObject *o2}
  Returns the ``bitwise or'' of \var{o1} and \var{o2} on success, or
  \NULL{} on failure.  The operation is done \emph{in-place} when
  \var{o1} supports it.  This is the equivalent of the Python
  statement \samp{\var{o1} |= \var{o2}}.
\end{cfuncdesc}

\begin{cfuncdesc}{int}{PyNumber_Coerce}{PyObject **p1, PyObject **p2}
  This function takes the addresses of two variables of type
  \ctype{PyObject*}.  If the objects pointed to by \code{*\var{p1}}
  and \code{*\var{p2}} have the same type, increment their reference
  count and return \code{0} (success). If the objects can be converted
  to a common numeric type, replace \code{*p1} and \code{*p2} by their
  converted value (with 'new' reference counts), and return \code{0}.
  If no conversion is possible, or if some other error occurs, return
  \code{-1} (failure) and don't increment the reference counts.  The
  call \code{PyNumber_Coerce(\&o1, \&o2)} is equivalent to the Python
  statement \samp{\var{o1}, \var{o2} = coerce(\var{o1}, \var{o2})}.
  \bifuncindex{coerce}
\end{cfuncdesc}

\begin{cfuncdesc}{PyObject*}{PyNumber_Int}{PyObject *o}
  Returns the \var{o} converted to an integer object on success, or
  \NULL{} on failure.  If the argument is outside the integer range
  a long object will be returned instead. This is the equivalent
  of the Python expression \samp{int(\var{o})}.\bifuncindex{int}
\end{cfuncdesc}

\begin{cfuncdesc}{PyObject*}{PyNumber_Long}{PyObject *o}
  Returns the \var{o} converted to a long integer object on success,
  or \NULL{} on failure.  This is the equivalent of the Python
  expression \samp{long(\var{o})}.\bifuncindex{long}
\end{cfuncdesc}

\begin{cfuncdesc}{PyObject*}{PyNumber_Float}{PyObject *o}
  Returns the \var{o} converted to a float object on success, or
  \NULL{} on failure.  This is the equivalent of the Python expression
  \samp{float(\var{o})}.\bifuncindex{float}
\end{cfuncdesc}


\section{Sequence Protocol \label{sequence}}

\begin{cfuncdesc}{int}{PySequence_Check}{PyObject *o}
  Return \code{1} if the object provides sequence protocol, and
  \code{0} otherwise.  This function always succeeds.
\end{cfuncdesc}

\begin{cfuncdesc}{int}{PySequence_Size}{PyObject *o}
  Returns the number of objects in sequence \var{o} on success, and
  \code{-1} on failure.  For objects that do not provide sequence
  protocol, this is equivalent to the Python expression
  \samp{len(\var{o})}.\bifuncindex{len}
\end{cfuncdesc}

\begin{cfuncdesc}{int}{PySequence_Length}{PyObject *o}
  Alternate name for \cfunction{PySequence_Size()}.
\end{cfuncdesc}

\begin{cfuncdesc}{PyObject*}{PySequence_Concat}{PyObject *o1, PyObject *o2}
  Return the concatenation of \var{o1} and \var{o2} on success, and
  \NULL{} on failure.   This is the equivalent of the Python
  expression \samp{\var{o1} + \var{o2}}.
\end{cfuncdesc}


\begin{cfuncdesc}{PyObject*}{PySequence_Repeat}{PyObject *o, int count}
  Return the result of repeating sequence object \var{o} \var{count}
  times, or \NULL{} on failure.  This is the equivalent of the Python
  expression \samp{\var{o} * \var{count}}.
\end{cfuncdesc}

\begin{cfuncdesc}{PyObject*}{PySequence_InPlaceConcat}{PyObject *o1,
                                                       PyObject *o2}
  Return the concatenation of \var{o1} and \var{o2} on success, and
  \NULL{} on failure.  The operation is done \emph{in-place} when
  \var{o1} supports it.  This is the equivalent of the Python
  expression \samp{\var{o1} += \var{o2}}.
\end{cfuncdesc}


\begin{cfuncdesc}{PyObject*}{PySequence_InPlaceRepeat}{PyObject *o, int count}
  Return the result of repeating sequence object \var{o} \var{count}
  times, or \NULL{} on failure.  The operation is done \emph{in-place}
  when \var{o} supports it.  This is the equivalent of the Python
  expression \samp{\var{o} *= \var{count}}.
\end{cfuncdesc}


\begin{cfuncdesc}{PyObject*}{PySequence_GetItem}{PyObject *o, int i}
  Return the \var{i}th element of \var{o}, or \NULL{} on failure.
  This is the equivalent of the Python expression
  \samp{\var{o}[\var{i}]}.
\end{cfuncdesc}


\begin{cfuncdesc}{PyObject*}{PySequence_GetSlice}{PyObject *o, int i1, int i2}
  Return the slice of sequence object \var{o} between \var{i1} and
  \var{i2}, or \NULL{} on failure. This is the equivalent of the
  Python expression \samp{\var{o}[\var{i1}:\var{i2}]}.
\end{cfuncdesc}


\begin{cfuncdesc}{int}{PySequence_SetItem}{PyObject *o, int i, PyObject *v}
  Assign object \var{v} to the \var{i}th element of \var{o}.  Returns
  \code{-1} on failure.  This is the equivalent of the Python
  statement \samp{\var{o}[\var{i}] = \var{v}}.  This function \emph{does not}
  steal a reference to \var{v}.
\end{cfuncdesc}

\begin{cfuncdesc}{int}{PySequence_DelItem}{PyObject *o, int i}
  Delete the \var{i}th element of object \var{o}.  Returns \code{-1}
  on failure.  This is the equivalent of the Python statement
  \samp{del \var{o}[\var{i}]}.
\end{cfuncdesc}

\begin{cfuncdesc}{int}{PySequence_SetSlice}{PyObject *o, int i1,
                                            int i2, PyObject *v}
  Assign the sequence object \var{v} to the slice in sequence object
  \var{o} from \var{i1} to \var{i2}.  This is the equivalent of the
  Python statement \samp{\var{o}[\var{i1}:\var{i2}] = \var{v}}.
\end{cfuncdesc}

\begin{cfuncdesc}{int}{PySequence_DelSlice}{PyObject *o, int i1, int i2}
  Delete the slice in sequence object \var{o} from \var{i1} to
  \var{i2}.  Returns \code{-1} on failure.  This is the equivalent of
  the Python statement \samp{del \var{o}[\var{i1}:\var{i2}]}.
\end{cfuncdesc}

\begin{cfuncdesc}{PyObject*}{PySequence_Tuple}{PyObject *o}
  Returns the \var{o} as a tuple on success, and \NULL{} on failure.
  This is equivalent to the Python expression \samp{tuple(\var{o})}.
  \bifuncindex{tuple}
\end{cfuncdesc}

\begin{cfuncdesc}{int}{PySequence_Count}{PyObject *o, PyObject *value}
  Return the number of occurrences of \var{value} in \var{o}, that is,
  return the number of keys for which \code{\var{o}[\var{key}] ==
  \var{value}}.  On failure, return \code{-1}.  This is equivalent to
  the Python expression \samp{\var{o}.count(\var{value})}.
\end{cfuncdesc}

\begin{cfuncdesc}{int}{PySequence_Contains}{PyObject *o, PyObject *value}
  Determine if \var{o} contains \var{value}.  If an item in \var{o} is
  equal to \var{value}, return \code{1}, otherwise return \code{0}.
  On error, return \code{-1}.  This is equivalent to the Python
  expression \samp{\var{value} in \var{o}}.
\end{cfuncdesc}

\begin{cfuncdesc}{int}{PySequence_Index}{PyObject *o, PyObject *value}
  Return the first index \var{i} for which \code{\var{o}[\var{i}] ==
  \var{value}}.  On error, return \code{-1}.    This is equivalent to
  the Python expression \samp{\var{o}.index(\var{value})}.
\end{cfuncdesc}

\begin{cfuncdesc}{PyObject*}{PySequence_List}{PyObject *o}
  Return a list object with the same contents as the arbitrary
  sequence \var{o}.  The returned list is guaranteed to be new.
\end{cfuncdesc}

\begin{cfuncdesc}{PyObject*}{PySequence_Tuple}{PyObject *o}
  Return a tuple object with the same contents as the arbitrary
  sequence \var{o}.  If \var{o} is a tuple, a new reference will be
  returned, otherwise a tuple will be constructed with the appropriate
  contents.
\end{cfuncdesc}

\begin{cfuncdesc}{PyObject*}{PySequence_Fast}{PyObject *o, const char *m}
  Returns the sequence \var{o} as a tuple, unless it is already a
  tuple or list, in which case \var{o} is returned.  Use
  \cfunction{PySequence_Fast_GET_ITEM()} to access the members of the
  result.  Returns \NULL{} on failure.  If the object is not a
  sequence, raises \exception{TypeError} with \var{m} as the message
  text.
\end{cfuncdesc}

\begin{cfuncdesc}{PyObject*}{PySequence_Fast_GET_ITEM}{PyObject *o, int i}
  Return the \var{i}th element of \var{o}, assuming that \var{o} was
  returned by \cfunction{PySequence_Fast()}, \var{o} is not \NULL,
  and that \var{i} is within bounds.
\end{cfuncdesc}

\begin{cfuncdesc}{PyObject**}{PySequence_Fast_ITEMS}{PyObject *o}
  Return the underlying array of PyObject pointers.  Assumes that
  \var{o} was returned by \cfunction{PySequence_Fast()} and
  \var{o} is not \NULL.
  \versionadded{2.4}  
\end{cfuncdesc}

\begin{cfuncdesc}{PyObject*}{PySequence_ITEM}{PyObject *o, int i}
  Return the \var{i}th element of \var{o} or \NULL{} on failure.
  Macro form of \cfunction{PySequence_GetItem()} but without checking
  that \cfunction{PySequence_Check(\var{o})} is true and without
  adjustment for negative indices.
  \versionadded{2.3}
\end{cfuncdesc}

\begin{cfuncdesc}{int}{PySequence_Fast_GET_SIZE}{PyObject *o}
  Returns the length of \var{o}, assuming that \var{o} was
  returned by \cfunction{PySequence_Fast()} and that \var{o} is
  not \NULL.  The size can also be gotten by calling
  \cfunction{PySequence_Size()} on \var{o}, but
  \cfunction{PySequence_Fast_GET_SIZE()} is faster because it can
  assume \var{o} is a list or tuple.
\end{cfuncdesc}


\section{Mapping Protocol \label{mapping}}

\begin{cfuncdesc}{int}{PyMapping_Check}{PyObject *o}
  Return \code{1} if the object provides mapping protocol, and
  \code{0} otherwise.  This function always succeeds.
\end{cfuncdesc}


\begin{cfuncdesc}{int}{PyMapping_Length}{PyObject *o}
  Returns the number of keys in object \var{o} on success, and
  \code{-1} on failure.  For objects that do not provide mapping
  protocol, this is equivalent to the Python expression
  \samp{len(\var{o})}.\bifuncindex{len}
\end{cfuncdesc}


\begin{cfuncdesc}{int}{PyMapping_DelItemString}{PyObject *o, char *key}
  Remove the mapping for object \var{key} from the object \var{o}.
  Return \code{-1} on failure.  This is equivalent to the Python
  statement \samp{del \var{o}[\var{key}]}.
\end{cfuncdesc}


\begin{cfuncdesc}{int}{PyMapping_DelItem}{PyObject *o, PyObject *key}
  Remove the mapping for object \var{key} from the object \var{o}.
  Return \code{-1} on failure.  This is equivalent to the Python
  statement \samp{del \var{o}[\var{key}]}.
\end{cfuncdesc}


\begin{cfuncdesc}{int}{PyMapping_HasKeyString}{PyObject *o, char *key}
  On success, return \code{1} if the mapping object has the key
  \var{key} and \code{0} otherwise.  This is equivalent to the Python
  expression \samp{\var{o}.has_key(\var{key})}.  This function always
  succeeds.
\end{cfuncdesc}


\begin{cfuncdesc}{int}{PyMapping_HasKey}{PyObject *o, PyObject *key}
  Return \code{1} if the mapping object has the key \var{key} and
  \code{0} otherwise.  This is equivalent to the Python expression
  \samp{\var{o}.has_key(\var{key})}.  This function always succeeds.
\end{cfuncdesc}


\begin{cfuncdesc}{PyObject*}{PyMapping_Keys}{PyObject *o}
  On success, return a list of the keys in object \var{o}.  On
  failure, return \NULL. This is equivalent to the Python expression
  \samp{\var{o}.keys()}.
\end{cfuncdesc}


\begin{cfuncdesc}{PyObject*}{PyMapping_Values}{PyObject *o}
  On success, return a list of the values in object \var{o}.  On
  failure, return \NULL. This is equivalent to the Python expression
  \samp{\var{o}.values()}.
\end{cfuncdesc}


\begin{cfuncdesc}{PyObject*}{PyMapping_Items}{PyObject *o}
  On success, return a list of the items in object \var{o}, where each
  item is a tuple containing a key-value pair.  On failure, return
  \NULL. This is equivalent to the Python expression
  \samp{\var{o}.items()}.
\end{cfuncdesc}


\begin{cfuncdesc}{PyObject*}{PyMapping_GetItemString}{PyObject *o, char *key}
  Return element of \var{o} corresponding to the object \var{key} or
  \NULL{} on failure. This is the equivalent of the Python expression
  \samp{\var{o}[\var{key}]}.
\end{cfuncdesc}

\begin{cfuncdesc}{int}{PyMapping_SetItemString}{PyObject *o, char *key,
                                                PyObject *v}
  Map the object \var{key} to the value \var{v} in object \var{o}.
  Returns \code{-1} on failure.  This is the equivalent of the Python
  statement \samp{\var{o}[\var{key}] = \var{v}}.
\end{cfuncdesc}


\section{Iterator Protocol \label{iterator}}

\versionadded{2.2}

There are only a couple of functions specifically for working with
iterators.

\begin{cfuncdesc}{int}{PyIter_Check}{PyObject *o}
  Return true if the object \var{o} supports the iterator protocol.
\end{cfuncdesc}

\begin{cfuncdesc}{PyObject*}{PyIter_Next}{PyObject *o}
  Return the next value from the iteration \var{o}.  If the object is
  an iterator, this retrieves the next value from the iteration, and
  returns \NULL{} with no exception set if there are no remaining
  items.  If the object is not an iterator, \exception{TypeError} is
  raised, or if there is an error in retrieving the item, returns
  \NULL{} and passes along the exception.
\end{cfuncdesc}

To write a loop which iterates over an iterator, the C code should
look something like this:

\begin{verbatim}
PyObject *iterator = PyObject_GetIter(obj);
PyObject *item;

if (iterator == NULL) {
    /* propagate error */
}

while (item = PyIter_Next(iterator)) {
    /* do something with item */
    ...
    /* release reference when done */
    Py_DECREF(item);
}

Py_DECREF(iterator);

if (PyErr_Occurred()) {
    /* propagate error */
}
else {
    /* continue doing useful work */
}
\end{verbatim}


\section{Buffer Protocol \label{abstract-buffer}}

\begin{cfuncdesc}{int}{PyObject_AsCharBuffer}{PyObject *obj,
                                              const char **buffer,
                                              int *buffer_len}
  Returns a pointer to a read-only memory location useable as character-
  based input.  The \var{obj} argument must support the single-segment
  character buffer interface.  On success, returns \code{0}, sets
  \var{buffer} to the memory location and \var{buffer_len} to the buffer
  length.  Returns \code{-1} and sets a \exception{TypeError} on error.
  \versionadded{1.6}
\end{cfuncdesc}

\begin{cfuncdesc}{int}{PyObject_AsReadBuffer}{PyObject *obj,
                                              const void **buffer,
                                              int *buffer_len}
  Returns a pointer to a read-only memory location containing
  arbitrary data.  The \var{obj} argument must support the
  single-segment readable buffer interface.  On success, returns
  \code{0}, sets \var{buffer} to the memory location and \var{buffer_len}
  to the buffer length.  Returns \code{-1} and sets a
  \exception{TypeError} on error.
  \versionadded{1.6}
\end{cfuncdesc}

\begin{cfuncdesc}{int}{PyObject_CheckReadBuffer}{PyObject *o}
  Returns \code{1} if \var{o} supports the single-segment readable
  buffer interface.  Otherwise returns \code{0}.
  \versionadded{2.2}
\end{cfuncdesc}

\begin{cfuncdesc}{int}{PyObject_AsWriteBuffer}{PyObject *obj,
                                               void **buffer,
                                               int *buffer_len}
  Returns a pointer to a writeable memory location.  The \var{obj}
  argument must support the single-segment, character buffer
  interface.  On success, returns \code{0}, sets \var{buffer} to the
  memory location and \var{buffer_len} to the buffer length.  Returns
  \code{-1} and sets a \exception{TypeError} on error.
  \versionadded{1.6}
\end{cfuncdesc}
