\chapter{Introduction \label{intro}}


The Application Programmer's Interface to Python gives C and
\Cpp{} programmers access to the Python interpreter at a variety of
levels.  The API is equally usable from \Cpp, but for brevity it is
generally referred to as the Python/C API.  There are two
fundamentally different reasons for using the Python/C API.  The first
reason is to write \emph{extension modules} for specific purposes;
these are C modules that extend the Python interpreter.  This is
probably the most common use.  The second reason is to use Python as a
component in a larger application; this technique is generally
referred to as \dfn{embedding} Python in an application.

Writing an extension module is a relatively well-understood process, 
where a ``cookbook'' approach works well.  There are several tools 
that automate the process to some extent.  While people have embedded 
Python in other applications since its early existence, the process of 
embedding Python is less straightforward than writing an extension.  

Many API functions are useful independent of whether you're embedding 
or extending Python; moreover, most applications that embed Python 
will need to provide a custom extension as well, so it's probably a 
good idea to become familiar with writing an extension before 
attempting to embed Python in a real application.


\section{Include Files \label{includes}}

All function, type and macro definitions needed to use the Python/C
API are included in your code by the following line:

\begin{verbatim}
#include "Python.h"
\end{verbatim}

This implies inclusion of the following standard headers:
\code{<stdio.h>}, \code{<string.h>}, \code{<errno.h>},
\code{<limits.h>}, and \code{<stdlib.h>} (if available).

\begin{notice}[warning]
  Since Python may define some pre-processor definitions which affect
  the standard headers on some systems, you \emph{must} include
  \file{Python.h} before any standard headers are included.
\end{notice}

All user visible names defined by Python.h (except those defined by
the included standard headers) have one of the prefixes \samp{Py} or
\samp{_Py}.  Names beginning with \samp{_Py} are for internal use by
the Python implementation and should not be used by extension writers.
Structure member names do not have a reserved prefix.

\strong{Important:} user code should never define names that begin
with \samp{Py} or \samp{_Py}.  This confuses the reader, and
jeopardizes the portability of the user code to future Python
versions, which may define additional names beginning with one of
these prefixes.

The header files are typically installed with Python.  On \UNIX, these 
are located in the directories
\file{\envvar{prefix}/include/python\var{version}/} and
\file{\envvar{exec_prefix}/include/python\var{version}/}, where
\envvar{prefix} and \envvar{exec_prefix} are defined by the
corresponding parameters to Python's \program{configure} script and
\var{version} is \code{sys.version[:3]}.  On Windows, the headers are
installed in \file{\envvar{prefix}/include}, where \envvar{prefix} is
the installation directory specified to the installer.

To include the headers, place both directories (if different) on your
compiler's search path for includes.  Do \emph{not} place the parent
directories on the search path and then use
\samp{\#include <python\shortversion/Python.h>}; this will break on
multi-platform builds since the platform independent headers under
\envvar{prefix} include the platform specific headers from
\envvar{exec_prefix}.

\Cpp{} users should note that though the API is defined entirely using
C, the header files do properly declare the entry points to be
\code{extern "C"}, so there is no need to do anything special to use
the API from \Cpp.


\section{Objects, Types and Reference Counts \label{objects}}

Most Python/C API functions have one or more arguments as well as a
return value of type \ctype{PyObject*}.  This type is a pointer
to an opaque data type representing an arbitrary Python
object.  Since all Python object types are treated the same way by the
Python language in most situations (e.g., assignments, scope rules,
and argument passing), it is only fitting that they should be
represented by a single C type.  Almost all Python objects live on the
heap: you never declare an automatic or static variable of type
\ctype{PyObject}, only pointer variables of type \ctype{PyObject*} can 
be declared.  The sole exception are the type objects\obindex{type};
since these must never be deallocated, they are typically static
\ctype{PyTypeObject} objects.

All Python objects (even Python integers) have a \dfn{type} and a
\dfn{reference count}.  An object's type determines what kind of object 
it is (e.g., an integer, a list, or a user-defined function; there are 
many more as explained in the \citetitle[../ref/ref.html]{Python
Reference Manual}).  For each of the well-known types there is a macro
to check whether an object is of that type; for instance,
\samp{PyList_Check(\var{a})} is true if (and only if) the object
pointed to by \var{a} is a Python list.


\subsection{Reference Counts \label{refcounts}}

The reference count is important because today's computers have a 
finite (and often severely limited) memory size; it counts how many 
different places there are that have a reference to an object.  Such a 
place could be another object, or a global (or static) C variable, or 
a local variable in some C function.  When an object's reference count 
becomes zero, the object is deallocated.  If it contains references to 
other objects, their reference count is decremented.  Those other 
objects may be deallocated in turn, if this decrement makes their 
reference count become zero, and so on.  (There's an obvious problem 
with objects that reference each other here; for now, the solution is 
``don't do that.'')

Reference counts are always manipulated explicitly.  The normal way is 
to use the macro \cfunction{Py_INCREF()}\ttindex{Py_INCREF()} to
increment an object's reference count by one, and
\cfunction{Py_DECREF()}\ttindex{Py_DECREF()} to decrement it by  
one.  The \cfunction{Py_DECREF()} macro is considerably more complex
than the incref one, since it must check whether the reference count
becomes zero and then cause the object's deallocator to be called.
The deallocator is a function pointer contained in the object's type
structure.  The type-specific deallocator takes care of decrementing
the reference counts for other objects contained in the object if this
is a compound object type, such as a list, as well as performing any
additional finalization that's needed.  There's no chance that the
reference count can overflow; at least as many bits are used to hold
the reference count as there are distinct memory locations in virtual
memory (assuming \code{sizeof(long) >= sizeof(char*)}).  Thus, the
reference count increment is a simple operation.

It is not necessary to increment an object's reference count for every 
local variable that contains a pointer to an object.  In theory, the 
object's reference count goes up by one when the variable is made to 
point to it and it goes down by one when the variable goes out of 
scope.  However, these two cancel each other out, so at the end the 
reference count hasn't changed.  The only real reason to use the 
reference count is to prevent the object from being deallocated as 
long as our variable is pointing to it.  If we know that there is at 
least one other reference to the object that lives at least as long as 
our variable, there is no need to increment the reference count 
temporarily.  An important situation where this arises is in objects 
that are passed as arguments to C functions in an extension module 
that are called from Python; the call mechanism guarantees to hold a 
reference to every argument for the duration of the call.

However, a common pitfall is to extract an object from a list and
hold on to it for a while without incrementing its reference count.
Some other operation might conceivably remove the object from the
list, decrementing its reference count and possible deallocating it.
The real danger is that innocent-looking operations may invoke
arbitrary Python code which could do this; there is a code path which
allows control to flow back to the user from a \cfunction{Py_DECREF()},
so almost any operation is potentially dangerous.

A safe approach is to always use the generic operations (functions 
whose name begins with \samp{PyObject_}, \samp{PyNumber_},
\samp{PySequence_} or \samp{PyMapping_}).  These operations always
increment the reference count of the object they return.  This leaves
the caller with the responsibility to call
\cfunction{Py_DECREF()} when they are done with the result; this soon
becomes second nature.


\subsubsection{Reference Count Details \label{refcountDetails}}

The reference count behavior of functions in the Python/C API is best 
explained in terms of \emph{ownership of references}.  Ownership
pertains to references, never to objects (objects are not owned: they
are always shared).  "Owning a reference" means being responsible for
calling Py_DECREF on it when the reference is no longer needed. 
Ownership can also be transferred, meaning that the code that receives
ownership of the reference then becomes responsible for eventually
decref'ing it by calling \cfunction{Py_DECREF()} or
\cfunction{Py_XDECREF()} when it's no longer needed --or passing on
this responsibility (usually to its caller).
When a function passes ownership of a reference on to its caller, the
caller is said to receive a \emph{new} reference.  When no ownership
is transferred, the caller is said to \emph{borrow} the reference.
Nothing needs to be done for a borrowed reference.

Conversely, when a calling function passes it a reference to an 
object, there are two possibilities: the function \emph{steals} a 
reference to the object, or it does not.  Few functions steal 
references; the two notable exceptions are
\cfunction{PyList_SetItem()}\ttindex{PyList_SetItem()} and
\cfunction{PyTuple_SetItem()}\ttindex{PyTuple_SetItem()}, which 
steal a reference to the item (but not to the tuple or list into which
the item is put!).  These functions were designed to steal a reference
because of a common idiom for populating a tuple or list with newly
created objects; for example, the code to create the tuple \code{(1,
2, "three")} could look like this (forgetting about error handling for
the moment; a better way to code this is shown below):

\begin{verbatim}
PyObject *t;

t = PyTuple_New(3);
PyTuple_SetItem(t, 0, PyInt_FromLong(1L));
PyTuple_SetItem(t, 1, PyInt_FromLong(2L));
PyTuple_SetItem(t, 2, PyString_FromString("three"));
\end{verbatim}

Incidentally, \cfunction{PyTuple_SetItem()} is the \emph{only} way to
set tuple items; \cfunction{PySequence_SetItem()} and
\cfunction{PyObject_SetItem()} refuse to do this since tuples are an
immutable data type.  You should only use
\cfunction{PyTuple_SetItem()} for tuples that you are creating
yourself.

Equivalent code for populating a list can be written using 
\cfunction{PyList_New()} and \cfunction{PyList_SetItem()}.  Such code
can also use \cfunction{PySequence_SetItem()}; this illustrates the
difference between the two (the extra \cfunction{Py_DECREF()} calls):

\begin{verbatim}
PyObject *l, *x;

l = PyList_New(3);
x = PyInt_FromLong(1L);
PySequence_SetItem(l, 0, x); Py_DECREF(x);
x = PyInt_FromLong(2L);
PySequence_SetItem(l, 1, x); Py_DECREF(x);
x = PyString_FromString("three");
PySequence_SetItem(l, 2, x); Py_DECREF(x);
\end{verbatim}

You might find it strange that the ``recommended'' approach takes more
code.  However, in practice, you will rarely use these ways of
creating and populating a tuple or list.  There's a generic function,
\cfunction{Py_BuildValue()}, that can create most common objects from
C values, directed by a \dfn{format string}.  For example, the
above two blocks of code could be replaced by the following (which
also takes care of the error checking):

\begin{verbatim}
PyObject *t, *l;

t = Py_BuildValue("(iis)", 1, 2, "three");
l = Py_BuildValue("[iis]", 1, 2, "three");
\end{verbatim}

It is much more common to use \cfunction{PyObject_SetItem()} and
friends with items whose references you are only borrowing, like
arguments that were passed in to the function you are writing.  In
that case, their behaviour regarding reference counts is much saner,
since you don't have to increment a reference count so you can give a
reference away (``have it be stolen'').  For example, this function
sets all items of a list (actually, any mutable sequence) to a given
item:

\begin{verbatim}
int
set_all(PyObject *target, PyObject *item)
{
    int i, n;

    n = PyObject_Length(target);
    if (n < 0)
        return -1;
    for (i = 0; i < n; i++) {
        if (PyObject_SetItem(target, i, item) < 0)
            return -1;
    }
    return 0;
}
\end{verbatim}
\ttindex{set_all()}

The situation is slightly different for function return values.  
While passing a reference to most functions does not change your 
ownership responsibilities for that reference, many functions that 
return a reference to an object give you ownership of the reference.
The reason is simple: in many cases, the returned object is created 
on the fly, and the reference you get is the only reference to the 
object.  Therefore, the generic functions that return object 
references, like \cfunction{PyObject_GetItem()} and 
\cfunction{PySequence_GetItem()}, always return a new reference (the
caller becomes the owner of the reference).

It is important to realize that whether you own a reference returned 
by a function depends on which function you call only --- \emph{the
plumage} (the type of the object passed as an
argument to the function) \emph{doesn't enter into it!}  Thus, if you 
extract an item from a list using \cfunction{PyList_GetItem()}, you
don't own the reference --- but if you obtain the same item from the
same list using \cfunction{PySequence_GetItem()} (which happens to
take exactly the same arguments), you do own a reference to the
returned object.

Here is an example of how you could write a function that computes the
sum of the items in a list of integers; once using 
\cfunction{PyList_GetItem()}\ttindex{PyList_GetItem()}, and once using
\cfunction{PySequence_GetItem()}\ttindex{PySequence_GetItem()}.

\begin{verbatim}
long
sum_list(PyObject *list)
{
    int i, n;
    long total = 0;
    PyObject *item;

    n = PyList_Size(list);
    if (n < 0)
        return -1; /* Not a list */
    for (i = 0; i < n; i++) {
        item = PyList_GetItem(list, i); /* Can't fail */
        if (!PyInt_Check(item)) continue; /* Skip non-integers */
        total += PyInt_AsLong(item);
    }
    return total;
}
\end{verbatim}
\ttindex{sum_list()}

\begin{verbatim}
long
sum_sequence(PyObject *sequence)
{
    int i, n;
    long total = 0;
    PyObject *item;
    n = PySequence_Length(sequence);
    if (n < 0)
        return -1; /* Has no length */
    for (i = 0; i < n; i++) {
        item = PySequence_GetItem(sequence, i);
        if (item == NULL)
            return -1; /* Not a sequence, or other failure */
        if (PyInt_Check(item))
            total += PyInt_AsLong(item);
        Py_DECREF(item); /* Discard reference ownership */
    }
    return total;
}
\end{verbatim}
\ttindex{sum_sequence()}


\subsection{Types \label{types}}

There are few other data types that play a significant role in 
the Python/C API; most are simple C types such as \ctype{int}, 
\ctype{long}, \ctype{double} and \ctype{char*}.  A few structure types 
are used to describe static tables used to list the functions exported 
by a module or the data attributes of a new object type, and another
is used to describe the value of a complex number.  These will 
be discussed together with the functions that use them.


\section{Exceptions \label{exceptions}}

The Python programmer only needs to deal with exceptions if specific 
error handling is required; unhandled exceptions are automatically 
propagated to the caller, then to the caller's caller, and so on, until
they reach the top-level interpreter, where they are reported to the 
user accompanied by a stack traceback.

For C programmers, however, error checking always has to be explicit.  
All functions in the Python/C API can raise exceptions, unless an 
explicit claim is made otherwise in a function's documentation.  In 
general, when a function encounters an error, it sets an exception, 
discards any object references that it owns, and returns an 
error indicator --- usually \NULL{} or \code{-1}.  A few functions 
return a Boolean true/false result, with false indicating an error.
Very few functions return no explicit error indicator or have an 
ambiguous return value, and require explicit testing for errors with 
\cfunction{PyErr_Occurred()}\ttindex{PyErr_Occurred()}.

Exception state is maintained in per-thread storage (this is 
equivalent to using global storage in an unthreaded application).  A 
thread can be in one of two states: an exception has occurred, or not.
The function \cfunction{PyErr_Occurred()} can be used to check for
this: it returns a borrowed reference to the exception type object
when an exception has occurred, and \NULL{} otherwise.  There are a
number of functions to set the exception state:
\cfunction{PyErr_SetString()}\ttindex{PyErr_SetString()} is the most
common (though not the most general) function to set the exception
state, and \cfunction{PyErr_Clear()}\ttindex{PyErr_Clear()} clears the
exception state.

The full exception state consists of three objects (all of which can 
be \NULL): the exception type, the corresponding exception 
value, and the traceback.  These have the same meanings as the Python
\withsubitem{(in module sys)}{
  \ttindex{exc_type}\ttindex{exc_value}\ttindex{exc_traceback}}
objects \code{sys.exc_type}, \code{sys.exc_value}, and
\code{sys.exc_traceback}; however, they are not the same: the Python
objects represent the last exception being handled by a Python 
\keyword{try} \ldots\ \keyword{except} statement, while the C level
exception state only exists while an exception is being passed on
between C functions until it reaches the Python bytecode interpreter's 
main loop, which takes care of transferring it to \code{sys.exc_type}
and friends.

Note that starting with Python 1.5, the preferred, thread-safe way to 
access the exception state from Python code is to call the function
\withsubitem{(in module sys)}{\ttindex{exc_info()}}
\function{sys.exc_info()}, which returns the per-thread exception state 
for Python code.  Also, the semantics of both ways to access the 
exception state have changed so that a function which catches an 
exception will save and restore its thread's exception state so as to 
preserve the exception state of its caller.  This prevents common bugs 
in exception handling code caused by an innocent-looking function 
overwriting the exception being handled; it also reduces the often 
unwanted lifetime extension for objects that are referenced by the 
stack frames in the traceback.

As a general principle, a function that calls another function to 
perform some task should check whether the called function raised an 
exception, and if so, pass the exception state on to its caller.  It 
should discard any object references that it owns, and return an 
error indicator, but it should \emph{not} set another exception ---
that would overwrite the exception that was just raised, and lose
important information about the exact cause of the error.

A simple example of detecting exceptions and passing them on is shown
in the \cfunction{sum_sequence()}\ttindex{sum_sequence()} example
above.  It so happens that that example doesn't need to clean up any
owned references when it detects an error.  The following example
function shows some error cleanup.  First, to remind you why you like
Python, we show the equivalent Python code:

\begin{verbatim}
def incr_item(dict, key):
    try:
        item = dict[key]
    except KeyError:
        item = 0
    dict[key] = item + 1
\end{verbatim}
\ttindex{incr_item()}

Here is the corresponding C code, in all its glory:

\begin{verbatim}
int
incr_item(PyObject *dict, PyObject *key)
{
    /* Objects all initialized to NULL for Py_XDECREF */
    PyObject *item = NULL, *const_one = NULL, *incremented_item = NULL;
    int rv = -1; /* Return value initialized to -1 (failure) */

    item = PyObject_GetItem(dict, key);
    if (item == NULL) {
        /* Handle KeyError only: */
        if (!PyErr_ExceptionMatches(PyExc_KeyError))
            goto error;

        /* Clear the error and use zero: */
        PyErr_Clear();
        item = PyInt_FromLong(0L);
        if (item == NULL)
            goto error;
    }
    const_one = PyInt_FromLong(1L);
    if (const_one == NULL)
        goto error;

    incremented_item = PyNumber_Add(item, const_one);
    if (incremented_item == NULL)
        goto error;

    if (PyObject_SetItem(dict, key, incremented_item) < 0)
        goto error;
    rv = 0; /* Success */
    /* Continue with cleanup code */

 error:
    /* Cleanup code, shared by success and failure path */

    /* Use Py_XDECREF() to ignore NULL references */
    Py_XDECREF(item);
    Py_XDECREF(const_one);
    Py_XDECREF(incremented_item);

    return rv; /* -1 for error, 0 for success */
}
\end{verbatim}
\ttindex{incr_item()}

This example represents an endorsed use of the \keyword{goto} statement 
in C!  It illustrates the use of
\cfunction{PyErr_ExceptionMatches()}\ttindex{PyErr_ExceptionMatches()} and
\cfunction{PyErr_Clear()}\ttindex{PyErr_Clear()} to
handle specific exceptions, and the use of
\cfunction{Py_XDECREF()}\ttindex{Py_XDECREF()} to
dispose of owned references that may be \NULL{} (note the
\character{X} in the name; \cfunction{Py_DECREF()} would crash when
confronted with a \NULL{} reference).  It is important that the
variables used to hold owned references are initialized to \NULL{} for
this to work; likewise, the proposed return value is initialized to
\code{-1} (failure) and only set to success after the final call made
is successful.


\section{Embedding Python \label{embedding}}

The one important task that only embedders (as opposed to extension
writers) of the Python interpreter have to worry about is the
initialization, and possibly the finalization, of the Python
interpreter.  Most functionality of the interpreter can only be used
after the interpreter has been initialized.

The basic initialization function is
\cfunction{Py_Initialize()}\ttindex{Py_Initialize()}.
This initializes the table of loaded modules, and creates the
fundamental modules \module{__builtin__}\refbimodindex{__builtin__},
\module{__main__}\refbimodindex{__main__}, \module{sys}\refbimodindex{sys},
and \module{exceptions}.\refbimodindex{exceptions}  It also initializes
the module search path (\code{sys.path}).%
\indexiii{module}{search}{path}
\withsubitem{(in module sys)}{\ttindex{path}}

\cfunction{Py_Initialize()} does not set the ``script argument list'' 
(\code{sys.argv}).  If this variable is needed by Python code that 
will be executed later, it must be set explicitly with a call to 
\code{PySys_SetArgv(\var{argc},
\var{argv})}\ttindex{PySys_SetArgv()} subsequent to the call to
\cfunction{Py_Initialize()}.

On most systems (in particular, on \UNIX{} and Windows, although the
details are slightly different),
\cfunction{Py_Initialize()} calculates the module search path based
upon its best guess for the location of the standard Python
interpreter executable, assuming that the Python library is found in a
fixed location relative to the Python interpreter executable.  In
particular, it looks for a directory named
\file{lib/python\shortversion} relative to the parent directory where
the executable named \file{python} is found on the shell command
search path (the environment variable \envvar{PATH}).

For instance, if the Python executable is found in
\file{/usr/local/bin/python}, it will assume that the libraries are in
\file{/usr/local/lib/python\shortversion}.  (In fact, this particular path
is also the ``fallback'' location, used when no executable file named
\file{python} is found along \envvar{PATH}.)  The user can override
this behavior by setting the environment variable \envvar{PYTHONHOME},
or insert additional directories in front of the standard path by
setting \envvar{PYTHONPATH}.

The embedding application can steer the search by calling 
\code{Py_SetProgramName(\var{file})}\ttindex{Py_SetProgramName()} \emph{before} calling 
\cfunction{Py_Initialize()}.  Note that \envvar{PYTHONHOME} still
overrides this and \envvar{PYTHONPATH} is still inserted in front of
the standard path.  An application that requires total control has to
provide its own implementation of
\cfunction{Py_GetPath()}\ttindex{Py_GetPath()},
\cfunction{Py_GetPrefix()}\ttindex{Py_GetPrefix()},
\cfunction{Py_GetExecPrefix()}\ttindex{Py_GetExecPrefix()}, and
\cfunction{Py_GetProgramFullPath()}\ttindex{Py_GetProgramFullPath()} (all
defined in \file{Modules/getpath.c}).

Sometimes, it is desirable to ``uninitialize'' Python.  For instance, 
the application may want to start over (make another call to 
\cfunction{Py_Initialize()}) or the application is simply done with its 
use of Python and wants to free all memory allocated by Python.  This
can be accomplished by calling \cfunction{Py_Finalize()}.  The function
\cfunction{Py_IsInitialized()}\ttindex{Py_IsInitialized()} returns
true if Python is currently in the initialized state.  More
information about these functions is given in a later chapter.
