\chapter{Undocumented Modules \label{undocumented-modules}}


The modules in this chapter are poorly documented (if at all).  If you
wish to contribute documentation of any of these modules, please get in
touch with
\ulink{\email{python-docs@python.org}}{mailto:python-docs@python.org}.

\localmoduletable


\section{\module{applesingle} --- AppleSingle decoder}
\declaremodule{standard}{applesingle}
  \platform{Mac}
\modulesynopsis{Rudimentary decoder for AppleSingle format files.}


\section{\module{buildtools} --- Helper module for BuildApplet and Friends}
\declaremodule{standard}{buildtools}
  \platform{Mac}
\modulesynopsis{Helper module for BuildApplet, BuildApplication and
                macfreeze.}


\section{\module{py_resource} --- Resources from Python code}
\declaremodule[pyresource]{standard}{py_resource}
  \platform{Mac}
\modulesynopsis{Helper to create \texttt{'PYC~'} resources for compiled
                applications.}

This module is primarily used as a help module for
\program{BuildApplet} and \program{BuildApplication}.  It is able to
store compiled Python code as \texttt{'PYC~'} resources in a file.


\section{\module{cfmfile} --- Code Fragment Resource module}
\declaremodule{standard}{cfmfile}
  \platform{Mac}
\modulesynopsis{Code Fragment Resource module.}

\module{cfmfile} is a module that understands Code Fragments and the
accompanying ``cfrg'' resources. It can parse them and merge them, and is
used by BuildApplication to combine all plugin modules to a single
executable.


\section{\module{icopen} --- Internet Config replacement for \method{open()}}
\declaremodule{standard}{icopen}
  \platform{Mac}
\modulesynopsis{Internet Config replacement for \method{open()}.}

Importing \module{icopen} will replace the builtin \method{open()}
with a version that uses Internet Config to set file type and creator
for new files.


\section{\module{macerrors} --- Mac OS Errors}
\declaremodule{standard}{macerrors}
  \platform{Mac}
\modulesynopsis{Constant definitions for many Mac OS error codes.}

\module{macerrors} cotains constant definitions for many Mac OS error
codes.


\section{\module{macfsn} --- NavServices calls}
\declaremodule{standard}{macfsn}
  \platform{Mac}
\modulesynopsis{NavServices versions of StandardFile calls.}


\module{macfsn} contains wrapper functions that have the same API as
the \refmodule{macfs} StandardFile calls, but are implemented with
Navigation Services through the
\refmodule{Nav}\refbimodindex{Nav} module.  Importing it will replace
the methods in \refmodule{macfs}\refbimodindex{macfs} with these, if
Navigation Services is available on your machine.


\section{\module{macresource} --- Locate script resources}
\declaremodule{standard}{macresource}
  \platform{Mac}
\modulesynopsis{Locate script resources.}

\module{macresource} helps scripts finding their resources, such as
dialogs and menus, without requiring special case code for when the
script is run under MacPython, as a MacPython applet or under OSX Python.

\section{\module{Nav} --- NavServices calls}
\declaremodule{standard}{Nac}
  \platform{Mac}
\modulesynopsis{Interface to Navigation Services.}

A low-level interface to Navigation Services. 


\section{\module{mactty} --- Serial line connections}
\declaremodule{standard}{mactty}
  \platform{Mac}
\modulesynopsis{Easy access serial to line connections.}

This module is only available under Mac OS 9 or earlier in classic PPC
MacPython.

\section{\module{mkcwproject} --- Create CodeWarrior projects}
\declaremodule{standard}{mkcwproject}
  \platform{Mac}
\modulesynopsis{Create CodeWarrior projects.}

\module{mkcwproject} creates project files for the Metrowerks CodeWarrior
development environment. It is a helper module for
\refmodule{distutils}\refbimodindex{distutils} but can be used separately for more
control.


\section{\module{nsremote} --- Wrapper around Netscape OSA modules}
\declaremodule{standard}{nsremote}
  \platform{Mac}
\modulesynopsis{Wrapper around Netscape OSA modules.}

\module{nsremote} is a wrapper around the Netscape OSA modules that
allows you to easily send your browser to a given URL.  A related
module that may be of interest is the \module{webbrowser} module,
documented in the \citetitle[../lib/lib.html]{Python Library
Reference}.


\section{\module{PixMapWrapper} --- Wrapper for PixMap objects}
\declaremodule{standard}{PixMapWrapper}
  \platform{Mac}
\modulesynopsis{Wrapper for PixMap objects.}

\module{PixMapWrapper} wraps a PixMap object with a Python object that
allows access to the fields by name. It also has methods to convert
to and from \module{PIL} images.


\section{\module{preferences} --- Application preferences manager}
\declaremodule{standard}{preferences}
  \platform{Mac}
\modulesynopsis{Nice application preferences manager with support for
                defaults.}

The \module{preferences} module allows storage of user preferences in
the system-wide preferences folder, with defaults coming from the
application itself and the possibility to override preferences for
specific situations.


\section{\module{pythonprefs} --- Preferences manager for Python}
\declaremodule{standard}{pythonprefs}
  \platform{Mac}
\modulesynopsis{Specialized preferences manager for the Python
                interpreter.}

This module is a specialization of the \refmodule{preferences} module
that allows reading and writing of the preferences for the Python
interpreter.


\section{\module{quietconsole} --- Non-visible standard output}
\declaremodule{standard}{quietconsole}
  \platform{Mac}
\modulesynopsis{Buffered, non-visible standard output.}

\module{quietconsole} allows you to keep stdio output in a buffer
without displaying it (or without displaying the stdout window
altogether, if set with \program{EditPythonPrefs}) until you try to read from
stdin or disable the buffering, at which point all the saved output is
sent to the window.  Good for programs with graphical user interfaces
that do want to display their output at a crash.


\section{\module{videoreader} --- Read QuickTime movies}
\declaremodule{standard}{videoreader}
  \platform{Mac}
\modulesynopsis{Read QuickTime movies frame by frame for further processing.}

\module{videoreader} reads and decodes QuickTime movies and passes
a stream of images to your program. It also provides some support for
audio tracks.

\section{\module{W} --- Widgets built on \module{FrameWork}}
\declaremodule{standard}{W}
  \platform{Mac}
\modulesynopsis{Widgets for the Mac, built on top of \refmodule{FrameWork}.}

The \module{W} widgets are used extensively in the \program{IDE}.

\section{\module{waste} --- non-Apple \program{TextEdit} replacement}
\declaremodule{standard}{waste}
  \platform{Mac}
\modulesynopsis{Interface to the ``WorldScript-Aware Styled Text Engine.''}

\begin{seealso}
  \seetitle[http://www.merzwaren.com/waste/]{About WASTE}{Information
            about the WASTE widget and library, including
            documentation and downloads.}
\end{seealso}

