\chapter{Introduction\label{introduction}}

This reference manual describes the Python programming language.
It is not intended as a tutorial.

While I am trying to be as precise as possible, I chose to use English
rather than formal specifications for everything except syntax and
lexical analysis.  This should make the document more understandable
to the average reader, but will leave room for ambiguities.
Consequently, if you were coming from Mars and tried to re-implement
Python from this document alone, you might have to guess things and in
fact you would probably end up implementing quite a different language.
On the other hand, if you are using
Python and wonder what the precise rules about a particular area of
the language are, you should definitely be able to find them here.
If you would like to see a more formal definition of the language,
maybe you could volunteer your time --- or invent a cloning machine
:-).

It is dangerous to add too many implementation details to a language
reference document --- the implementation may change, and other
implementations of the same language may work differently.  On the
other hand, there is currently only one Python implementation in
widespread use (although alternate implementations exist), and
its particular quirks are sometimes worth being mentioned, especially
where the implementation imposes additional limitations.  Therefore,
you'll find short ``implementation notes'' sprinkled throughout the
text.

Every Python implementation comes with a number of built-in and
standard modules.  These are not documented here, but in the separate
\citetitle[../lib/lib.html]{Python Library Reference} document.  A few
built-in modules are mentioned when they interact in a significant way
with the language definition.


\section{Alternate Implementations\label{implementations}}

Though there is one Python implementation which is by far the most
popular, there are some alternate implementations which are of
particular interest to different audiences.

Known implementations include:

\begin{itemize}
\item[CPython]
This is the original and most-maintained implementation of Python,
written in C.  New language features generally appear here first.

\item[Jython]
Python implemented in Java.  This implementation can be used as a
scripting language for Java applications, or can be used to create
applications using the Java class libraries.  It is also often used to
create tests for Java libraries.  More information can be found at
\ulink{the Jython website}{http://www.jython.org/}.

\item[Python for .NET]
This implementation actually uses the CPython implementation, but is a
managed .NET application and makes .NET libraries available.  This was
created by Brian Lloyd.  For more information, see the \ulink{Python
for .NET home page}{http://www.zope.org/Members/Brian/PythonNet}.

\item[IronPython]
An alternate Python for\ .NET.  Unlike Python.NET, this is a complete
Python implementation that generates IL, and compiles Python code
directly to\ .NET assemblies.  It was created by Jim Hugunin, the
original creator of Jython.  For more information, see \ulink{the
IronPython website}{http://workspaces.gotdotnet.com/ironpython}.

\item[PyPy]
An implementation of Python written in Python; even the bytecode
interpreter is written in Python.  This is executed using CPython as
the underlying interpreter.  One of the goals of the project is to
encourage experimentation with the language itself by making it easier
to modify the interpreter (since it is written in Python).  Additional
information is available on \ulink{the PyPy project's home
page}{http://codespeak.net/pypy/}.
\end{itemize}

Each of these implementations varies in some way from the language as
documented in this manual, or introduces specific information beyond
what's covered in the standard Python documentation.  Please refer to
the implementation-specific documentation to determine what else you
need to know about the specific implementation you're using.


\section{Notation\label{notation}}

The descriptions of lexical analysis and syntax use a modified BNF
grammar notation.  This uses the following style of definition:
\index{BNF}
\index{grammar}
\index{syntax}
\index{notation}

\begin{productionlist}
  \production{name}{\token{lc_letter} (\token{lc_letter} | "_")*}
  \production{lc_letter}{"a"..."z"}
\end{productionlist}

The first line says that a \code{name} is an \code{lc_letter} followed by
a sequence of zero or more \code{lc_letter}s and underscores.  An
\code{lc_letter} in turn is any of the single characters \character{a}
through \character{z}.  (This rule is actually adhered to for the
names defined in lexical and grammar rules in this document.)

Each rule begins with a name (which is the name defined by the rule)
and \code{::=}.  A vertical bar (\code{|}) is used to separate
alternatives; it is the least binding operator in this notation.  A
star (\code{*}) means zero or more repetitions of the preceding item;
likewise, a plus (\code{+}) means one or more repetitions, and a
phrase enclosed in square brackets (\code{[ ]}) means zero or one
occurrences (in other words, the enclosed phrase is optional).  The
\code{*} and \code{+} operators bind as tightly as possible;
parentheses are used for grouping.  Literal strings are enclosed in
quotes.  White space is only meaningful to separate tokens.
Rules are normally contained on a single line; rules with many
alternatives may be formatted alternatively with each line after the
first beginning with a vertical bar.

In lexical definitions (as the example above), two more conventions
are used: Two literal characters separated by three dots mean a choice
of any single character in the given (inclusive) range of \ASCII{}
characters.  A phrase between angular brackets (\code{<...>}) gives an
informal description of the symbol defined; e.g., this could be used
to describe the notion of `control character' if needed.
\index{lexical definitions}
\index{ASCII@\ASCII}

Even though the notation used is almost the same, there is a big
difference between the meaning of lexical and syntactic definitions:
a lexical definition operates on the individual characters of the
input source, while a syntax definition operates on the stream of
tokens generated by the lexical analysis.  All uses of BNF in the next
chapter (``Lexical Analysis'') are lexical definitions; uses in
subsequent chapters are syntactic definitions.
