\chapter{Simple statements \label{simple}}
\indexii{simple}{statement}

Simple statements are comprised within a single logical line.
Several simple statements may occur on a single line separated
by semicolons.  The syntax for simple statements is:

\begin{productionlist}
  \production{simple_stmt}{\token{expression_stmt}}
  \productioncont{| \token{assert_stmt}}
  \productioncont{| \token{assignment_stmt}}
  \productioncont{| \token{augmented_assignment_stmt}}
  \productioncont{| \token{pass_stmt}}
  \productioncont{| \token{del_stmt}}
  \productioncont{| \token{print_stmt}}
  \productioncont{| \token{return_stmt}}
  \productioncont{| \token{yield_stmt}}
  \productioncont{| \token{raise_stmt}}
  \productioncont{| \token{break_stmt}}
  \productioncont{| \token{continue_stmt}}
  \productioncont{| \token{import_stmt}}
  \productioncont{| \token{global_stmt}}
  \productioncont{| \token{exec_stmt}}
\end{productionlist}


\section{Expression statements \label{exprstmts}}
\indexii{expression}{statement}

Expression statements are used (mostly interactively) to compute and
write a value, or (usually) to call a procedure (a function that
returns no meaningful result; in Python, procedures return the value
\code{None}).  Other uses of expression statements are allowed and
occasionally useful.  The syntax for an expression statement is:

\begin{productionlist}
  \production{expression_stmt}
             {\token{expression_list}}
\end{productionlist}

An expression statement evaluates the expression list (which may be a
single expression).
\indexii{expression}{list}

In interactive mode, if the value is not \code{None}, it is converted
to a string using the built-in \function{repr()}\bifuncindex{repr}
function and the resulting string is written to standard output (see
section~\ref{print}) on a line by itself.  (Expression statements
yielding \code{None} are not written, so that procedure calls do not
cause any output.)
\obindex{None}
\indexii{string}{conversion}
\index{output}
\indexii{standard}{output}
\indexii{writing}{values}
\indexii{procedure}{call}


\section{Assert statements \label{assert}}

Assert statements\stindex{assert} are a convenient way to insert
debugging assertions\indexii{debugging}{assertions} into a program:

\begin{productionlist}
  \production{assert_stmt}
             {"assert" \token{expression} ["," \token{expression}]}
\end{productionlist}

The simple form, \samp{assert expression}, is equivalent to

\begin{verbatim}
if __debug__:
   if not expression: raise AssertionError
\end{verbatim}

The extended form, \samp{assert expression1, expression2}, is
equivalent to

\begin{verbatim}
if __debug__:
   if not expression1: raise AssertionError, expression2
\end{verbatim}

These equivalences assume that \code{__debug__}\ttindex{__debug__} and
\exception{AssertionError}\exindex{AssertionError} refer to the built-in
variables with those names.  In the current implementation, the
built-in variable \code{__debug__} is \code{True} under normal
circumstances, \code{False} when optimization is requested (command line
option -O).  The current code generator emits no code for an assert
statement when optimization is requested at compile time.  Note that it
is unnecessary to include the source code for the expression that failed
in the error message;
it will be displayed as part of the stack trace.

Assignments to \code{__debug__} are illegal.  The value for the
built-in variable is determined when the interpreter starts.


\section{Assignment statements \label{assignment}}

Assignment statements\indexii{assignment}{statement} are used to
(re)bind names to values and to modify attributes or items of mutable
objects:
\indexii{binding}{name}
\indexii{rebinding}{name}
\obindex{mutable}
\indexii{attribute}{assignment}

\begin{productionlist}
  \production{assignment_stmt}
             {(\token{target_list} "=")+ \token{expression_list}}
  \production{target_list}
             {\token{target} ("," \token{target})* [","]}
  \production{target}
             {\token{identifier}}
  \productioncont{| "(" \token{target_list} ")"}
  \productioncont{| "[" \token{target_list} "]"}
  \productioncont{| \token{attributeref}}
  \productioncont{| \token{subscription}}
  \productioncont{| \token{slicing}}
\end{productionlist}

(See section~\ref{primaries} for the syntax definitions for the last
three symbols.)

An assignment statement evaluates the expression list (remember that
this can be a single expression or a comma-separated list, the latter
yielding a tuple) and assigns the single resulting object to each of
the target lists, from left to right.
\indexii{expression}{list}

Assignment is defined recursively depending on the form of the target
(list).  When a target is part of a mutable object (an attribute
reference, subscription or slicing), the mutable object must
ultimately perform the assignment and decide about its validity, and
may raise an exception if the assignment is unacceptable.  The rules
observed by various types and the exceptions raised are given with the
definition of the object types (see section~\ref{types}).
\index{target}
\indexii{target}{list}

Assignment of an object to a target list is recursively defined as
follows.
\indexiii{target}{list}{assignment}

\begin{itemize}
\item
If the target list is a single target: The object is assigned to that
target.

\item
If the target list is a comma-separated list of targets: The object
must be a sequence with the same number of items as there are
targets in the target list, and the items are assigned, from left to
right, to the corresponding targets.  (This rule is relaxed as of
Python 1.5; in earlier versions, the object had to be a tuple.  Since
strings are sequences, an assignment like \samp{a, b = "xy"} is
now legal as long as the string has the right length.)

\end{itemize}

Assignment of an object to a single target is recursively defined as
follows.

\begin{itemize} % nested

\item
If the target is an identifier (name):

\begin{itemize}

\item
If the name does not occur in a \keyword{global} statement in the current
code block: the name is bound to the object in the current local
namespace.
\stindex{global}

\item
Otherwise: the name is bound to the object in the current global
namespace.

\end{itemize} % nested

The name is rebound if it was already bound.  This may cause the
reference count for the object previously bound to the name to reach
zero, causing the object to be deallocated and its
destructor\index{destructor} (if it has one) to be called.

\item
If the target is a target list enclosed in parentheses or in square
brackets: The object must be a sequence with the same number of items
as there are targets in the target list, and its items are assigned,
from left to right, to the corresponding targets.

\item
If the target is an attribute reference: The primary expression in the
reference is evaluated.  It should yield an object with assignable
attributes; if this is not the case, \exception{TypeError} is raised.  That
object is then asked to assign the assigned object to the given
attribute; if it cannot perform the assignment, it raises an exception
(usually but not necessarily \exception{AttributeError}).
\indexii{attribute}{assignment}

\item
If the target is a subscription: The primary expression in the
reference is evaluated.  It should yield either a mutable sequence
object (e.g., a list) or a mapping object (e.g., a dictionary).  Next,
the subscript expression is evaluated.
\indexii{subscription}{assignment}
\obindex{mutable}

If the primary is a mutable sequence object (e.g., a list), the subscript
must yield a plain integer.  If it is negative, the sequence's length
is added to it.  The resulting value must be a nonnegative integer
less than the sequence's length, and the sequence is asked to assign
the assigned object to its item with that index.  If the index is out
of range, \exception{IndexError} is raised (assignment to a subscripted
sequence cannot add new items to a list).
\obindex{sequence}
\obindex{list}

If the primary is a mapping object (e.g., a dictionary), the subscript must
have a type compatible with the mapping's key type, and the mapping is
then asked to create a key/datum pair which maps the subscript to
the assigned object.  This can either replace an existing key/value
pair with the same key value, or insert a new key/value pair (if no
key with the same value existed).
\obindex{mapping}
\obindex{dictionary}

\item
If the target is a slicing: The primary expression in the reference is
evaluated.  It should yield a mutable sequence object (e.g., a list).  The
assigned object should be a sequence object of the same type.  Next,
the lower and upper bound expressions are evaluated, insofar they are
present; defaults are zero and the sequence's length.  The bounds
should evaluate to (small) integers.  If either bound is negative, the
sequence's length is added to it.  The resulting bounds are clipped to
lie between zero and the sequence's length, inclusive.  Finally, the
sequence object is asked to replace the slice with the items of the
assigned sequence.  The length of the slice may be different from the
length of the assigned sequence, thus changing the length of the
target sequence, if the object allows it.
\indexii{slicing}{assignment}

\end{itemize}
        
(In the current implementation, the syntax for targets is taken
to be the same as for expressions, and invalid syntax is rejected
during the code generation phase, causing less detailed error
messages.)

WARNING: Although the definition of assignment implies that overlaps
between the left-hand side and the right-hand side are `safe' (e.g.,
\samp{a, b = b, a} swaps two variables), overlaps \emph{within} the
collection of assigned-to variables are not safe!  For instance, the
following program prints \samp{[0, 2]}:

\begin{verbatim}
x = [0, 1]
i = 0
i, x[i] = 1, 2
print x
\end{verbatim}


\subsection{Augmented assignment statements \label{augassign}}

Augmented assignment is the combination, in a single statement, of a binary
operation and an assignment statement:
\indexii{augmented}{assignment}
\index{statement!assignment, augmented}

\begin{productionlist}
  \production{augmented_assignment_stmt}
             {\token{target} \token{augop} \token{expression_list}}
  \production{augop}
             {"+=" | "-=" | "*=" | "/=" | "\%=" | "**="}
  % The empty groups below prevent conversion to guillemets.
  \productioncont{| ">{}>=" | "<{}<=" | "\&=" | "\textasciicircum=" | "|="}
\end{productionlist}

(See section~\ref{primaries} for the syntax definitions for the last
three symbols.)

An augmented assignment evaluates the target (which, unlike normal
assignment statements, cannot be an unpacking) and the expression
list, performs the binary operation specific to the type of assignment
on the two operands, and assigns the result to the original
target.  The target is only evaluated once.

An augmented assignment expression like \code{x += 1} can be rewritten as
\code{x = x + 1} to achieve a similar, but not exactly equal effect. In the
augmented version, \code{x} is only evaluated once. Also, when possible, the
actual operation is performed \emph{in-place}, meaning that rather than
creating a new object and assigning that to the target, the old object is
modified instead.

With the exception of assigning to tuples and multiple targets in a single
statement, the assignment done by augmented assignment statements is handled
the same way as normal assignments. Similarly, with the exception of the
possible \emph{in-place} behavior, the binary operation performed by
augmented assignment is the same as the normal binary operations.

For targets which are attribute references, the initial value is
retrieved with a \method{getattr()} and the result is assigned with a
\method{setattr()}.  Notice that the two methods do not necessarily
refer to the same variable.  When \method{getattr()} refers to a class
variable, \method{setattr()} still writes to an instance variable.
For example:

\begin{verbatim}
class A:
    x = 3    # class variable
a = A()
a.x += 1     # writes a.x as 4 leaving A.x as 3
\end{verbatim}


\section{The \keyword{pass} statement \label{pass}}
\stindex{pass}

\begin{productionlist}
  \production{pass_stmt}
             {"pass"}
\end{productionlist}

\keyword{pass} is a null operation --- when it is executed, nothing
happens.  It is useful as a placeholder when a statement is
required syntactically, but no code needs to be executed, for example:
\indexii{null}{operation}

\begin{verbatim}
def f(arg): pass    # a function that does nothing (yet)

class C: pass       # a class with no methods (yet)
\end{verbatim}


\section{The \keyword{del} statement \label{del}}
\stindex{del}

\begin{productionlist}
  \production{del_stmt}
             {"del" \token{target_list}}
\end{productionlist}

Deletion is recursively defined very similar to the way assignment is
defined. Rather that spelling it out in full details, here are some
hints.
\indexii{deletion}{target}
\indexiii{deletion}{target}{list}

Deletion of a target list recursively deletes each target, from left
to right.

Deletion of a name removes the binding of that name 
from the local or global namespace, depending on whether the name
occurs in a \keyword{global} statement in the same code block.  If the
name is unbound, a \exception{NameError} exception will be raised.
\stindex{global}
\indexii{unbinding}{name}

It is illegal to delete a name from the local namespace if it occurs
as a free variable\indexii{free}{variable} in a nested block.

Deletion of attribute references, subscriptions and slicings
is passed to the primary object involved; deletion of a slicing
is in general equivalent to assignment of an empty slice of the
right type (but even this is determined by the sliced object).
\indexii{attribute}{deletion}


\section{The \keyword{print} statement \label{print}}
\stindex{print}

\begin{productionlist}
  \production{print_stmt}
             {"print" ( \optional{\token{expression} ("," \token{expression})* \optional{","}}}
  \productioncont{| ">\code{>}" \token{expression}
                  \optional{("," \token{expression})+ \optional{","}} )}
\end{productionlist}

\keyword{print} evaluates each expression in turn and writes the
resulting object to standard output (see below).  If an object is not
a string, it is first converted to a string using the rules for string
conversions.  The (resulting or original) string is then written.  A
space is written before each object is (converted and) written, unless
the output system believes it is positioned at the beginning of a
line.  This is the case (1) when no characters have yet been written
to standard output, (2) when the last character written to standard
output is \character{\e n}, or (3) when the last write operation on
standard output was not a \keyword{print} statement.  (In some cases
it may be functional to write an empty string to standard output for
this reason.)  \note{Objects which act like file objects but which are
not the built-in file objects often do not properly emulate this
aspect of the file object's behavior, so it is best not to rely on
this.}
\index{output}
\indexii{writing}{values}

A \character{\e n} character is written at the end, unless the
\keyword{print} statement ends with a comma.  This is the only action
if the statement contains just the keyword \keyword{print}.
\indexii{trailing}{comma}
\indexii{newline}{suppression}

Standard output is defined as the file object named \code{stdout}
in the built-in module \module{sys}.  If no such object exists, or if
it does not have a \method{write()} method, a \exception{RuntimeError}
exception is raised.
\indexii{standard}{output}
\refbimodindex{sys}
\withsubitem{(in module sys)}{\ttindex{stdout}}
\exindex{RuntimeError}

\keyword{print} also has an extended\index{extended print statement}
form, defined by the second portion of the syntax described above.
This form is sometimes referred to as ``\keyword{print} chevron.''
In this form, the first expression after the \code{>}\code{>} must
evaluate to a ``file-like'' object, specifically an object that has a
\method{write()} method as described above.  With this extended form,
the subsequent expressions are printed to this file object.  If the
first expression evaluates to \code{None}, then \code{sys.stdout} is
used as the file for output.


\section{The \keyword{return} statement \label{return}}
\stindex{return}

\begin{productionlist}
  \production{return_stmt}
             {"return" [\token{expression_list}]}
\end{productionlist}

\keyword{return} may only occur syntactically nested in a function
definition, not within a nested class definition.
\indexii{function}{definition}
\indexii{class}{definition}

If an expression list is present, it is evaluated, else \code{None}
is substituted.

\keyword{return} leaves the current function call with the expression
list (or \code{None}) as return value.

When \keyword{return} passes control out of a \keyword{try} statement
with a \keyword{finally} clause, that \keyword{finally} clause is executed
before really leaving the function.
\kwindex{finally}

In a generator function, the \keyword{return} statement is not allowed
to include an \grammartoken{expression_list}.  In that context, a bare
\keyword{return} indicates that the generator is done and will cause
\exception{StopIteration} to be raised.


\section{The \keyword{yield} statement \label{yield}}
\stindex{yield}

\begin{productionlist}
  \production{yield_stmt}
             {"yield" \token{expression_list}}
\end{productionlist}

\index{generator!function}
\index{generator!iterator}
\index{function!generator}
\exindex{StopIteration}

The \keyword{yield} statement is only used when defining a generator
function, and is only used in the body of the generator function.
Using a \keyword{yield} statement in a function definition is
sufficient to cause that definition to create a generator function
instead of a normal function.

When a generator function is called, it returns an iterator known as a
generator iterator, or more commonly, a generator.  The body of the
generator function is executed by calling the generator's
\method{next()} method repeatedly until it raises an exception.

When a \keyword{yield} statement is executed, the state of the
generator is frozen and the value of \grammartoken{expression_list} is
returned to \method{next()}'s caller.  By ``frozen'' we mean that all
local state is retained, including the current bindings of local
variables, the instruction pointer, and the internal evaluation stack:
enough information is saved so that the next time \method{next()} is
invoked, the function can proceed exactly as if the \keyword{yield}
statement were just another external call.

The \keyword{yield} statement is not allowed in the \keyword{try}
clause of a \keyword{try} ...\ \keyword{finally} construct.  The
difficulty is that there's no guarantee the generator will ever be
resumed, hence no guarantee that the \keyword{finally} block will ever
get executed.

\begin{notice}
In Python 2.2, the \keyword{yield} statement is only allowed
when the \code{generators} feature has been enabled.  It will always
be enabled in Python 2.3.  This \code{__future__} import statement can
be used to enable the feature:

\begin{verbatim}
from __future__ import generators
\end{verbatim}
\end{notice}


\begin{seealso}
  \seepep{0255}{Simple Generators}
         {The proposal for adding generators and the \keyword{yield}
          statement to Python.}
\end{seealso}


\section{The \keyword{raise} statement \label{raise}}
\stindex{raise}

\begin{productionlist}
  \production{raise_stmt}
             {"raise" [\token{expression} ["," \token{expression}
              ["," \token{expression}]]]}
\end{productionlist}

If no expressions are present, \keyword{raise} re-raises the last
expression that was active in the current scope.  If no exception is
active in the current scope, an exception is raised indicating this error.
\index{exception}
\indexii{raising}{exception}

Otherwise, \keyword{raise} evaluates the expressions to get three
objects, using \code{None} as the value of omitted expressions.  The
first two objects are used to determine the \emph{type} and
\emph{value} of the exception.

If the first object is an instance, the type of the exception is the
class of the instance, the instance itself is the value, and the
second object must be \code{None}.

If the first object is a class, it becomes the type of the exception.
The second object is used to determine the exception value: If it is
an instance of the class, the instance becomes the exception value.
If the second object is a tuple, it is used as the argument list for
the class constructor; if it is \code{None}, an empty argument list is
used, and any other object is treated as a single argument to the
constructor.  The instance so created by calling the constructor is
used as the exception value.

If a third object is present and not \code{None}, it must be a
traceback\obindex{traceback} object (see section~\ref{traceback}), and
it is substituted instead of the current location as the place where
the exception occurred.  If the third object is present and not a
traceback object or \code{None}, a \exception{TypeError} exception is
raised.  The three-expression form of \keyword{raise} is useful to
re-raise an exception transparently in an except clause, but
\keyword{raise} with no expressions should be preferred if the
exception to be re-raised was the most recently active exception in
the current scope.

Additional information on exceptions can be found in
section~\ref{exceptions}, and information about handling exceptions is
in section~\ref{try}.


\section{The \keyword{break} statement \label{break}}
\stindex{break}

\begin{productionlist}
  \production{break_stmt}
             {"break"}
\end{productionlist}

\keyword{break} may only occur syntactically nested in a \keyword{for}
or \keyword{while} loop, but not nested in a function or class definition
within that loop.
\stindex{for}
\stindex{while}
\indexii{loop}{statement}

It terminates the nearest enclosing loop, skipping the optional
\keyword{else} clause if the loop has one.
\kwindex{else}

If a \keyword{for} loop is terminated by \keyword{break}, the loop control
target keeps its current value.
\indexii{loop control}{target}

When \keyword{break} passes control out of a \keyword{try} statement
with a \keyword{finally} clause, that \keyword{finally} clause is executed
before really leaving the loop.
\kwindex{finally}


\section{The \keyword{continue} statement \label{continue}}
\stindex{continue}

\begin{productionlist}
  \production{continue_stmt}
             {"continue"}
\end{productionlist}

\keyword{continue} may only occur syntactically nested in a \keyword{for} or
\keyword{while} loop, but not nested in a function or class definition or
\keyword{try} statement within that loop.\footnote{It may
occur within an \keyword{except} or \keyword{else} clause.  The
restriction on occurring in the \keyword{try} clause is implementor's
laziness and will eventually be lifted.}
It continues with the next cycle of the nearest enclosing loop.
\stindex{for}
\stindex{while}
\indexii{loop}{statement}
\kwindex{finally}


\section{The \keyword{import} statement \label{import}}
\stindex{import}
\index{module!importing}
\indexii{name}{binding}
\kwindex{from}

\begin{productionlist}
  \production{import_stmt}
             {"import" \token{module} ["as" \token{name}]
                ( "," \token{module} ["as" \token{name}] )*}
  \productioncont{| "from" \token{module} "import" \token{identifier}
                    ["as" \token{name}]}
  \productioncont{  ( "," \token{identifier} ["as" \token{name}] )*}
  \productioncont{| "from" \token{module} "import" "(" \token{identifier}
                    ["as" \token{name}]}
  \productioncont{  ( "," \token{identifier} ["as" \token{name}] )* [","] ")"}
  \productioncont{| "from" \token{module} "import" "*"}
  \production{module}
             {(\token{identifier} ".")* \token{identifier}}
\end{productionlist}

Import statements are executed in two steps: (1) find a module, and
initialize it if necessary; (2) define a name or names in the local
namespace (of the scope where the \keyword{import} statement occurs).
The first form (without \keyword{from}) repeats these steps for each
identifier in the list.  The form with \keyword{from} performs step
(1) once, and then performs step (2) repeatedly.

In this context, to ``initialize'' a built-in or extension module means to
call an initialization function that the module must provide for the purpose
(in the reference implementation, the function's name is obtained by
prepending string ``init'' to the module's name); to ``initialize'' a
Python-coded module means to execute the module's body.
  
The system maintains a table of modules that have been or are being
initialized,
indexed by module name.  This table is
accessible as \code{sys.modules}.  When a module name is found in
this table, step (1) is finished.  If not, a search for a module
definition is started.  When a module is found, it is loaded.  Details
of the module searching and loading process are implementation and
platform specific.  It generally involves searching for a ``built-in''
module with the given name and then searching a list of locations
given as \code{sys.path}.
\withsubitem{(in module sys)}{\ttindex{modules}}
\ttindex{sys.modules}
\indexii{module}{name}
\indexii{built-in}{module}
\indexii{user-defined}{module}
\refbimodindex{sys}
\indexii{filename}{extension}
\indexiii{module}{search}{path}

If a built-in module is found,\indexii{module}{initialization} its
built-in initialization code is executed and step (1) is finished.  If
no matching file is found,
\exception{ImportError}\exindex{ImportError} is raised.
\index{code block}If a file is found, it is parsed,
yielding an executable code block.  If a syntax error occurs,
\exception{SyntaxError}\exindex{SyntaxError} is raised.  Otherwise, an
empty module of the given name is created and inserted in the module
table, and then the code block is executed in the context of this
module.  Exceptions during this execution terminate step (1).

When step (1) finishes without raising an exception, step (2) can
begin.

The first form of \keyword{import} statement binds the module name in
the local namespace to the module object, and then goes on to import
the next identifier, if any.  If the module name is followed by
\keyword{as}, the name following \keyword{as} is used as the local
name for the module. 

The \keyword{from} form does not bind the module name: it goes through the
list of identifiers, looks each one of them up in the module found in step
(1), and binds the name in the local namespace to the object thus found. 
As with the first form of \keyword{import}, an alternate local name can be
supplied by specifying "\keyword{as} localname".  If a name is not found,
\exception{ImportError} is raised.  If the list of identifiers is replaced
by a star (\character{*}), all public names defined in the module are
bound in the local namespace of the \keyword{import} statement..
\indexii{name}{binding}
\exindex{ImportError}

The \emph{public names} defined by a module are determined by checking
the module's namespace for a variable named \code{__all__}; if
defined, it must be a sequence of strings which are names defined or
imported by that module.  The names given in \code{__all__} are all
considered public and are required to exist.  If \code{__all__} is not
defined, the set of public names includes all names found in the
module's namespace which do not begin with an underscore character
(\character{_}).  \code{__all__} should contain the entire public API.
It is intended to avoid accidentally exporting items that are not part
of the API (such as library modules which were imported and used within
the module).
\withsubitem{(optional module attribute)}{\ttindex{__all__}}

The \keyword{from} form with \samp{*} may only occur in a module
scope.  If the wild card form of import --- \samp{import *} --- is
used in a function and the function contains or is a nested block with
free variables, the compiler will raise a \exception{SyntaxError}.

\kwindex{from}
\stindex{from}

\strong{Hierarchical module names:}\indexiii{hierarchical}{module}{names}
when the module names contains one or more dots, the module search
path is carried out differently.  The sequence of identifiers up to
the last dot is used to find a ``package''\index{packages}; the final
identifier is then searched inside the package.  A package is
generally a subdirectory of a directory on \code{sys.path} that has a
file \file{__init__.py}.\ttindex{__init__.py}
%
[XXX Can't be bothered to spell this out right now; see the URL
\url{http://www.python.org/doc/essays/packages.html} for more details, also
about how the module search works from inside a package.]

The built-in function \function{__import__()} is provided to support
applications that determine which modules need to be loaded
dynamically; refer to \ulink{Built-in
Functions}{../lib/built-in-funcs.html} in the
\citetitle[../lib/lib.html]{Python Library Reference} for additional
information.
\bifuncindex{__import__}

\subsection{Future statements \label{future}}

A \dfn{future statement}\indexii{future}{statement} is a directive to
the compiler that a particular module should be compiled using syntax
or semantics that will be available in a specified future release of
Python.  The future statement is intended to ease migration to future
versions of Python that introduce incompatible changes to the
language.  It allows use of the new features on a per-module basis
before the release in which the feature becomes standard.

\begin{productionlist}[*]
  \production{future_statement}
             {"from" "__future__" "import" feature ["as" name] ("," feature ["as" name])*}
  \productioncont{| "from" "__future__" "import" "(" feature ["as" name] ("," feature ["as" name])* [","] ")"}
  \production{feature}{identifier}
  \production{name}{identifier}
\end{productionlist}

A future statement must appear near the top of the module.  The only
lines that can appear before a future statement are:

\begin{itemize}

\item the module docstring (if any),
\item comments,
\item blank lines, and
\item other future statements.

\end{itemize}

The features recognized by Python 2.3 are \samp{generators},
\samp{division} and \samp{nested_scopes}.  \samp{generators} and
\samp{nested_scopes} are redundant in 2.3 because they are always
enabled. 

A future statement is recognized and treated specially at compile
time: Changes to the semantics of core constructs are often
implemented by generating different code.  It may even be the case
that a new feature introduces new incompatible syntax (such as a new
reserved word), in which case the compiler may need to parse the
module differently.  Such decisions cannot be pushed off until
runtime.

For any given release, the compiler knows which feature names have been
defined, and raises a compile-time error if a future statement contains
a feature not known to it.

The direct runtime semantics are the same as for any import statement:
there is a standard module \module{__future__}, described later, and
it will be imported in the usual way at the time the future statement
is executed.

The interesting runtime semantics depend on the specific feature
enabled by the future statement.

Note that there is nothing special about the statement:

\begin{verbatim}
import __future__ [as name]
\end{verbatim}

That is not a future statement; it's an ordinary import statement with
no special semantics or syntax restrictions.

Code compiled by an exec statement or calls to the builtin functions
\function{compile()} and \function{execfile()} that occur in a module
\module{M} containing a future statement will, by default, use the new 
syntax or semantics associated with the future statement.  This can,
starting with Python 2.2 be controlled by optional arguments to
\function{compile()} --- see the documentation of that function in the 
library reference for details.

A future statement typed at an interactive interpreter prompt will
take effect for the rest of the interpreter session.  If an
interpreter is started with the \programopt{-i} option, is passed a
script name to execute, and the script includes a future statement, it
will be in effect in the interactive session started after the script
is executed.

\section{The \keyword{global} statement \label{global}}
\stindex{global}

\begin{productionlist}
  \production{global_stmt}
             {"global" \token{identifier} ("," \token{identifier})*}
\end{productionlist}

The \keyword{global} statement is a declaration which holds for the
entire current code block.  It means that the listed identifiers are to be
interpreted as globals.  It would be impossible to assign to a global
variable without \keyword{global}, although free variables may refer
to globals without being declared global.
\indexiii{global}{name}{binding}

Names listed in a \keyword{global} statement must not be used in the same
code block textually preceding that \keyword{global} statement.

Names listed in a \keyword{global} statement must not be defined as formal
parameters or in a \keyword{for} loop control target, \keyword{class}
definition, function definition, or \keyword{import} statement.

(The current implementation does not enforce the latter two
restrictions, but programs should not abuse this freedom, as future
implementations may enforce them or silently change the meaning of the
program.)

\strong{Programmer's note:}
the \keyword{global} is a directive to the parser.  It
applies only to code parsed at the same time as the \keyword{global}
statement.  In particular, a \keyword{global} statement contained in an
\keyword{exec} statement does not affect the code block \emph{containing}
the \keyword{exec} statement, and code contained in an \keyword{exec}
statement is unaffected by \keyword{global} statements in the code
containing the \keyword{exec} statement.  The same applies to the
\function{eval()}, \function{execfile()} and \function{compile()} functions.
\stindex{exec}
\bifuncindex{eval}
\bifuncindex{execfile}
\bifuncindex{compile}


\section{The \keyword{exec} statement \label{exec}}
\stindex{exec}

\begin{productionlist}
  \production{exec_stmt}
             {"exec" \token{expression}
              ["in" \token{expression} ["," \token{expression}]]}
\end{productionlist}

This statement supports dynamic execution of Python code.  The first
expression should evaluate to either a string, an open file object, or
a code object.  If it is a string, the string is parsed as a suite of
Python statements which is then executed (unless a syntax error
occurs).  If it is an open file, the file is parsed until \EOF{} and
executed.  If it is a code object, it is simply executed.

In all cases, if the optional parts are omitted, the code is executed
in the current scope.  If only the first expression after \keyword{in}
is specified, it should be a dictionary, which will be used for both
the global and the local variables.  If two expressions are given,
they are used for the global and local variables, respectively.
If provided, \var{locals} can be any mapping object.
\versionchanged[formerly \var{locals} was required to be a dictionary]{2.4}

As a side effect, an implementation may insert additional keys into
the dictionaries given besides those corresponding to variable names
set by the executed code.  For example, the current implementation
may add a reference to the dictionary of the built-in module
\module{__builtin__} under the key \code{__builtins__} (!).
\ttindex{__builtins__}
\refbimodindex{__builtin__}

\strong{Programmer's hints:}
dynamic evaluation of expressions is supported by the built-in
function \function{eval()}.  The built-in functions
\function{globals()} and \function{locals()} return the current global
and local dictionary, respectively, which may be useful to pass around
for use by \keyword{exec}.
\bifuncindex{eval}
\bifuncindex{globals}
\bifuncindex{locals}

  

