\section{Standard Module \sectcode{cgi}}
\stmodindex{cgi}
\indexii{WWW}{server}
\indexii{CGI}{protocol}
\indexii{HTTP}{protocol}
\indexii{MIME}{headers}
\index{URL}

\renewcommand{\indexsubitem}{(in module cgi)}

This module makes it easy to write Python scripts that run in a WWW
server using the Common Gateway Interface.  It was written by Michael
McLay and subsequently modified by Steve Majewski and Guido van
Rossum.

When a WWW server finds that a URL contains a reference to a file in a
particular subdirectory (usually \code{/cgibin}), it runs the file as
a subprocess.  Information about the request such as the full URL, the
originating host etc., is passed to the subprocess in the shell
environment; additional input from the client may be read from
standard input.  Standard output from the subprocess is sent back
across the network to the client as the response from the request.
The CGI protocol describes what the environment variables passed to
the subprocess mean and how the output should be formatted.  The
official reference documentation for the CGI protocol can be found on
the World-Wide Web at
\code{<URL:http://hoohoo.ncsa.uiuc.edu/cgi/overview.html>}.  The
\code{cgi} module was based on version 1.1 of the protocol and should
also work with version 1.0.

The \code{cgi} module defines several classes that make it easy to
access the information passed to the subprocess from a Python script;
in particular, it knows how to parse the input sent by an HTML
``form'' using either a POST or a GET request (these are alternatives
for submitting forms in the HTTP protocol).

The formatting of the output is so trivial that no additional support
is needed.  All you need to do is print a minimal set of MIME headers
describing the output format, followed by a blank line and your actual
output.  E.g. if you want to generate HTML, your script could start as
follows:

\begin{verbatim}
# Header -- one or more lines:
print "Content-type: text/html"
# Blank line separating header from body:
print
# Body, in HTML format:
print "<TITLE>The Amazing SPAM Homepage!</TITLE>"
# etc...
\end{verbatim}

The server will add some header lines of its own, but it won't touch
the output following the header.

The \code{cgi} module defines the following functions:

\begin{funcdesc}{parse}{}
Read and parse the form submitted to the script and return a
dictionary containing the form's fields.  This should be called at
most once per script invocation, as it may consume standard input (if
the form was submitted through a POST request).  The keys in the
resulting dictionary are the field names used in the submission; the
values are {\em lists} of the field values (since field name may be
used multiple times in a single form).  \samp{\%} escapes in the
values are translated to their single-character equivalent using
\code{urllib.unquote()}.  As a side effect, this function sets
\code{environ['QUERY_STRING']} to the raw query string, if it isn't
already set.
\end{funcdesc}

\begin{funcdesc}{print_environ_usage}{}
Print a piece of HTML listing the environment variables that may be
set by the CGI protocol.
This is mainly useful when learning about writing CGI scripts.
\end{funcdesc}

\begin{funcdesc}{print_environ}{}
Print a piece of HTML text showing the entire contents of the shell
environment.  This is mainly useful when debugging a CGI script.
\end{funcdesc}

\begin{funcdesc}{print_form}{form}
Print a piece of HTML text showing the contents of the \var{form} (a
dictionary, an instance of the \code{FormContentDict} class defined
below, or a subclass thereof).
This is mainly useful when debugging a CGI script.
\end{funcdesc}

\begin{funcdesc}{escape}{string}
Convert special characters in \var{string} to HTML escapes.  In
particular, ``\code{\&}'' is replaced with ``\code{\&amp;}'',
``\code{<}'' is replaced with ``\code{\&lt;}'', and ``\code{>}'' is
replaced with ``\code{\&gt;}''.  This is useful when printing (almost)
arbitrary text in an HTML context.  Note that for inclusion in quoted
tag attributes (e.g. \code{<A HREF="...">}), some additional
characters would have to be converted --- in particular the string
quote.  There is currently no function that does this.
\end{funcdesc}

The module defines the following classes.  Since the base class
initializes itself by calling \code{parse()}, at most one instance of
at most one of these classes should be created per script invocation:

\begin{funcdesc}{FormContentDict}{}
This class behaves like a (read-only) dictionary and has the same keys
and values as the dictionary returned by \code{parse()} (i.e. each
field name maps to a list of values).  Additionally, it initializes
its data member \code{query_string} to the raw query sent from the
server.
\end{funcdesc}

\begin{funcdesc}{SvFormContentDict}{}
This class, derived from \code{FormContentDict}, is a little more
user-friendly when you are expecting that each field name is only used
once in the form.  When you access for a particular field (using
\code{form[fieldname]}), it will return the string value of that item
if it is unique, or raise \code{IndexError} if the field was specified
more than once in the form.  (If the field wasn't specified at all,
\code{KeyError} is raised.)  To access fields that are specified
multiple times, use \code{form.getlist(fieldname)}.  The
\code{values()} and \code{items()} methods return mixed lists ---
containing strings for singly-defined fields, and lists of strings for
multiply-defined fields.
\end{funcdesc}

(It currently defines some more classes, but these are experimental
and/or obsolescent, and are thus not documented --- see the source for
more informations.)

The module defines the following variable:

\begin{datadesc}{environ}
The shell environment, exactly as received from the http server.  See
the CGI documentation for a description of the various fields.
\end{datadesc}

\subsection{Example}

This example assumes that you have a WWW server up and running,
e.g.\ NCSA's \code{httpd}.

Place the following file in a convenient spot in the WWW server's
directory tree.  E.g., if you place it in the subdirectory \file{test}
of the root directory and call it \file{test.html}, its URL will be
\file{http://\var{yourservername}/test/test.html}.

\begin{verbatim}
<TITLE>Test Form Input</TITLE>
<H1>Test Form Input</H1>
<FORM METHOD="POST" ACTION="/cgi-bin/test.py">
<INPUT NAME=Name> (Name)<br>
<INPUT NAME=Address> (Address)<br>
<INPUT TYPE=SUBMIT>
</FORM>
\end{verbatim}

Selecting this file's URL from a forms-capable browser such as Mosaic
or Netscape will bring up a simple form with two text input fields and
a ``submit'' button.

But wait.  Before pressing ``submit'', a script that responds to the
form must also be installed.  The test file as shown assumes that the
script is called \file{test.py} and lives in the server's
\code{cgi-bin} directory.  Here's the test script:

\begin{verbatim}
#!/usr/local/bin/python

import cgi

print "Content-type: text/html"
print                                   # End of headers!
print "<TITLE>Test Form Output</TITLE>"
print "<H1>Test Form Output</H1>"

form = cgi.SvFormContentDict()          # Load the form

name = addr = None                      # Default: no name and address

# Extract name and address from the form, if given

if form.has_key('Name'):
        name = form['Name']
if form.has_key('Address'):
        addr = form['Address']
        
# Print an unnumbered list of the name and address, if present

print "<UL>"
if name is not None:
        print "<LI>Name:", cgi.escape(name)
if addr is not None:
        print "<LI>Address:", cgi.escape(addr)
print "</UL>"
\end{verbatim}

The script should be made executable (\samp{chmod +x \var{script}}).
If the Python interpreter is not located at
\file{/usr/local/bin/python} but somewhere else, the first line of the
script should be modified accordingly.

Now that everything is installed correctly, we can try out the form.
Bring up the test form in your WWW browser, fill in a name and address
in the form, and press the ``submit'' button.  The script should now
run and its output is sent back to your browser.  This should roughly
look as follows:

\strong{Test Form Output}

\begin{itemize}
\item Name: \var{the name you entered}
\item Address: \var{the address you entered}
\end{itemize}

If you didn't enter a name or address, the corresponding line will be
missing (since the browser doesn't send empty form fields to the
server).
