\section{Built-in module \sectcode{pickle}}
\stmodindex{pickle}
\index{persistency}
\indexii{persistent}{objects}
\indexii{serializing}{objects}
\indexii{marshalling}{objects}
\indexii{flattening}{objects}
\indexii{pickling}{objects}

The \code{pickle} module implements a basic but powerful algorithm for
``pickling'' (a.k.a. serializing, marshalling or flattening) nearly
arbitrary Python objects.  This is a more primitive notion than
persistency --- although \code{pickle} reads and writes file objects,
it does not handle the issue of naming persistent objects, nor the
(even more complicated) area of concurrent access to persistent
objects.  The \code{pickle} module can transform a complex object into
a byte stream and it can transform the byte stream into an object with
the same internal structure.  The most obvious thing to do with these
byte streams is to write them onto a file, but it is also conceivable
to send them across a network or store them in a database.  The module
\code{shelve} provides a simple interface to pickle and unpickle
objects on ``dbm''-style database files.
\stmodindex{shelve}

Unlike the built-in module \code{marshal}, \code{pickle} handles the
following correctly:
\stmodindex{marshal}

\begin{itemize}

\item recursive objects

\item pointer sharing

\item instances uf user-defined classes

\end{itemize}

The data format used by \code{pickle} is Python-specific.  This has
the advantage that there are no restrictions imposed by external
standards such as CORBA (which probably can't represent pointer
sharing or recursive objects); however it means that non-Python
programs may not be able to reconstruct pickled Python objects.

The \code{pickle} data format uses a printable ASCII representation.
This is slightly more voluminous than a binary representation.
However, small integers actually take {\em less} space when
represented as minimal-size decimal strings than when represented as
32-bit binary numbers, and strings are only much longer if they
contain many control characters or 8-bit characters.  The big
advantage of using printable ASCII (and of some other characteristics
of \code{pickle}'s representation) is that for debugging or recovery
purposes it is possible for a human to read the pickled file with a
standard text editor.  (I could have gone a step further and used a
notation like S-expressions, but the parser would have been
considerably more complicated and slower, and the files would probably
have become much larger.)

The \code{pickle} module doesn't handle code objects, which the
\code{marshal} module does.  I suppose \code{pickle} could, and maybe
it should, but there's probably no great need for it right now (as
long as \code{marshal} continues to be used for reading and writing
code objects), and at least this avoids the possibility of smuggling
Trojan horses into a program.
\stmodindex{marshal}

For the benefit of persistency modules written using \code{pickle}, it
supports the notion of a reference to an object outside the pickled
data stream.  Such objects are referenced by a name, which is an
arbitrary string of printable ASCII characters.  The resolution of
such names is not defined by the \code{pickle} module --- the
persistent object module will have to implement a method
\code{persistent_load}.  To write references to persistent objects,
the persistent module must define a method \code{persistent_id} which
returns either \code{None} or the persistent ID of the object.

There are some restrictions on the pickling of class instances.

First of all, the class must be defined at the top level in a module.

Next, it must normally be possible to create class instances by
calling the class without arguments.  If this is undesirable, the
class can define a method \code{__getinitargs__()}, which should
return a {\em tuple} containing the arguments to be passed to the
class constructor (\code{__init__()}).
\ttindex{__getinitargs__}
\ttindex{__init__}

Classes can further influence how they are pickled --- if the class
defines the method \code{__getstate__()}, it is called and the return
state is pickled as the contents for the instance, and if the class
defines the method \code{__setstate__()}, it is called with the
unpickled state.  (Note that these methods can also be used to
implement copying class instances.)  If there is no
\code{__getstate__()} method, the instance's \code{__dict__} is
pickled.  If there is no \code{__setstate__()} method, the pickled
object must be a dictionary and its items are assigned to the new
instance's dictionary.  (If a class defines both \code{__getstate__()}
and \code{__setstate__()}, the state object needn't be a dictionary
--- these methods can do what they want.)  This protocol is also used
by the shallow and deep copying operations defined in the \code{copy}
module.
\ttindex{__getstate__}
\ttindex{__setstate__}
\ttindex{__dict__}

Note that when class instances are pickled, their class's code and
data is not pickled along with them.  Only the instance data is
pickled.  This is done on purpose, so you can fix bugs in a class or
add methods and still load objects that were created with an earlier
version of the class.  If you plan to have long-lived objects that
will see many versions of a class, it may be worth to put a version
number in the objects so that suitable conversions can be made by the
class's \code{__setstate__()} method.

The interface can be summarized as follows.

To pickle an object \code{x} onto a file \code{f}, open for writing:

\begin{verbatim}
p = pickle.Pickler(f)
p.dump(x)
\end{verbatim}

To unpickle an object \code{x} from a file \code{f}, open for reading:

\begin{verbatim}
u = pickle.Unpickler(f)
x = u.load(x)
\end{verbatim}

The \code{Pickler} class only calls the method \code{f.write} with a
string argument.  The \code{Unpickler} calls the methods \code{f.read}
(with an integer argument) and \code{f.readline} (without argument),
both returning a string.  It is explicitly allowed to pass non-file
objects here, as long as they have the right methods.

The following types can be pickled:
\begin{itemize}

\item \code{None}

\item integers, long integers, floating point numbers

\item strings

\item tuples, lists and dictionaries containing only picklable objects

\item class instances whose \code{__dict__} or \code{__setstate__()}
is picklable

\end{itemize}

Attempts to pickle unpicklable objects will raise an exception; when
this happens, an unspecified number of bytes may have been written to
the file argument.

It is possible to make multiple calls to \code{Pickler.dump()} or to
\code{Unpickler.load()}, as long as there is a one-to-one
correspondence between pickler and \code{Unpickler} objects and
between \code{dump} and \code{load} calls for any pair of
corresponding \code{Pickler} and \code{Unpicklers}.  {\em Warning}:
this is intended for pickling multiple objects without intervening
modifications to the objects or their parts.  If you modify an object
and then pickle it again using the same \code{Pickler} instance, the
object is not pickled again --- a reference to it is pickled and the
\code{Unpickler} will return the old value, not the modified one.  (There
are two problems here: (a) detecting changes, and (b) marshalling a
minimal set of changes.  I have no answers.  Garbage Collection may
also become a problem here.)
